---
title: "2nd pass annotations for the PIPAC Xenium data"
author: "heinin"
date: "2025-02-10"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

### Packages and environment variables

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  library(workflowr)
  library(arrow)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(googlesheets4)
  library(workflowr)})

```

### Environment variables and helper functions

```{r, message = F, warning = F}

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(9999)
options(scipen = 99999)
options(ggrepel.max.overlaps = Inf)

source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

```

### Import data

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

nonimmune <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/nonimmune_complete_subset_clustered_NC50_NN20_PC20_Seurat.rds")
nonimmune <- NormalizeData(nonimmune)

immune <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/immune_complete_subset_clustered_NC50_NN20_PC20_Seurat.rds")
immune <- NormalizeData(immune)

p1 <- DimPlot(immune,
        group.by = "leiden_res0.5",
        cols = pipac_immune_cluster_col,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

p2 <- DimPlot(nonimmune,
        group.by = "leiden_res0.5",
        cols = pipac_nonimmune_cluster_col,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

p1 + p2

```

### Marker information and annotations

```{r, message = F, warning = F, fig.width = 6, fig.height = 4}

gs4_deauth()
metadata  <- gs4_get("https://docs.google.com/spreadsheets/d/1sXXwOreLxjMSUoPt79c6jmaQpluWkaxA5P5HfDsed3I/edit?usp=sharing")
markers <- read_sheet(metadata, sheet = "Markers")
second_pass_nonimmune <- read_sheet(metadata, sheet = "Tumor 2nd pass")
second_pass_immune <- read_sheet(metadata, sheet = "Immune 2nd pass")

# Cell type markers by group
gs4_deauth()
markers_by_group  <- gs4_get("https://docs.google.com/spreadsheets/d/1w0wrL-5KwNEWi_VpmriN7YmfwpBPlEYX0MKlwI2ZQu8/edit?usp=sharing")
markers_by_group <- read_sheet(markers_by_group, sheet = "Markers by group")

```

### Top cluster markers

```{r, message = F, warning = F, fig.width = 5, fig.height = 4}

Idents(immune) <- immune$leiden_res0.5
immune_markers <- FindAllMarkers(immune,
                                 return.thresh = 0.01,
                                 logfc.threshold = 0.5,
                                 min.pct = 0.20,
                                 only.pos = T,
                                 verbose = F)

table(immune_markers$cluster)

top_immune_markers <- immune_markers %>%
  arrange(dplyr::desc(abs(avg_log2FC))) %>%
  group_by(cluster) %>%
  dplyr::slice(1:10)

Idents(nonimmune) <- nonimmune$leiden_res0.5
nonimmune_markers <- FindAllMarkers(nonimmune,
                                    return.thresh = 0.01,
                                    logfc.threshold = 0.5,
                                    min.pct = 0.20,
                                    only.pos = T,
                                    verbose = F)

table(nonimmune_markers$cluster)

top_nonimmune_markers <- nonimmune_markers %>%
  arrange(dplyr::desc(abs(avg_log2FC))) %>%
  group_by(cluster) %>%
  dplyr::slice(1:10)

```

### Dotplots of top cluster markers

#### Immune

```{r, message = F, warning = F, fig.width = 6, fig.height = 18}

create_dotplot_heatmap(seurat_object = immune,
                       plot_features = unique(top_immune_markers$gene),
                       group_var = "leiden_res0.5",
                       group_colors = pipac_immune_cluster_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

#### Non-immune

```{r, message = F, warning = F, fig.width = 6, fig.height = 18}

create_dotplot_heatmap(seurat_object = nonimmune,
                       plot_features = unique(top_nonimmune_markers$gene),
                       group_var = "leiden_res0.5",
                       group_colors = pipac_nonimmune_cluster_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

### Saving top markers and marker annotations

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

output_immune_markers <- immune_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:30)

output_cluster_markers <- merge(output_immune_markers, markers, by.x = "gene", by.y = "Gene")

write.table(output_cluster_markers, "/home/hnatri/PIPAC_spatial/immune_cluster_marker_annotations_2ndpass.tsv",
            quote = F, row.names = F, sep = "\t")

output_nonimmune_markers <- nonimmune_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:30)

output_nonimmune_markers <- merge(output_nonimmune_markers, markers, by.x = "gene", by.y = "Gene")

write.table(output_nonimmune_markers, "/home/hnatri/PIPAC_spatial/nonimmune_cluster_marker_annotations_2ndpass.tsv",
            quote = F, row.names = F, sep = "\t")

```

### Lineage markers

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

plot_features <- c("PTPRC",
                   "CD3D", "CD3E", "CD4", "CD8A", # T cells
                   "STAT4", "STAT3", "TIGIT", "GZMB",
                   "SELL", "CD19", # B cells
                   "CD68", "CD44", "MARCO", "APOE", # Macrophages
                   "C1QB", "C1QBP",
                   "MUC5AC", "NOTCH3", "MS4A1", "PGA5", # Lineage markers
                   "FN1", "DCN", "LUM", # Fibroblasts
                   "EGR3", "TP53", "JUN", "KIT", # Tumor
                   "SOX9", "RNF43", "EPCAM")

DotPlot(immune,
        group.by = "leiden_res0.5",
        features = plot_features,
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Immune")

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 22}

FeaturePlot(immune,
            features = plot_features,
            cols = c("gray89", "tomato3"),
            order = T,
            ncol = 4) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  plot_theme #ggtitle("Immune")

```


```{r, message = F, warning = F, fig.width = 10, fig.height = 4}

DotPlot(nonimmune,
        group.by = "leiden_res0.5",
        features = plot_features,
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Non-immune")

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 24}

FeaturePlot(nonimmune,
            features = plot_features,
            cols = c("gray89", "tomato3"),
            order = T,
            ncol = 4) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  plot_theme #ggtitle("Non-immune")

```

### Adding annotations to the subsets and the full object

```{r, message = F, warning = F, fig.width = 10, fig.height = 4, eval = F}

immune$annot2 <- mapvalues(x = immune$leiden_res0.5,
                           from = second_pass_immune$leiden_res0.5,
                           to = second_pass_immune$Annotation)

nonimmune$annot2 <- mapvalues(x = nonimmune$leiden_res0.5,
                              from = second_pass_nonimmune$leiden_res0.5,
                              to = second_pass_nonimmune$Annotation)

saveRDS(immune, "/tgen_labs/banovich/PIPAC/Seurat/immune_complete_subset_clustered_NC50_NN20_PC20_Seurat_annotated.rds")
saveRDS(nonimmune, "/tgen_labs/banovich/PIPAC/Seurat/nonimmune_complete_subset_clustered_NC50_NN20_PC20_Seurat_annotated.rds")

immune_nonimmune <- merge(immune, nonimmune)

seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/merged_spatial_filtered_split_clustered_NC50_NN20_PC20_Seurat.rds")

seurat_data$Annotation <- mapvalues(x = colnames(seurat_data),
                                    from = colnames(immune_nonimmune),
                                    to = immune_nonimmune$annot2)

unique(seurat_data$Annotation)

saveRDS(seurat_data, "/tgen_labs/banovich/PIPAC/Seurat/PIPAC_NC50_NN20_PC20_Seurat_annotated.rds")

```

### Plotting cell type markers by group

```{r, message = F, warning = F, fig.width = 16, fig.height = 6, eval = T}

seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/PIPAC_NC50_NN20_PC20_Seurat_annotated.rds")

DotPlot(seurat_data,
        group.by = "Annotation",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "Cell-cell crosstalk"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Cell-cell crosstalk")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

DotPlot(seurat_data,
        group.by = "Annotation",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "Macrophage signaling"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Macrophage signaling")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

DotPlot(seurat_data,
        group.by = "Annotation",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "DNA damage and cell faith"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("DNA damage and cell faith")

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 6, eval = T}

DotPlot(seurat_data,
        group.by = "Annotation",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "Functional markers"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Functional markers")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

DotPlot(seurat_data,
        group.by = "Annotation",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "Mesothelial"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Mesothelial")

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 6, eval = F}

DotPlot(seurat_data,
        group.by = "Annotation",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "Macrophages"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Macrophages")

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 6, eval = T}

DotPlot(seurat_data,
        group.by = "Annotation",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "CAFs"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("CAFs")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

DotPlot(seurat_data,
        group.by = "Annotation",
        features = grep("COL", rownames(seurat_data), value = T),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Collagen")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

DotPlot(seurat_data,
        group.by = "Annotation",
        features = grep("COL", rownames(seurat_data), value = T),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Collagen")

```

### Cell cycle markers

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

DotPlot(seurat_data,
        group.by = "Annotation",
        features = c(s.genes, g2m.genes),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("S/G2M genes")

```

### Saving markers one more time

```{r, message = F, warning = F, fig.width = 10, fig.height = 4, eval = T}

#seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/PIPAC_NC50_NN20_PC20_Seurat_annotated.rds")

Idents(seurat_data) <- seurat_data$Annotation
celltype_markers <- FindAllMarkers(seurat_data,
                                   return.thresh = 0.01,
                                   #logfc.threshold = 0.5,
                                   min.pct = 0.20,
                                   only.pos = T,
                                   verbose = F)

output_celltype_markers <- celltype_markers %>%
  arrange(dplyr::desc(abs(avg_log2FC))) %>%
  group_by(cluster) %>%
  dplyr::slice(1:30)

output_celltype_markers <- merge(output_celltype_markers, markers, by.x = "gene", by.y = "Gene")

write.table(output_celltype_markers, "/home/hnatri/PIPAC_spatial/celltype_markers.tsv",
            quote = F, row.names = F, sep = "\t")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = F}

DimPlot(seurat_data,
        reduction = "sp",
        group.by = "Sample",
        split.by = "TMA",
        ncol = 3,
        raster = T) +
    coord_fixed(ratio = 1) +
    theme_minimal() +
    NoLegend() +
    ggtitle("")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = F}

DimPlot(seurat_data,
        group.by = "leiden_res0.5",
        cols = pipac_nonimmune_cluster_col,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

```
