---
title: "PIPAC Arm 3 differential expression analysis"
author: "heinin"
date: "2025-03-25"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

### Load packages

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  library(workflowr)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(googlesheets4)
  library(workflowr)
  library(patchwork)
  library(ggpmisc)})

```

### Environment variables and helper functions

```{r, message = F, warning = F}

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(9999)
options(scipen = 99999)
options(ggrepel.max.overlaps = Inf)

source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

# Calling DEGs between two groups for each cell type
get_DEGs <- function(seuratdata, celltypes, groupvar, group1, group2){
  #message(xx)
  DEGlist <- lapply(celltypes, function(xx){
    data_subset <- subset(seuratdata, subset = Annotation == xx)
    Idents(data_subset) <- as.character(unlist(data_subset[[groupvar]]))
    
    if (min(table(unlist(data_subset[[groupvar]])))<20){
      return(NULL)
    }
    
    if (all((c(group1, group2) %in% unlist(data_subset[[groupvar]]) == c(T, T)))){
      markers <- FindMarkers(data_subset,
                             ident.1 = group1,
                             ident.2 = group2,
                             assay = "RNA",
                             verbose = F)
      markers$feature <- rownames(markers)
      markers$celltype <- xx
      
      return(markers)
    } else {
      return(NULL)
    }
    })
  
  as.data.frame(do.call(rbind, DEGlist))
}

```

### Marker information

```{r, message = F, warning = F, fig.width = 6, fig.height = 4}

gs4_deauth()
metadata  <- gs4_get("https://docs.google.com/spreadsheets/d/1sXXwOreLxjMSUoPt79c6jmaQpluWkaxA5P5HfDsed3I/edit?usp=sharing")
markers <- read_sheet(metadata, sheet = "Markers")
second_pass_nonimmune <- read_sheet(metadata, sheet = "Tumor 2nd pass")
second_pass_immune <- read_sheet(metadata, sheet = "Immune 2nd pass")

# Cell type markers by group
gs4_deauth()
markers_by_group  <- gs4_get("https://docs.google.com/spreadsheets/d/1w0wrL-5KwNEWi_VpmriN7YmfwpBPlEYX0MKlwI2ZQu8/edit?usp=sharing")
markers_by_group <- read_sheet(markers_by_group, sheet = "Markers by group")

```

### Import and subset data

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/PIPAC_NC50_NN20_PC20_Seurat_annotated_metadata_niches.rds")
DefaultAssay(seurat_data) <- "RNA"
seurat_data <- seurat_data %>%
  NormalizeData() %>%
  ScaleData()

seurat_arm3 <- subset(seurat_data, subset = Arm == "Arm3")
seurat_arm3 <- seurat_arm3 %>%
  NormalizeData() %>%
  ScaleData()

```

### DEGs by response, all tumors

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

response_DEGs <- get_DEGs(seuratdata = seurat_arm3,
                          celltypes = unique(seurat_arm3$Annotation),
                          groupvar = "BestResponse",
                          group1 = "SD",
                          group2 = "PR")

# Distribution of p-values and log2FC
hist(response_DEGs$p_val)
hist(response_DEGs$p_val_adj)
hist(response_DEGs$avg_log2FC)

response_DEGs_sig <- response_DEGs %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

table(response_DEGs_sig$celltype)

```

### DEGs by response, baseline tumors

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

baseline_tumor <- subset(seurat_arm3, subset = Tissue == "Tumor" &
                           Timepoint == 0)
baseline_tumor <- baseline_tumor %>%
  NormalizeData() %>%
  ScaleData()

response_DEGs_baseline <- get_DEGs(seuratdata = baseline_tumor,
                                   celltypes = unique(baseline_tumor$Annotation),
                                   groupvar = "BestResponse",
                                   group1 = "SD",
                                   group2 = "PR")

# Distribution of p-values and log2FC
hist(response_DEGs_baseline$p_val)
hist(response_DEGs_baseline$p_val_adj)
hist(response_DEGs_baseline$avg_log2FC)

response_DEGs_baseline_sig <- response_DEGs_baseline %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

table(response_DEGs_baseline_sig$celltype)

```

### DEGs by response, post-treatment tumors

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

post_treatment_tumor <- subset(seurat_arm3, subset = Tissue == "Tumor" &
                                 Timepoint %in% c(6, 12))
post_treatment_tumor <- post_treatment_tumor %>%
  NormalizeData() %>%
  ScaleData()

posttreatment_response_DEGs <- get_DEGs(seuratdata = post_treatment_tumor,
                                        celltypes = unique(post_treatment_tumor$Annotation),
                                        groupvar = "BestResponse",
                                        group1 = "SD",
                                        group2 = "PR")

# Distribution of p-values and log2FC
hist(posttreatment_response_DEGs$p_val)
hist(posttreatment_response_DEGs$p_val_adj)
hist(posttreatment_response_DEGs$avg_log2FC)

posttreatment_response_DEGs_sig <- posttreatment_response_DEGs %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

table(posttreatment_response_DEGs_sig$celltype)

```

#### Plotting

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

#plot_list <- lapply(unique(seurat_arm3$Annotation))

```

### DEGs for myeloid, lymphoid, and fibroblast subsets

#### Baseline tumors, myeloid

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

sort(unique(baseline_tumor$Annotation))

Idents(baseline_tumor) <- baseline_tumor$BestResponse

myel_celltypes <- c("B1", "B2", "DC", "Mast", "Plasma1", "Plasma2", "Plasma3",
                    paste0("M", seq(1,9)))

baseline_myel <- subset(baseline_tumor, subset = Annotation %in% myel_celltypes)
baseline_myel <- baseline_myel %>%
  NormalizeData() %>%
  ScaleData()

response_baseline_myel <- FindMarkers(baseline_myel,
                                      ident.1 = "SD",
                                      ident.2 = "PR",
                                      assay = "RNA",
                                      verbose = F)
response_baseline_myel$feature <- rownames(response_baseline_myel)

# Distribution of p-values and log2FC
hist(response_baseline_myel$p_val)
hist(response_baseline_myel$p_val_adj)
hist(response_baseline_myel$avg_log2FC)

response_baseline_myel_sig <- response_baseline_myel %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

response_baseline_myel_sig$feature

response_baseline_myel_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 8}

VlnPlot(subset(baseline_myel, subset = BestResponse %in% c("SD", "PR")),
        group.by = "BestResponse",
        features = response_baseline_myel_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

VlnPlot(subset(baseline_myel, subset = BestResponse %in% c("SD", "PR")),
        group.by = "Annotation",
        split.by = "BestResponse",
        split.plot = T,
        features = response_baseline_myel_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Baseline tumors, lymphoid

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

lymph_celltypes <- paste0("L", seq(1,5))

baseline_lymph <- subset(baseline_tumor, subset = Annotation %in% lymph_celltypes)
baseline_lymph <- baseline_lymph %>%
  NormalizeData() %>%
  ScaleData()

response_baseline_lymph <- FindMarkers(baseline_lymph,
                                       ident.1 = "SD",
                                       ident.2 = "PR",
                                       assay = "RNA",
                                       verbose = F)
response_baseline_lymph$feature <- rownames(response_baseline_lymph)

# Distribution of p-values and log2FC
hist(response_baseline_lymph$p_val)
hist(response_baseline_lymph$p_val_adj)
hist(response_baseline_lymph$avg_log2FC)

response_baseline_lymph_sig <- response_baseline_lymph %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

response_baseline_lymph_sig$feature

response_baseline_lymph_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 8}

VlnPlot(subset(baseline_lymph, subset = BestResponse %in% c("SD", "PR")),
        group.by = "BestResponse",
        features = response_baseline_lymph_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

VlnPlot(subset(baseline_lymph, subset = BestResponse %in% c("SD", "PR")),
        group.by = "Annotation",
        split.by = "BestResponse",
        split.plot = T,
        features = response_baseline_lymph_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Baseline tumors, fibroblast

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

fibro_celltypes <- paste0("F", seq(1,5))

baseline_fibro <- subset(baseline_tumor, subset = Annotation %in% fibro_celltypes)
baseline_fibro <- baseline_fibro %>%
  NormalizeData() %>%
  ScaleData()

response_baseline_fibro <- FindMarkers(baseline_fibro,
                                       ident.1 = "SD",
                                       ident.2 = "PR",
                                       assay = "RNA",
                                       verbose = F)
response_baseline_fibro$feature <- rownames(response_baseline_fibro)

# Distribution of p-values and log2FC
hist(response_baseline_fibro$p_val)
hist(response_baseline_fibro$p_val_adj)
hist(response_baseline_fibro$avg_log2FC)

response_baseline_fibro_sig <- response_baseline_fibro %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

response_baseline_fibro_sig$feature

response_baseline_fibro_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 14}

VlnPlot(subset(baseline_fibro, subset = BestResponse %in% c("SD", "PR")),
        group.by = "BestResponse",
        features = response_baseline_fibro_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 14}

VlnPlot(subset(baseline_fibro, subset = BestResponse %in% c("SD", "PR")),
        group.by = "Annotation",
        split.by = "BestResponse",
        split.plot = T,
        features = response_baseline_fibro_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Post-treatment tumors, myeloid

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

sort(unique(post_treatment_tumor$Annotation))

Idents(post_treatment_tumor) <- post_treatment_tumor$BestResponse

myel_celltypes <- c("B1", "B2", "DC", "Mast", "Plasma1", "Plasma2", "Plasma3",
                    paste0("M", seq(1,9)))

treated_myel <- subset(post_treatment_tumor, subset = Annotation %in% myel_celltypes)
treated_myel <- treated_myel %>%
  NormalizeData() %>%
  ScaleData()

response_treated_myel <- FindMarkers(treated_myel,
                                     ident.1 = "SD",
                                     ident.2 = "PR",
                                     assay = "RNA",
                                     verbose = F)
response_treated_myel$feature <- rownames(response_treated_myel)

# Distribution of p-values and log2FC
hist(response_treated_myel$p_val)
hist(response_treated_myel$p_val_adj)
hist(response_treated_myel$avg_log2FC)

response_treated_myel_sig <- response_treated_myel %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

response_treated_myel_sig$feature

response_treated_myel_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 8}

VlnPlot(subset(treated_myel, subset = BestResponse %in% c("SD", "PR")),
        group.by = "BestResponse",
        features = response_treated_myel_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

VlnPlot(subset(treated_myel, subset = BestResponse %in% c("SD", "PR")),
        group.by = "Annotation",
        split.by = "BestResponse",
        split.plot = T,
        features = response_treated_myel_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Post-treatment tumors, lymphoid

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

lymph_celltypes <- paste0("L", seq(1,5))

treated_lymph <- subset(post_treatment_tumor, subset = Annotation %in% lymph_celltypes)
treated_lymph <- treated_lymph %>%
  NormalizeData() %>%
  ScaleData()

response_treated_lymph <- FindMarkers(treated_lymph,
                                      ident.1 = "SD",
                                      ident.2 = "PR",
                                      assay = "RNA",
                                      verbose = F)
response_treated_lymph$feature <- rownames(response_treated_lymph)

# Distribution of p-values and log2FC
hist(response_treated_lymph$p_val)
hist(response_treated_lymph$p_val_adj)
hist(response_treated_lymph$avg_log2FC)

response_treated_lymph_sig <- response_treated_lymph %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

response_treated_lymph_sig$feature

response_treated_lymph_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 8}

VlnPlot(subset(treated_lymph, subset = BestResponse %in% c("SD", "PR")),
        group.by = "BestResponse",
        features = response_treated_lymph_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

VlnPlot(subset(treated_lymph, subset = BestResponse %in% c("SD", "PR")),
        group.by = "Annotation",
        split.by = "BestResponse",
        split.plot = T,
        features = response_treated_lymph_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Post-treatment tumors, fibroblast

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

fibro_celltypes <- paste0("F", seq(1,5))

treated_fibro <- subset(post_treatment_tumor, subset = Annotation %in% fibro_celltypes)
treated_fibro <- treated_fibro %>%
  NormalizeData() %>%
  ScaleData()

response_treated_fibro <- FindMarkers(treated_fibro,
                                      ident.1 = "SD",
                                      ident.2 = "PR",
                                      assay = "RNA",
                                      verbose = F)
response_treated_fibro$feature <- rownames(response_treated_fibro)

# Distribution of p-values and log2FC
hist(response_treated_fibro$p_val)
hist(response_treated_fibro$p_val_adj)
hist(response_treated_fibro$avg_log2FC)

response_treated_fibro_sig <- response_treated_fibro %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

response_treated_fibro_sig$feature

response_treated_fibro_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 12}

VlnPlot(subset(treated_fibro, subset = BestResponse %in% c("SD", "PR")),
        group.by = "BestResponse",
        features = response_treated_fibro_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 12}

VlnPlot(subset(treated_fibro, subset = BestResponse %in% c("SD", "PR")),
        group.by = "Annotation",
        split.by = "BestResponse",
        split.plot = T,
        features = response_treated_fibro_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

