---
title: "Arm 3 niche-based comparative analysis"
author: "heinin"
date: "2025-03-21"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

### Load packages

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  library(workflowr)
  library(arrow)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(googlesheets4)
  library(workflowr)
  library(patchwork)
  library(ggpmisc)
  library(scProportionTest)})

```

### Environment variables and helper functions

```{r, message = F, warning = F}

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(9999)
options(scipen = 99999)
options(ggrepel.max.overlaps = Inf)

source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

```

### Marker information

```{r, message = F, warning = F, fig.width = 6, fig.height = 4}

gs4_deauth()
metadata  <- gs4_get("https://docs.google.com/spreadsheets/d/1sXXwOreLxjMSUoPt79c6jmaQpluWkaxA5P5HfDsed3I/edit?usp=sharing")
markers <- read_sheet(metadata, sheet = "Markers")
second_pass_nonimmune <- read_sheet(metadata, sheet = "Tumor 2nd pass")
second_pass_immune <- read_sheet(metadata, sheet = "Immune 2nd pass")

# Cell type markers by group
gs4_deauth()
markers_by_group  <- gs4_get("https://docs.google.com/spreadsheets/d/1w0wrL-5KwNEWi_VpmriN7YmfwpBPlEYX0MKlwI2ZQu8/edit?usp=sharing")
markers_by_group <- read_sheet(markers_by_group, sheet = "Markers by group")

```

### Import and subset data

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/PIPAC_NC50_NN20_PC20_Seurat_annotated_metadata_niches.rds")
DefaultAssay(seurat_data) <- "RNA"
seurat_data <- seurat_data %>%
  NormalizeData() %>%
  ScaleData()

seurat_arm3 <- subset(seurat_data, subset = Arm == "Arm3")
seurat_arm3 <- seurat_arm3 %>%
  NormalizeData() %>%
  ScaleData()

```

## Niche abundance by sample, group, and correlation with survival

```{r, message = F, warning = F, fig.width = 10, fig.height = 24}

barplot_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
  seurat_arm3@meta.data[,niche_column] <- factor(seurat_arm3@meta.data[,niche_column],
                                                 levels = niches)
  niche_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(niches))
  names(niche_col) <- levels(seurat_arm3@meta.data[,niche_column])
  
  create_barplot(seurat_arm3,
                 group_var = "Sample",
                 plot_var = niche_column,
                 plot_levels = levels(seurat_arm3@meta.data[,niche_column]),
                 group_levels = sort(unique(seurat_arm3$Sample)),
                 plot_colors = niche_col,
                 var_names =  c("Frequency (%)", ""),
                 legend_title = "Niche")
})

wrap_plots(barplot_list, ncol = 1)

```

### All tumors

```{r, message = F, warning = F, fig.width = 6, fig.height = 24}

barplot_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
  seurat_arm3@meta.data[,niche_column] <- factor(seurat_arm3@meta.data[,niche_column],
                                                 levels = niches)
  niche_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(niches))
  names(niche_col) <- niches
  
  create_barplot(seurat_arm3,
                 group_var = "Sample",
                 plot_var = niche_column,
                 plot_levels = levels(seurat_arm3@meta.data[,niche_column]),
                 group_levels = sort(unique(seurat_arm3$Sample)),
                 plot_colors = niche_col,
                 var_names =  c("Frequency (%)", ""),
                 legend_title = "Niche")
})

wrap_plots(barplot_list, ncol = 1)

```

#### Proportion plots by best response

```{r, message = F, warning = F, fig.width = 10, fig.height = 12}

barplot_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
  seurat_arm3@meta.data[,niche_column] <- factor(seurat_arm3@meta.data[,niche_column],
                                                 levels = niches)
  niche_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(niches))
  names(niche_col) <- levels(seurat_arm3@meta.data[,niche_column])
  
  create_barplot(seurat_arm3,
                 group_var = "BestResponse",
                 plot_var = niche_column,
                 plot_levels = levels(seurat_arm3@meta.data[,niche_column]),
                 group_levels = sort(unique(seurat_arm3$BestResponse)),
                 plot_colors = niche_col,
                 var_names =  c("Frequency (%)", ""),
                 legend_title = "Niche")
})

wrap_plots(barplot_list, ncol = 2)

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 12}

scatterplot_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
  seurat_arm3@meta.data[,niche_column] <- factor(seurat_arm3@meta.data[,niche_column],
                                                 levels = niches)
  niche_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(niches))
  names(niche_col) <- levels(seurat_arm3@meta.data[,niche_column])
   
  create_clusterpropplot(subset(seurat_arm3, subset = BestResponse %in% c("SD", "PR")),
                         group_var = "BestResponse",
                         plot_var = niche_column,
                         group1 = "SD",
                         group2 = "PR",
                         plot_colors = niche_col,
                         var_names =  c("Frequency (%)", ""),
                         legend_title = "Niche")
})

wrap_plots(scatterplot_list, ncol = 2)

```

### Testing for significance with scProportionTest

```{r, message = F, warning = F, fig.width = 6, fig.height = 6}

propplot_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
   
  prop_test <- sc_utils(seurat_arm3)
  prop_test <- permutation_test(
	   prop_test, cluster_identity = niche_column,
	   sample_1 = "PR", sample_2 = "SD",
	   sample_identity = "BestResponse")

  permutation_plot(prop_test) +
    ggtitle("All tumors")
})

wrap_plots(propplot_list, ncol = 2)

```

#### Niche abundance correlation with survival in tumor samples

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

seurat_arm3 <- subset(seurat_arm3, subset = Tissue == "Tumor")

# Combining the quadrant samples for each patient/timepoint
# Note: there is only one sample for per timepoint, so doesn't make a
# difference here
seurat_arm3$Patient_Tissue_Timepoint <- paste0(seurat_arm3$Patient_ID, "_",
                                               seurat_arm3$Tissue, "_",
                                               seurat_arm3$Timepoint)

metadata <- seurat_arm3@meta.data %>%
  dplyr::select("Sample", "Patient_ID", "Patient_Tissue_Timepoint", "DISEASESITE", "osmonths", "pfsmonths", "No_of_PIPACs") %>%
  distinct()

list_of_plotlists <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
  seurat_arm3@meta.data[,niche_column] <- factor(seurat_arm3@meta.data[,niche_column],
                                                 levels = niches)
  niche_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(niches))
  names(niche_col) <- levels(seurat_arm3@meta.data[,niche_column])
  
  prop_table <- as.data.frame(table(unlist(seurat_arm3[[niche_column]]),   as.character(seurat_arm3@meta.data$Patient_Tissue_Timepoint)))
  colnames(prop_table) <- c("Niche", "Patient_Tissue_Timepoint", "Freq")
  prop_table <- spread(prop_table, Niche, Freq)
  # Converting to percentage
  prop_table[,2:length(prop_table)] <- (prop_table[,2:length(prop_table)]/rowSums(prop_table[,2:length(prop_table)]))*100
  prop_table <- gather(prop_table, Niche, Freq, names(prop_table)[2:length(names(prop_table))], factor_key=TRUE)
  
  # Adding survival data
  prop_survival <- merge(prop_table, metadata, by = "Patient_Tissue_Timepoint")
  prop_survival$osmonths <- unlist(prop_survival$osmonths)
  prop_survival$pfmonths <- unlist(prop_survival$pfmonths)
  
  # Plotting
  plot_list <- lapply(niches, function(niche){
    prop_survival %>%
      filter(Niche == niche) %>%
      ggplot(aes(x = Freq, y = osmonths, color = Niche)) +
      geom_point() +
      scale_color_manual(name = "Niche", values = niche_col) +
      stat_poly_line() +
      stat_poly_eq(formula = y ~ x,
                   rr.digits = 3, p.digits = 3, small.p = T,
                   aes(label = paste(after_stat(rr.label), after_stat(p.value.label), sep = "*`,`~")),
                   label.x = 0.01, label.y = 1, parse = T) +
      ylab("Overall Survival") +
      xlab(paste0("% of niche ", niche, " cells")) +
      theme_classic() +
      theme(axis.text.x = element_text(color = "black"),
            axis.text.y = element_text(color = "black")) +
      NoLegend()
  })
  return(plot_list)
})

```

K = 3

```{r, message = F, warning = F, fig.width = 10, fig.height = 3}

wrap_plots(list_of_plotlists[[1]], ncol = 3)

```

K = 4

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

wrap_plots(list_of_plotlists[[2]], ncol = 3)

```

K = 5

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

wrap_plots(list_of_plotlists[[3]], ncol = 3)

```

K = 6

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

wrap_plots(list_of_plotlists[[4]], ncol = 3)

```

K = 7

```{r, message = F, warning = F, fig.width = 10, fig.height = 9}

wrap_plots(list_of_plotlists[[5]], ncol = 3)

```

K = 8

```{r, message = F, warning = F, fig.width = 10, fig.height = 9}

wrap_plots(list_of_plotlists[[6]], ncol = 3)

```

### Baseline tumors

```{r, message = F, warning = F, fig.width = 6, fig.height = 24}

unique(seurat_arm3$Timepoint)
baseline_tumor <- subset(seurat_arm3, subset = Tissue == "Tumor" &
                           Timepoint == 0)
baseline_tumor <- baseline_tumor %>%
  NormalizeData() %>%
  ScaleData()

barplot_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
  niche_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(niches))
  names(niche_col) <- niches
  
  create_barplot(baseline_tumor,
                 group_var = "Sample",
                 plot_var = niche_column,
                 plot_levels = levels(baseline_tumor@meta.data[,niche_column]),
                 group_levels = sort(unique(baseline_tumor$Sample)),
                 plot_colors = niche_col,
                 var_names =  c("Frequency (%)", ""),
                 legend_title = "Niche")
})

wrap_plots(barplot_list, ncol = 1)

```

### Testing for significance with scProportionTest

```{r, message = F, warning = F, fig.width = 6, fig.height = 6}

propplot_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
   
  prop_test <- sc_utils(baseline_tumor)
  prop_test <- permutation_test(
	   prop_test, cluster_identity = niche_column,
	   sample_1 = "PR", sample_2 = "SD",
	   sample_identity = "BestResponse")

  permutation_plot(prop_test) +
    ggtitle("Baseline tumors")
})

wrap_plots(propplot_list, ncol = 2)

```

#### Niche abundance correlation with survival

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

metadata <- baseline_tumor@meta.data %>%
  dplyr::select("Sample", "Patient_ID", "DISEASESITE", "osmonths", "pfsmonths", "No_of_PIPACs") %>%
  distinct()

list_of_plotlists <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
  niche_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(niches))
  names(niche_col) <- niches
  
  prop_table <- as.data.frame(table(unlist(baseline_tumor[[niche_column]]), as.character(baseline_tumor@meta.data$Sample)))
  colnames(prop_table) <- c("Niche", "Sample", "Freq")
  prop_table <- spread(prop_table, Niche, Freq)
  # Converting to percentage
  prop_table[,2:length(prop_table)] <- (prop_table[,2:length(prop_table)]/rowSums(prop_table[,2:length(prop_table)]))*100
  prop_table <- gather(prop_table, Niche, Freq, names(prop_table)[2:length(names(prop_table))], factor_key=TRUE)
  
  # Adding survival data
  prop_survival <- merge(prop_table, metadata, by = "Sample")
  prop_survival$osmonths <- unlist(prop_survival$osmonths)
  prop_survival$pfmonths <- unlist(prop_survival$pfmonths)
  
  # Plotting
  plot_list <- lapply(niches, function(niche){
    prop_survival %>%
      filter(Niche == niche) %>%
      ggplot(aes(x = Freq, y = osmonths, color = Niche)) +
      geom_point() +
      scale_color_manual(name = "Niche", values = niche_col) +
      stat_poly_line() +
      stat_poly_eq(formula = y ~ x,
                   rr.digits = 3, p.digits = 3, small.p = T,
                   aes(label = paste(after_stat(rr.label), after_stat(p.value.label), sep = "*`,`~")),
                   label.x = 0.01, label.y = 1, parse = T) +
      ylab("Overall Survival") +
      xlab(paste0("% of niche ", niche, " cells")) +
      theme_classic() +
      theme(axis.text.x = element_text(color = "black"),
            axis.text.y = element_text(color = "black")) +
      NoLegend()
  })
  return(plot_list)
})

```

K = 3

```{r, message = F, warning = F, fig.width = 10, fig.height = 3}

wrap_plots(list_of_plotlists[[1]], ncol = 3)

```

K = 4

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

wrap_plots(list_of_plotlists[[2]], ncol = 3)

```

K = 5

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

wrap_plots(list_of_plotlists[[3]], ncol = 3)

```

K = 6

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

wrap_plots(list_of_plotlists[[4]], ncol = 3)

```

K = 7

```{r, message = F, warning = F, fig.width = 10, fig.height = 9}

wrap_plots(list_of_plotlists[[5]], ncol = 3)

```

K = 8

```{r, message = F, warning = F, fig.width = 10, fig.height = 9}

wrap_plots(list_of_plotlists[[6]], ncol = 3)

```

### Post-treatment tumors (D6)

```{r, message = F, warning = F, fig.width = 6, fig.height = 24}

unique(seurat_arm3$Timepoint)
later_tumor <- subset(seurat_arm3, subset = Tissue == "Tumor" &
                        Timepoint == 6)
later_tumor <- later_tumor %>%
  NormalizeData() %>%
  ScaleData()

barplot_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
  niche_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(niches))
  names(niche_col) <- niches
  
  create_barplot(later_tumor,
                 group_var = "Sample",
                 plot_var = niche_column,
                 plot_levels = levels(later_tumor@meta.data[,niche_column]),
                 group_levels = sort(unique(later_tumor$Sample)),
                 plot_colors = niche_col,
                 var_names =  c("Frequency (%)", ""),
                 legend_title = "Niche")
})

wrap_plots(barplot_list, ncol = 1)

```

### Testing for significance with scProportionTest

```{r, message = F, warning = F, fig.width = 6, fig.height = 6}

propplot_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
   
  prop_test <- sc_utils(later_tumor)
  prop_test <- permutation_test(
	   prop_test, cluster_identity = niche_column,
	   sample_1 = "PR", sample_2 = "SD",
	   sample_identity = "BestResponse")
  
  permutation_plot(prop_test) +
    ggtitle("After treatment tumors")
})

wrap_plots(propplot_list, ncol = 2)

```

#### Niche abundance correlation with survival

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

metadata <- later_tumor@meta.data %>%
  dplyr::select("Sample", "Patient_ID", "DISEASESITE", "osmonths", "pfsmonths", "No_of_PIPACs") %>%
  distinct()

list_of_plotlists <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  niches <- seq(1, k)
  niche_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(niches))
  names(niche_col) <- niches
  
  prop_table <- as.data.frame(table(unlist(later_tumor[[niche_column]]), as.character(later_tumor@meta.data$Sample)))
  colnames(prop_table) <- c("Niche", "Sample", "Freq")
  prop_table <- spread(prop_table, Niche, Freq)
  # Converting to percentage
  prop_table[,2:length(prop_table)] <- (prop_table[,2:length(prop_table)]/rowSums(prop_table[,2:length(prop_table)]))*100
  prop_table <- gather(prop_table, Niche, Freq, names(prop_table)[2:length(names(prop_table))], factor_key=TRUE)
  
  # Adding survival data
  prop_survival <- merge(prop_table, metadata, by = "Sample")
  prop_survival$osmonths <- unlist(prop_survival$osmonths)
  prop_survival$pfmonths <- unlist(prop_survival$pfmonths)
  
  # Plotting
  plot_list <- lapply(niches, function(niche){
    prop_survival %>%
      filter(Niche == niche) %>%
      ggplot(aes(x = Freq, y = osmonths, color = Niche)) +
      geom_point() +
      scale_color_manual(name = "Niche", values = niche_col) +
      stat_poly_line() +
      stat_poly_eq(formula = y ~ x,
                   rr.digits = 3, p.digits = 3, small.p = T,
                   aes(label = paste(after_stat(rr.label), after_stat(p.value.label), sep = "*`,`~")),
                   label.x = 0.01, label.y = 1, parse = T) +
      ylab("Overall Survival") +
      xlab(paste0("% of niche ", niche, " cells")) +
      theme_classic() +
      theme(axis.text.x = element_text(color = "black"),
            axis.text.y = element_text(color = "black")) +
      NoLegend()
  })
  return(plot_list)
})

```

K = 3

```{r, message = F, warning = F, fig.width = 10, fig.height = 3}

wrap_plots(list_of_plotlists[[1]], ncol = 3)

```

K = 4

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

wrap_plots(list_of_plotlists[[2]], ncol = 3)

```

K = 5

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

wrap_plots(list_of_plotlists[[3]], ncol = 3)

```

K = 6

```{r, message = F, warning = F, fig.width = 10, fig.height = 6}

wrap_plots(list_of_plotlists[[4]], ncol = 3)

```

K = 7

```{r, message = F, warning = F, fig.width = 10, fig.height = 9}

wrap_plots(list_of_plotlists[[5]], ncol = 3)

```

K = 8

```{r, message = F, warning = F, fig.width = 10, fig.height = 9}

wrap_plots(list_of_plotlists[[6]], ncol = 3)

```

## Niche pseudo-bulk gene expression correlation with survival

### Baseline tumors

```{r, message = F, warning = F, fig.width = 8, fig.height = 4, eval = F}

# Looping through each k
exp_corr_list <- lapply(seq(3, 8), function(k){
  niche_column <- paste0("niche_k", k, "_n30")
  message(niche_column)
  
  baseline_tumor@meta.data$Niche <- unlist(baseline_tumor[[niche_column]])
  # Looping through each gene in each niche and testing for significance
  corr_res_tables <- lapply(seq(1, k), function(niche){
    message(niche)
    
    niche_data <- subset(baseline_tumor, subset = Niche == niche)
    
    if(length(unique(niche_data$Patient_ID)) < 5){
      return(NULL)
    }
    
    # Log-normalized pseudobulk data
    lognorm_counts <- LayerData(niche_data,
                                assay = "RNA",
                                layer = "data")
    lognorm_counts <- lognorm_counts %>% t() %>% as.data.frame()
    lognorm_counts$Patient_ID <- niche_data$Patient_ID
    
    lognorm_counts <- lognorm_counts %>% group_by(Patient_ID) %>%
      summarise_at(vars(-group_cols()), mean) %>%
      as.data.frame()
    
    lognorm_counts <- t(lognorm_counts) %>%
      as.data.frame()
      
    colnames(lognorm_counts) <- lognorm_counts[1,]
    lognorm_counts <- lognorm_counts[2:nrow(lognorm_counts),]
    
    pseudobulk_exp <- lognorm_counts
    
    pseudobulk_exp <- sapply(pseudobulk_exp, as.numeric )
    rownames(pseudobulk_exp) <- rownames(lognorm_counts)
  
    corr_res_table <- as.data.frame(matrix(NA, nrow = nrow(lognorm_counts), ncol = 4))
    rownames(corr_res_table) <- rownames(lognorm_counts)
    
    colnames(corr_res_table) <- c("niche", "cor", "pval", "k")
  
    for(gene in rownames(lognorm_counts)){
      pseudobulk_gene_exp <- pseudobulk_exp %>%
        as.data.frame() %>%
        filter(row.names(.) == gene) %>%
        pivot_longer(cols = colnames(pseudobulk_exp),
                     names_to = "Patient_ID",
                     values_to = gene)
      
      # Adding survival info
      exp_survival <- merge(pseudobulk_gene_exp, metadata, by = "Patient_ID")
      
      # R2 and p for correlation
      cor_res <- cor.test(exp_survival[,gene], unlist(exp_survival$osmonths),
                          method = "pearson")
      
      cor <- as.numeric(cor_res$estimate)
      pval <- as.numeric(cor_res$p.value)
      
      corr_res_table[gene, "niche"] <- niche
      corr_res_table[gene, "cor"] <- cor
      corr_res_table[gene, "pval"] <- pval
      corr_res_table[gene, "k"] <- niche_column
    }
    
    return(corr_res_table)
    
  })
  
  
})


names(plot_list) <- test_genes

```








