---
title: "PIPAC Arm 3 pre- and post-treatment comparison"
author: "heinin"
date: "2025-04-23"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Load packages

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  library(workflowr)
  library(arrow)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(googlesheets4)
  library(workflowr)
  library(patchwork)
  library(ggpmisc)
  library(scProportionTest)})

```

## Environment variables and helper functions

```{r, message = F, warning = F}

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(9999)
options(scipen = 99999)
options(ggrepel.max.overlaps = Inf)

source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

# Calling DEGs between two groups for each cell type
get_DEGs <- function(seuratdata, celltypes, groupvar, group1, group2){
  #message(xx)
  DEGlist <- lapply(celltypes, function(xx){
    data_subset <- subset(seuratdata, subset = Annotation == xx)
    Idents(data_subset) <- as.character(unlist(data_subset[[groupvar]]))
    
    if (min(table(unlist(data_subset[[groupvar]])))<20){
      return(NULL)
    }
    
    if (all((c(group1, group2) %in% unlist(data_subset[[groupvar]]) == c(T, T)))){
      markers <- FindMarkers(data_subset,
                             ident.1 = group1,
                             ident.2 = group2,
                             assay = "RNA",
                             verbose = F)
      markers$feature <- rownames(markers)
      markers$celltype <- xx
      
      return(markers)
    } else {
      return(NULL)
    }
    })
  
  as.data.frame(do.call(rbind, DEGlist))
}

```

## Marker information

```{r, message = F, warning = F, fig.width = 6, fig.height = 4}

gs4_deauth()
metadata  <- gs4_get("https://docs.google.com/spreadsheets/d/1sXXwOreLxjMSUoPt79c6jmaQpluWkaxA5P5HfDsed3I/edit?usp=sharing")
markers <- read_sheet(metadata, sheet = "Markers")
second_pass_nonimmune <- read_sheet(metadata, sheet = "Tumor 2nd pass")
second_pass_immune <- read_sheet(metadata, sheet = "Immune 2nd pass")

# Cell type markers by group
gs4_deauth()
markers_by_group  <- gs4_get("https://docs.google.com/spreadsheets/d/1w0wrL-5KwNEWi_VpmriN7YmfwpBPlEYX0MKlwI2ZQu8/edit?usp=sharing")
markers_by_group <- read_sheet(markers_by_group, sheet = "Markers by group")

```

## Import and subset data

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/PIPAC_NC50_NN20_PC20_Seurat_annotated_metadata_niches.rds")
DefaultAssay(seurat_data) <- "RNA"
VariableFeatures(seurat_data) <- rownames(seurat_data)

seurat_data <- seurat_data %>%
  NormalizeData() %>%
  ScaleData()

seurat_arm3 <- subset(seurat_data, subset = Arm == "Arm3")
seurat_arm3 <- seurat_arm3 %>%
  NormalizeData() %>%
  ScaleData()

length(unique(seurat_arm3$Patient_ID))
length(unique(seurat_arm3$Sample))
unique(seurat_arm3$Timepoint)

arm3_tumor <- subset(seurat_arm3, subset = Tissue == "Tumor")
arm3_tumor <- arm3_tumor %>%
  NormalizeData() %>%
  ScaleData()

table(arm3_tumor$Patient_ID)
table(as.character(arm3_tumor$Sample))

arm3_tumor@meta.data %>%
  dplyr::select(c("Patient_ID", "Sample")) %>%
  distinct()

arm3_normal <- subset(seurat_arm3, subset = Tissue == "Normal")
arm3_normal <- arm3_normal %>%
  NormalizeData() %>%
  ScaleData()

table(arm3_normal$Patient_ID)
table(as.character(arm3_normal$Sample))

arm3_normal@meta.data %>%
  dplyr::select(c("Patient_ID", "Sample")) %>%
  distinct()


```

## Tumor cell type proportion differences between timepoints (before and after treatment)

```{r, message = F, warning = F, fig.width = 4, fig.height = 5}

arm3_tumor$PrePost <- ifelse(arm3_tumor$Timepoint == 0, "Baseline", "Treated")

create_barplot(arm3_tumor,
               group_var = "PrePost",
               plot_var = "Annotation",
               plot_levels = sort((unique(arm3_tumor$Annotation))),
               group_levels = sort(unique(arm3_tumor$PrePost)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6}

prop_test <- sc_utils(arm3_tumor)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Annotation",
	sample_1 = "Baseline", sample_2 = "Treated",
	sample_identity = "PrePost")

permutation_plot(prop_test) +
  ggtitle("Baseline vs. treated")

```

## Tumor-adjacent cell type proportion differences between timepoints (before and after treatment)

```{r, message = F, warning = F, fig.width = 4, fig.height = 5}

arm3_normal$PrePost <- ifelse(arm3_normal$Timepoint == 0, "Baseline", "Treated")

create_barplot(arm3_normal,
               group_var = "PrePost",
               plot_var = "Annotation",
               plot_levels = sort((unique(arm3_normal$Annotation))),
               group_levels = sort(unique(arm3_normal$PrePost)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6}

prop_test <- sc_utils(arm3_normal)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Annotation",
	sample_1 = "Baseline", sample_2 = "Treated",
	sample_identity = "PrePost")

permutation_plot(prop_test) +
  ggtitle("Baseline vs. treated")

```

## DEGs

### Tumor

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

tumor_DEGs <- get_DEGs(seuratdata = arm3_tumor,
                       celltypes = unique(arm3_tumor$Annotation),
                       groupvar = "PrePost",
                       group1 = "Treated",
                       group2 = "Baseline")

# Distribution of p-values and log2FC
hist(tumor_DEGs$p_val)
hist(tumor_DEGs$p_val_adj)
hist(tumor_DEGs$avg_log2FC)

tumor_DEGs_sig <- tumor_DEGs %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

table(tumor_DEGs_sig$celltype)

```

### Tumor-adjacent

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

normal_DEGs <- get_DEGs(seuratdata = arm3_normal,
                        celltypes = unique(arm3_normal$Annotation),
                        groupvar = "PrePost",
                        group1 = "Treated",
                        group2 = "Baseline")

# Distribution of p-values and log2FC
hist(normal_DEGs$p_val)
hist(normal_DEGs$p_val_adj)
hist(normal_DEGs$avg_log2FC)

normal_DEGs_sig <- normal_DEGs %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

table(normal_DEGs_sig$celltype)

```

#### Plotting

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

#plot_list <- lapply(unique(seurat_arm3$Annotation))

```

### DEGs for myeloid, lymphoid, and fibroblast subsets

#### Tumors, myeloid

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

sort(unique(arm3_tumor$Annotation))

Idents(arm3_tumor) <- arm3_tumor$PrePost

myel_celltypes <- c("B1", "B2", "DC", "Mast", "Plasma1", "Plasma2", "Plasma3",
                    paste0("M", seq(1,9)))

tumor_myel <- subset(arm3_tumor, subset = Annotation %in% myel_celltypes)
tumor_myel <- tumor_myel %>%
  NormalizeData() %>%
  ScaleData()

degs_tumor_myel <- FindMarkers(arm3_tumor,
                               ident.1 = "Treated",
                               ident.2 = "Baseline",
                               assay = "RNA",
                               verbose = F)
degs_tumor_myel$feature <- rownames(degs_tumor_myel)

# Distribution of p-values and log2FC
hist(degs_tumor_myel$p_val)
hist(degs_tumor_myel$p_val_adj)
hist(degs_tumor_myel$avg_log2FC)

degs_tumor_myel_sig <- degs_tumor_myel %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

degs_tumor_myel_sig$feature

degs_tumor_myel_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 8, eval = F}

VlnPlot(tumor_myel,
        group.by = "PrePost",
        features = degs_tumor_myel_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8, eval = F}

VlnPlot(tumor_myel,
        group.by = "Annotation",
        split.by = "PrePost",
        split.plot = T,
        features = degs_tumor_myel_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Tumors, lymphoid

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

lymph_celltypes <- paste0("L", seq(1,5))

tumor_lymph <- subset(arm3_tumor, subset = Annotation %in% lymph_celltypes)
tumor_lymph <- tumor_lymph %>%
  NormalizeData() %>%
  ScaleData()

degs_tumor_lymph <- FindMarkers(tumor_lymph,
                           ident.1 = "Treated",
                           ident.2 = "Baseline",
                           assay = "RNA",
                           verbose = F)
degs_tumor_lymph$feature <- rownames(degs_tumor_lymph)

# Distribution of p-values and log2FC
hist(degs_tumor_lymph$p_val)
hist(degs_tumor_lymph$p_val_adj)
hist(degs_tumor_lymph$avg_log2FC)

degs_tumor_lymph_sig <- degs_tumor_lymph %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

degs_tumor_lymph_sig$feature

degs_tumor_lymph_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 8, eval = F}

VlnPlot(tumor_lymph,
        group.by = "PrePost",
        features = degs_tumor_lymph_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8, eval = F}

VlnPlot(tumor_lymph,
        group.by = "Annotation",
        split.by = "PrePost",
        split.plot = T,
        features = degs_tumor_lymph_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Tumors, fibroblast

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

fibro_celltypes <- paste0("F", seq(1,5))

tumor_fibro <- subset(arm3_tumor, subset = Annotation %in% fibro_celltypes)
tumor_fibro <- tumor_fibro %>%
  NormalizeData() %>%
  ScaleData()

degs_tumor_fibro <- FindMarkers(tumor_fibro,
                                ident.1 = "Treated",
                                ident.2 = "Baseline",
                                assay = "RNA",
                                verbose = F)
degs_tumor_fibro$feature <- rownames(degs_tumor_fibro)

# Distribution of p-values and log2FC
hist(degs_tumor_fibro$p_val)
hist(degs_tumor_fibro$p_val_adj)
hist(degs_tumor_fibro$avg_log2FC)

degs_tumor_fibro_sig <- degs_tumor_fibro %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

degs_tumor_fibro_sig$feature

degs_tumor_fibro_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 14, eval = F}

VlnPlot(tumor_fibro,
        group.by = "PrePost",
        features = degs_tumor_fibro_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 14, eval = F}

VlnPlot(tumor_fibro,
        group.by = "Annotation",
        split.by = "PrePost",
        split.plot = T,
        features = degs_tumor_fibro_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Tumor-adjacent, myeloid

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

sort(unique(arm3_normal$Annotation))

Idents(arm3_normal) <- arm3_normal$PrePost

myel_celltypes <- c("B1", "B2", "DC", "Mast", "Plasma1", "Plasma2", "Plasma3",
                    paste0("M", seq(1,9)))

normal_myel <- subset(arm3_normal, subset = Annotation %in% myel_celltypes)
normal_myel <- normal_myel %>%
  NormalizeData() %>%
  ScaleData()

degs_normal_myel <- FindMarkers(normal_myel,
                                ident.1 = "Treated",
                                ident.2 = "Baseline",
                                assay = "RNA",
                                verbose = F)
degs_normal_myel$feature <- rownames(degs_normal_myel)

# Distribution of p-values and log2FC
hist(degs_normal_myel$p_val)
hist(degs_normal_myel$p_val_adj)
hist(degs_normal_myel$avg_log2FC)

degs_normal_myel_sig <- degs_normal_myel %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

degs_normal_myel_sig$feature

degs_normal_myel_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 8, eval = F}

VlnPlot(normal_myel,
        group.by = "PrePost",
        features = degs_normal_myel_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8, eval = F}

VlnPlot(normal_myel,
        group.by = "Annotation",
        split.by = "PrePost",
        split.plot = T,
        features = degs_normal_myel_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Tumor-adjacent, lymphoid

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

lymph_celltypes <- paste0("L", seq(1,5))

normal_lymph <- subset(arm3_normal, subset = Annotation %in% lymph_celltypes)
normal_lymph <- normal_lymph %>%
  NormalizeData() %>%
  ScaleData()

degs_normal_lymph <- FindMarkers(normal_lymph,
                                 ident.1 = "Treated",
                                 ident.2 = "Baseline",
                                 assay = "RNA",
                                 verbose = F)
degs_normal_lymph$feature <- rownames(degs_normal_lymph)

# Distribution of p-values and log2FC
hist(degs_normal_lymph$p_val)
hist(degs_normal_lymph$p_val_adj)
hist(degs_normal_lymph$avg_log2FC)

degs_normal_lymph_sig <- degs_normal_lymph %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

degs_normal_lymph_sig$feature

degs_normal_lymph_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 8, eval = F}

VlnPlot(normal_lymph,
        group.by = "PrePost",
        features = degs_normal_lymph_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8, eval = F}

VlnPlot(normal_lymph,
        group.by = "Annotation",
        split.by = "PrePost",
        split.plot = T,
        features = degs_normal_lymph_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

#### Tumor-adjacent, fibroblast

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

fibro_celltypes <- paste0("F", seq(1,5))

normal_fibro <- subset(arm3_normal, subset = Annotation %in% fibro_celltypes)
normal_fibro <- normal_fibro %>%
  NormalizeData() %>%
  ScaleData()

degs_normal_fibro <- FindMarkers(normal_fibro,
                                 ident.1 = "Treated",
                                 ident.2 = "Baseline",
                                 assay = "RNA",
                                 verbose = F)
degs_normal_fibro$feature <- rownames(degs_normal_fibro)

# Distribution of p-values and log2FC
hist(degs_normal_fibro$p_val)
hist(degs_normal_fibro$p_val_adj)
hist(degs_normal_fibro$avg_log2FC)

degs_normal_fibro_sig <- degs_normal_fibro %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) > 1.5)

degs_normal_fibro_sig$feature

degs_normal_fibro_sig %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 12, eval = F}

VlnPlot(normal_fibro,
        group.by = "PrePost",
        features = degs_normal_fibro_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 12, eval = F}

VlnPlot(normal_fibro,
        group.by = "Annotation",
        split.by = "PrePost",
        split.plot = T,
        features = degs_normal_fibro_sig$feature,
        layer = "data",
        pt.size = 0,
        ncol = 3) &
  theme_classic() &
  RotatedAxis()

```


