---
title: "PIPAC Arm 3 comparative analysis"
author: "heinin"
date: "2025-03-19"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Load packages

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  library(workflowr)
  library(arrow)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(googlesheets4)
  library(workflowr)
  library(patchwork)
  library(ggpmisc)
  library(scProportionTest)})

```

## Environment variables and helper functions

```{r, message = F, warning = F}

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(9999)
options(scipen = 99999)
options(ggrepel.max.overlaps = Inf)

source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

```

## Marker information

```{r, message = F, warning = F, fig.width = 6, fig.height = 4}

gs4_deauth()
metadata  <- gs4_get("https://docs.google.com/spreadsheets/d/1sXXwOreLxjMSUoPt79c6jmaQpluWkaxA5P5HfDsed3I/edit?usp=sharing")
markers <- read_sheet(metadata, sheet = "Markers")
second_pass_nonimmune <- read_sheet(metadata, sheet = "Tumor 2nd pass")
second_pass_immune <- read_sheet(metadata, sheet = "Immune 2nd pass")

# Cell type markers by group
gs4_deauth()
markers_by_group  <- gs4_get("https://docs.google.com/spreadsheets/d/1w0wrL-5KwNEWi_VpmriN7YmfwpBPlEYX0MKlwI2ZQu8/edit?usp=sharing")
markers_by_group <- read_sheet(markers_by_group, sheet = "Markers by group")

```

## Import and subset data

```{r, message = F, warning = F, fig.width = 8, fig.height = 4}

seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/cell_merged_spatial_filtered_splitsamples_clustered_NN30_PC50_Seurat_denoIST_annotated_updated.rds")

# Colors
celltype_final <- as.factor(sort(unique(seurat_data$Celltype_Final)))
celltype_final_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(celltype_final))
names(celltype_final_col) <- levels(celltype_final)

pipac_celltype_col <- celltype_final_col

DefaultAssay(seurat_data) <- "RNA"
seurat_data <- seurat_data %>%
  NormalizeData() %>%
  ScaleData()

seurat_arm3 <- subset(seurat_data, subset = Arm == "Arm3")
seurat_arm3 <- seurat_arm3 %>%
  NormalizeData() %>%
  ScaleData()

# pre tumor cells: 3296+3296
# post: 829
# ((3296+3296)-829)/(3296+3296)

keeps <- table(seurat_arm3$Celltype_Final) %>%
  as.data.frame() %>%
  filter(Freq> 500)

seurat_arm3_subset <- subset(seurat_arm3, subset = Celltype_Final %in% keeps$Var1)

table(seurat_arm3_subset$Celltype_Final)
table(seurat_arm3_subset$Celltype_Final, seurat_arm3_subset$Tissue)

```

## All tumors

### Cell type proportions by sample and sample type

```{r, message = F, warning = F, fig.width = 24, fig.height = 6}

create_barplot(seurat_arm3_subset,
               group_var = "Sample",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3_subset$Celltype_Final))),
               group_levels = sort(unique(seurat_arm3_subset$Sample)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

### Cell type proportions by timepoint in tumors

```{r, message = F, warning = F, fig.width = 5, fig.height = 5}

create_barplot(subset(seurat_arm3_subset, subset = Tissue == "Tumor"),
               group_var = "Timepoint",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3_subset$"Celltype_Final"))),
               group_levels = sort(unique(seurat_arm3_subset$Timepoint)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

### Cell type proportions by timepoint in tumor-adjacent samples

```{r, message = F, warning = F, fig.width = 5, fig.height = 5}

create_barplot(subset(seurat_arm3_subset, subset = Tissue == "Normal"),
               group_var = "Timepoint",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3_subset$"Celltype_Final"))),
               group_levels = sort(unique(seurat_arm3_subset$Timepoint)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

### Cell numbers in tumor and normal across all samples

```{r, message = F, warning = F, fig.width = 5, fig.height = 5}

ct_table <- as.data.frame(table(seurat_arm3_subset$"Celltype_Final", seurat_arm3_subset$Tissue))
colnames(ct_table) <- c("Celltype", "Tissue", "Freq")
ct_table <- spread(ct_table, Celltype, Freq)
# Converting to percetange 
#ct_table[,2:length(ct_table)] <- (ct_table[,2:length(ct_table)]/rowSums(ct_table[,2:length(ct_table)]))*100
ct_table <- gather(ct_table, Celltype, Freq, names(ct_table)[2:length(names(ct_table))], factor_key=TRUE)
    
ct_table$Tissue <- factor(ct_table$Tissue, levels = c("Normal", "Tumor"))

ggplot(ct_table, aes(x = Tissue, y = Freq, fill = Celltype)) +
       geom_bar(stat="identity", position='stack', width = 0.8) +
       scale_fill_manual(name = "Celltype", values = pipac_celltype_col) +
       xlab("") +
       ylab("Cell count") +
       theme_classic() +
       theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

```

### Subsetting arm 3 tumor and tumor adjacent

```{r, message = F, warning = F, fig.width = 4, fig.height = 5, eval = T}

arm3_tumor <- subset(seurat_arm3_subset, subset = Tissue == "Tumor")
arm3_tumor <- arm3_tumor %>%
  NormalizeData() %>%
  ScaleData()

arm3_normal <- subset(seurat_arm3_subset, subset = Tissue == "Normal")
arm3_normal <- arm3_normal %>%
  NormalizeData() %>%
  ScaleData()

```

### Tumor cell type proportions by response

```{r, message = F, warning = F, fig.width = 4, fig.height = 5, eval = F}

table(seurat_arm3_subset$Patient_ID)
table(as.character(seurat_arm3_subset$Sample))

seurat_arm3_subset@meta.data %>%
  dplyr::select(c("Patient_ID", "Sample")) %>%
  distinct()

table(seurat_arm3_subset$BestResponse)

create_barplot(seurat_arm3_subset,
               group_var = "BestResponse",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(arm3_tumor$"Celltype_Final"))),
               group_levels = sort(unique(arm3_tumor$BestResponse)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

#### Testing for significance with scProportionTest

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = F}

prop_test <- sc_utils(seurat_arm3_subset)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "PR", sample_2 = "SD",
	sample_identity = "BestResponse")

permutation_plot(prop_test) +
  ggtitle("All tumors")

```

### Tumor-adjacent cell type proportions by response

```{r, message = F, warning = F, fig.width = 4, fig.height = 5, eval = F}

table(seurat_arm3_subset$Patient_ID)
table(as.character(seurat_arm3_subset$Sample))

seurat_arm3_subset@meta.data %>%
  dplyr::select(c("Patient_ID", "Sample")) %>%
  distinct()

table(seurat_arm3_subset$BestResponse)

create_barplot(seurat_arm3_subset,
               group_var = "BestResponse",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3_subset$"Celltype_Final"))),
               group_levels = sort(unique(seurat_arm3_subset$BestResponse)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

#### Testing for significance with scProportionTest

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = F}

prop_test <- sc_utils(arm3_normal)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "PR", sample_2 = "SD",
	sample_identity = "BestResponse")

permutation_plot(prop_test) +
  ggtitle("All tumors")

```

### Cell type proportion correlations with survival time

Linear regression

```{r, message = F, warning = F, fig.width = 12, fig.height = 30, eval = F}

metadata <- seurat_arm3_subset@meta.data %>%
  dplyr::select("Patient_ID", "DISEASESITE", "osmonths", "pfsmonths", "No_of_PIPACs") %>%
  distinct()

prop_table <- as.data.frame(table(seurat_arm3_subset@meta.data$"Celltype_Final", as.character(seurat_arm3@meta.data$Patient_ID)))
colnames(prop_table) <- c("Celltype_Final", "Patient_ID", "Freq")
prop_table <- spread(prop_table, Celltype_Final, Freq)
# Converting to percentage
prop_table[,2:length(prop_table)] <- (prop_table[,2:length(prop_table)]/rowSums(prop_table[,2:length(prop_table)]))*100
prop_table <- gather(prop_table, Celltype_Final, Freq, names(prop_table)[2:length(names(prop_table))], factor_key=TRUE)

# Adding survival data
prop_survival <- merge(prop_table, metadata, by = "Patient_ID")
prop_survival$osmonths <- unlist(prop_survival$osmonths)
prop_survival$pfmonths <- unlist(prop_survival$pfmonths)

# Plotting
plot_list <- lapply(sort(unique(seurat_arm3_subset$"Celltype_Final")), function(ct){
  prop_survival %>%
    filter("Celltype_Final" == ct) %>%
    ggplot(aes(x = Freq, y = osmonths, color = "Celltype_Final")) +
    geom_point() +
    scale_color_manual(name = "Celltype_Final", values = pipac_celltype_col) +
    stat_poly_line() +
    stat_poly_eq(formula = y ~ x,
                 rr.digits = 3, p.digits = 3, small.p = T,
                 aes(label = paste(after_stat(rr.label), after_stat(p.value.label), sep = "*`,`~")),
                 label.x = 0.01, label.y = 1, parse = T) +
    ylab("Overall Survival") +
    xlab(paste0("% of ", ct, " cells")) +
    theme_classic() +
    theme(axis.text.x = element_text(color = "black"),
          axis.text.y = element_text(color = "black")) +
    NoLegend()
})

wrap_plots(plot_list, ncol = 3)

names(plot_list) <- sort(unique(seurat_arm3_subset$"Celltype_Final"))

```

### Fibroblast marker expression

```{r, message = F, warning = F, fig.width = 6, fig.height = 5, eval = F}

DotPlot(subset(seurat_arm3_subset, subset = Celltype_Final %in% grep("^F", seurat_arm3_subset$Celltype_Final, value = T),
        group.by = "Celltype_Final",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "CAFs"),]$feature),
        cols = c("azure", "tomato3")) +
  coord_flip() +
  ggtitle("CAFs, all timepoints")

```

## Baseline tumors

### Cell type proportions by sample

```{r, message = F, warning = F, fig.width = 8, fig.height = 5}

unique(seurat_arm3_subset$Timepoint)
baseline_tumor <- subset(seurat_arm3_subset, subset = Tissue == "Tumor" &
                           Timepoint == 0)
baseline_tumor <- baseline_tumor %>%
  NormalizeData() %>%
  ScaleData()

table(baseline_tumor$Patient_ID)
table(as.character(baseline_tumor$Sample))

baseline_tumor@meta.data %>%
  dplyr::select(c("Patient_ID", "Sample")) %>%
  distinct()

table(baseline_tumor$BestResponse)

create_barplot(baseline_tumor,
               group_var = "Sample",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(baseline_tumor$"Celltype_Final"))),
               group_levels = sort(unique(baseline_tumor$Sample)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

### Cell type proportions by response

```{r, message = F, warning = F, fig.width = 4, fig.height = 5, eval = F}

create_barplot(baseline_tumor,
               group_var = "BestResponse",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(baseline_tumor$"Celltype_Final"))),
               group_levels = sort(unique(baseline_tumor$BestResponse)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

#### Testing for significance with scProportionTest

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = F}

prop_test <- sc_utils(baseline_tumor)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "PR", sample_2 = "SD",
	sample_identity = "BestResponse")

permutation_plot(prop_test) +
  ggtitle("Baseline tumors")

```

### Cell type proportions by response in tumor-adjacent

```{r, message = F, warning = F, fig.width = 4, fig.height = 5, eval = F}

baseline_normal <- subset(seurat_arm3_subset, subset = Tissue == "Normal" &
                           Timepoint == 0)
baseline_normal <- baseline_normal %>%
  NormalizeData() %>%
  ScaleData()

table(baseline_normal$Patient_ID)
table(as.character(baseline_normal$Sample))

baseline_normal@meta.data %>%
  dplyr::select(c("Patient_ID", "Sample")) %>%
  distinct()

table(baseline_normal$BestResponse)
table(baseline_normal$"Celltype_Final")

create_barplot(baseline_normal,
               group_var = "BestResponse",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(baseline_normal$"Celltype_Final"))),
               group_levels = sort(unique(baseline_normal$BestResponse)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

#### Testing for significance with scProportionTest

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = F}

prop_test <- sc_utils(baseline_normal)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "PR", sample_2 = "SD",
	sample_identity = "BestResponse")

permutation_plot(prop_test) +
  ggtitle("Baseline tumor-adjacent")

```

### Tumor cell type proportion correlations with survival time

Linear regression

```{r, message = F, warning = F, fig.width = 12, fig.height = 30, eval = F}

metadata <- baseline_tumor@meta.data %>%
  dplyr::select("Patient_ID", "DISEASESITE", "osmonths", "pfsmonths", "No_of_PIPACs") %>%
  distinct()

prop_table <- as.data.frame(table(baseline_tumor@meta.data$"Celltype_Final", as.character(baseline_tumor@meta.data$Patient_ID)))
colnames(prop_table) <- c("Celltype_Final", "Patient_ID", "Freq")
prop_table <- spread(prop_table, Celltype_Final, Freq)
# Converting to percentage
prop_table[,2:length(prop_table)] <- (prop_table[,2:length(prop_table)]/rowSums(prop_table[,2:length(prop_table)]))*100
prop_table <- gather(prop_table, Celltype_Final, Freq, names(prop_table)[2:length(names(prop_table))], factor_key=TRUE)

# Adding survival data
prop_survival <- merge(prop_table, metadata, by = "Patient_ID")
prop_survival$osmonths <- unlist(prop_survival$osmonths)
prop_survival$pfmonths <- unlist(prop_survival$pfmonths)

# Plotting
plot_list <- lapply(sort(unique(baseline_tumor$"Celltype_Final")), function(ct){
  prop_survival %>%
    filter("Celltype_Final" == ct) %>%
    ggplot(aes(x = Freq, y = osmonths, color = "Celltype_Final")) +
    geom_point() +
    scale_color_manual(name = "Celltype_Final", values = pipac_celltype_col) +
    stat_poly_line() +
    stat_poly_eq(formula = y ~ x,
                 rr.digits = 3, p.digits = 3, small.p = T,
                 aes(label = paste(after_stat(rr.label), after_stat(p.value.label), sep = "*`,`~")),
                 label.x = 0.01, label.y = 1, parse = T) +
    ylab("Overall Survival") +
    xlab(paste0("% of ", ct, " cells")) +
    theme_classic() +
    theme(axis.text.x = element_text(color = "black"),
          axis.text.y = element_text(color = "black")) +
    NoLegend()
})

wrap_plots(plot_list, ncol = 3)

```

## After treatment tumor

### Cell type proportions by sample

```{r, message = F, warning = F, fig.width = 12, fig.height = 5}

unique(seurat_arm3_subset$Timepoint)
post_treatment_tumor <- subset(seurat_arm3_subset, subset = Tissue == "Tumor" &
                                 Timepoint %in% c(6, 12))
post_treatment_tumor <- post_treatment_tumor %>%
  NormalizeData() %>%
  ScaleData()

table(post_treatment_tumor$Patient_ID)
table(as.character(post_treatment_tumor$Sample))

post_treatment_tumor@meta.data %>%
  dplyr::select(c("Patient_ID", "Sample")) %>%
  distinct()

table(post_treatment_tumor$BestResponse)

create_barplot(post_treatment_tumor,
               group_var = "Sample",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(post_treatment_tumor$"Celltype_Final"))),
               group_levels = sort(unique(post_treatment_tumor$Sample)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

### Cell type proportions by response

```{r, message = F, warning = F, fig.width = 4, fig.height = 5, eval = F}

create_barplot(post_treatment_tumor,
               group_var = "BestResponse",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(post_treatment_tumor$"Celltype_Final"))),
               group_levels = sort(unique(post_treatment_tumor$BestResponse)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

#### Testing for significance with scProportionTest

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = F}

prop_test <- sc_utils(post_treatment_tumor)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "PR", sample_2 = "SD",
	sample_identity = "BestResponse")

permutation_plot(prop_test) +
  ggtitle("Post-treatment tumors")

```

### Cell type proportions by response in tumor-adjacent

```{r, message = F, warning = F, fig.width = 4, fig.height = 5, eval = F}

unique(seurat_arm3$Timepoint)
treated_normal <- subset(seurat_arm3, subset = Tissue == "Normal" &
                           Timepoint == c(6, 12))
treated_normal <- treated_normal %>%
  NormalizeData() %>%
  ScaleData()

table(treated_normal$Patient_ID)
table(as.character(treated_normal$Sample))

treated_normal@meta.data %>%
  dplyr::select(c("Patient_ID", "Sample")) %>%
  distinct()

table(treated_normal$BestResponse)
table(treated_normal$"Celltype_Final")

create_barplot(treated_normal,
               group_var = "BestResponse",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(treated_normal$"Celltype_Final"))),
               group_levels = sort(unique(treated_normal$BestResponse)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

#### Testing for significance with scProportionTest

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = F}

prop_test <- sc_utils(treated_normal)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "PR", sample_2 = "SD",
	sample_identity = "BestResponse")

permutation_plot(prop_test) +
  ggtitle("Post-treatment tumor-adjacent")

```

#### Tumor cell type proportion correlations with survival time

Linear regression

```{r, message = F, warning = F, fig.width = 12, fig.height = 30, eval = F}

metadata <- post_treatment_tumor@meta.data %>%
  dplyr::select("Patient_ID", "DISEASESITE", "osmonths", "pfsmonths", "No_of_PIPACs") %>%
  distinct()

prop_table <- as.data.frame(table(post_treatment_tumor@meta.data$"Celltype_Final", as.character(post_treatment_tumor@meta.data$Patient_ID)))
colnames(prop_table) <- c("Celltype_Final", "Patient_ID", "Freq")
prop_table <- spread(prop_table, Celltype_Final, Freq)
# Converting to percentage
prop_table[,2:length(prop_table)] <- (prop_table[,2:length(prop_table)]/rowSums(prop_table[,2:length(prop_table)]))*100
prop_table <- gather(prop_table, Celltype_Final, Freq, names(prop_table)[2:length(names(prop_table))], factor_key=TRUE)

# Adding survival data
prop_survival <- merge(prop_table, metadata, by = "Patient_ID")
prop_survival$osmonths <- unlist(prop_survival$osmonths)
prop_survival$pfmonths <- unlist(prop_survival$pfmonths)

# Plotting
plot_list <- lapply(sort(unique(post_treatment_tumor$"Celltype_Final")), function(ct){
  prop_survival %>%
    filter("Celltype_Final" == ct) %>%
    ggplot(aes(x = Freq, y = osmonths, color = "Celltype_Final")) +
    geom_point() +
    scale_color_manual(name = "Celltype_Final", values = pipac_celltype_col) +
    stat_poly_line() +
    stat_poly_eq(formula = y ~ x,
                 rr.digits = 3, p.digits = 3, small.p = T,
                 aes(label = paste(after_stat(rr.label), after_stat(p.value.label), sep = "*`,`~")),
                 label.x = 0.01, label.y = 1, parse = T) +
    ylab("Overall Survival") +
    xlab(paste0("% of ", ct, " cells")) +
    theme_classic() +
    theme(axis.text.x = element_text(color = "black"),
          axis.text.y = element_text(color = "black")) +
    NoLegend()
})

wrap_plots(plot_list, ncol = 3)

names(plot_list) <- sort(unique(post_treatment_tumor$"Celltype_Final"))

```

```{r, message = F, warning = F, fig.width = 4, fig.height = 3, eval = F}

plot_list[["F5"]]

```

Calculating Hazard ratios

```{r, message = F, warning = F, fig.width = 5, fig.height = 5, eval = F}

prop_table <- as.data.frame(table(immune_fibro@meta.data$celltype, as.character(immune_fibro@meta.data$UPN)))
colnames(prop_table) <- c("celltype", "UPN", "Freq")
prop_table <- spread(prop_table, celltype, Freq)
# Converting to percentage
prop_table[,2:length(prop_table)] <- (prop_table[,2:length(prop_table)]/rowSums(prop_table[,2:length(prop_table)]))*100
prop_table <- gather(prop_table, celltype, Freq, names(prop_table)[2:length(names(prop_table))], factor_key=TRUE)

prop_table <- prop_table %>% pivot_wider(names_from = celltype,
                                         values_from = Freq)

# Adding survival data
metadata <- immune_fibro@meta.data %>%
  dplyr::select("UPN",
                "Death.Status",
                "Survival.time.in.Months.from.surgery",
                "binary_response",
                "Grade",
                "Diagnosis.Histology",
                "EGFR.Amplification",
                "EGFRvIII..Exon.2.7.deletion",
                "EGFR.Missense",
                "TP53.Frameshift",
                "TP53.Missense",
                "TP53.Nonsense",
                "IDH1.Missense",
                "IDH2.Missense",
                "TERT.Promoter.Mutation",
                "PTEN.Codon.Deletion",
                "PTEN.Frameshift",
                "PTEN.Interference.of.splice.acceptor.site.in.Intron.6") %>%
  distinct()

prop_survival <- merge(prop_table, metadata, by = "UPN")

# Fit survival data using the Kaplan-Meier method
surv_object <- Surv(time = prop_survival$Survival.time.in.Months.from.surgery,
                    event = prop_survival$Death.Status)
surv_object 

# Fit a Cox proportional hazards model for a single gene
fit.coxph <- coxph(surv_object ~ M1 + M2 + M3 + M4 + M5 + M6 + M7 + M8 + M9 + N1 +
                     L1 + L2 + L3 + L4 + L5 + L6 + L7 + L8 + L8 + L9 + L10 + B1 +
                     F1 + F2 + F3,
                   data = prop_survival)

p1 <- ggforest(fit.coxph, data = prop_survival)

p1

```

### Inspecting F5

#### Fibroblast marker expression

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = F}

DotPlot(subset(post_treatment_tumor, subset = Celltype_Final %in% paste0("F", seq(1,5))),
        group.by = "Celltype_Final",
        features = c(unique(markers_by_group[which(markers_by_group$annotation1 == "CAFs"),]$feature), "DCN", "LUM", "EPCAM"),
        cols = c("gray88", "tomato3")) +
  coord_flip() +
  ggtitle("CAFs, post-treatment (D6/12)")

#Idents(seurat_data) <- seurat_data$"Celltype_Final"
#celltype_markers <- FindAllMarkers(seurat_data,
#                                   return.thresh = 0.01,
#                                   logfc.threshold = 0.5,
#                                   min.pct = 0.20,
#                                   only.pos = T,
#                                   verbose = F)
#
#table(celltype_markers$cluster)
#
#output_celltype_markers <- celltype_markers %>%
#  arrange(dplyr::desc(abs(avg_log2FC))) %>%
#  group_by(cluster) %>%
#  dplyr::slice(1:30)
#
#output_celltype_markers <- merge(output_celltype_markers, markers, by.x = "gene", by.y = "Gene")

```

## Tumor cell type proportion differences between timepoints (before and after treatment)

```{r, message = F, warning = F, fig.width = 4, fig.height = 5}

arm3_tumor$PrePost <- ifelse(arm3_tumor$Timepoint == 0, "Baseline", "Treated")

create_barplot(arm3_tumor,
               group_var = "PrePost",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(arm3_tumor$"Celltype_Final"))),
               group_levels = sort(unique(arm3_tumor$PrePost)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6}

prop_test <- sc_utils(arm3_tumor)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "Baseline", sample_2 = "Treated",
	sample_identity = "PrePost")

permutation_plot(prop_test) +
  ggtitle("Baseline vs. treated")

```

## Tumor-adjacent cell type proportion differences between timepoints (before and after treatment)

```{r, message = F, warning = F, fig.width = 4, fig.height = 5}

arm3_normal$PrePost <- ifelse(arm3_normal$Timepoint == 0, "Baseline", "Treated")

create_barplot(arm3_normal,
               group_var = "PrePost",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(arm3_normal$"Celltype_Final"))),
               group_levels = sort(unique(arm3_normal$PrePost)),
               plot_colors = pipac_celltype_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6}


prop_test <- sc_utils(arm3_normal)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "Baseline", sample_2 = "Treated",
	sample_identity = "PrePost")

permutation_plot(prop_test) +
  ggtitle("Baseline vs. treated")

# Rscript -e "rmarkdown::render('arm3_comparative_analysis.Rmd')"

```
