---
title: "CAF analysis"
author: "heinin"
date: "2026-02-02"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Load packages

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  library(workflowr)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(googlesheets4)
  library(workflowr)
  library(patchwork)
  library(ggpmisc)
  library(scProportionTest)})

```

## Environment variables and helper functions

```{r, message = F, warning = F}

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(9999)
options(scipen = 99999)
options(ggrepel.max.overlaps = Inf)

source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

```

## Marker information

```{r, message = F, warning = F, fig.width = 6, fig.height = 4}

# Markers
main_plot_features <- c("PTPRC", "CD3D", "CD3E", "CD4", "CD8A", "JAK1", "JAK2", "STAT4", "STAT3", "TIGIT", "GZMB", "SELL", "CD19", "CD68", "CD44", "MARCO", "APOE", "C1QB", "C1QBP", "CD163", "SPP1", "VCAN", "MUC5AC", "NOTCH3", "RGS5", "MS4A1", "PGA5", "PLVAP", "CCL21", "LYVE", "SPARCL1", "FN1", "DCN", "LUM", "COL1A2", "PDPN", "ACTA2", "LMNA", "EGR3", "TP53", "JUN", "FOS", "BRAF", "KIT", "SOX9", "RNF43", "EPCAM", "PECAM", "CEACAM", "TGFB1", "JUNB", "TCF7", "GZMA", "GZMK", "CD47", "CD74", "LYZ", "TREM2", "CD247", "COL5A1", "SPARC", "CD36", "CTLA4", "RGS1", "CCL2", "SELENOP", "LYVE1", "CTSD", "MMP9", "CDK1", "FLT1", "RAMP2", "CCL14", "MCAM", "NRP1", "CEACAM6", "IL32", "TGFBI", "VEGFA", "LAMP1", "IGFBP7", "POSTN", "C1S", "C1R", "CSF1", "NOTCH1", "FOXP3", "IL2RB", "TUBA1B", "S100A9", "C3", "CXCR2", "SDC1", "CD38", "IGHG1", "PRDX4", "ICAM1", "GATA2", "CXCL3")

mycafs <- c("TGFB1", "ACTA2", "TAGLN") #"HOPX", "MYL9"
icafs <- c("IL6", "CXCL12", "CXCL14") #"CXCL13", 
abcafs <- c("CD74", "HLA-DPB1", "HLA-DPB2", "HLA-DQA1", "HLA-DRA", "HLA-DRB1")
ficafs <- c("SFRP2", "SFRP4", "COMP", "CXCL14", "PLA2G2A", "ASPN")

main_plot_features <- c(main_plot_features, mycafs, icafs, abcafs, ficafs)
main_plot_features <- unique(main_plot_features)

gs4_deauth()
metadata  <- gs4_get("https://docs.google.com/spreadsheets/d/1sXXwOreLxjMSUoPt79c6jmaQpluWkaxA5P5HfDsed3I/edit?usp=sharing")
markers <- read_sheet(metadata, sheet = "Markers")
final_annotations <- read_sheet(metadata, sheet = "CT update clean")

# Cell type markers by group
gs4_deauth()
markers_by_group  <- gs4_get("https://docs.google.com/spreadsheets/d/1w0wrL-5KwNEWi_VpmriN7YmfwpBPlEYX0MKlwI2ZQu8/edit?usp=sharing")
markers_by_group <- read_sheet(markers_by_group, sheet = "Markers by group")

```

## Import and subset data

```{r, message = F, warning = F, fig.width = 8, fig.height = 4, eval = T}

seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/Freeze/cell_merged_spatial_filtered_splitsamples_clustered_NN30_PC50_Seurat_denoIST_annotated_updated.rds")

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 4, eval = F}

#DefaultAssay(seurat_data) <- "RNA"
#seurat_data <- seurat_data %>%
#  NormalizeData() %>%
#  ScaleData()

sort(unique(seurat_data$Celltype_Final))

fcts <- grep("^F", unique(seurat_data$Celltype_Final), value = T)

fibro <- subset(seurat_data, subset = Celltype_Final %in% fcts)
fibro <- fibro %>%
  NormalizeData() %>%
  ScaleData()

#saveRDS(fibro, "/scratch/hnatri/PIPAC/fibro_seurat_data.rds")

#as.data.frame(table(seurat_data$Celltype_Final))
#
#write.table(as.data.frame(table(seurat_data$Celltype_Final)),
#            "/home/hnatri/PIPAC_spatial/ncells.tsv",
#            quote = F, row.names = F, sep = "\t")

```

## Marker expression

```{r, message = F, warning = F, fig.width = 4, fig.height = 4}

fibro <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/fibro_clustered_NN20_PC20_Seurat.rds")

DefaultAssay(fibro) <- "RNA"
fibro <- fibro %>%
  NormalizeData() %>%
  ScaleData()

fibro$Celltype_Final <- mapvalues(colnames(fibro),
                                  from = colnames(seurat_data),
                                  to = seurat_data$Celltype_Final)

celltype_final <- as.factor(sort(unique(fibro$Celltype_Final)))
celltype_final_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(celltype_final))
names(celltype_final_col) <- levels(celltype_final)

DimPlot(fibro,
        group.by = "Celltype_Final",
        cols = celltype_final_col,
        reduction = "umap",
        raster = T,
        label = T,
        label.box = T) +
  coord_fixed(ratio = 1) +
  theme_classic() +
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

FeaturePlot(fibro,
            features = c(mycafs, icafs, abcafs, ficafs),
            cols = c("gray98", "tomato3"),
            reduction = "umap",
            raster = T,
            ncol = 3) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 13, fig.height = 8}

FeaturePlot(fibro,
            features = c("nCount_RNA", "nFeature_RNA", "total_counts",
                         "cell_area", "nucleus_area", "nucleus_count"),
            cols = c("gray98", "tomato3"),
            reduction = "umap",
            raster = T,
            ncol = 3) &
  coord_fixed(ratio = 1) &
  theme_bw()

```

```{r, message = F, warning = F, fig.width = 13, fig.height = 6}

create_dotplot_heatmap(seurat_object = fibro,
                       plot_features = unique(c(mycafs, icafs, abcafs, ficafs)),
                       group_var = "Celltype_Final",
                       group_colors = celltype_final_col,
                       column_title = "",
                       row_km = 1,
                       col_km = 6,
                       row.order = NULL,
                       col.order = NULL)

```


```{r, message = F, warning = F, fig.width = 13, fig.height = 26}

create_dotplot_heatmap(seurat_object = fibro,
                       plot_features = main_plot_features,
                       group_var = "Celltype_Final",
                       group_colors = celltype_final_col,
                       column_title = "",
                       row_km = 12,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

### Plotting cell type markers by group

```{r, message = F, warning = F, fig.width = 16, fig.height = 8, eval = T}

DotPlot(fibro,
        group.by = "Celltype_Final",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "Cell-cell crosstalk"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Cell-cell crosstalk")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 8, eval = T}

DotPlot(fibro,
        group.by = "Celltype_Final",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "DNA damage and cell faith"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("DNA damage and cell faith")

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 8, eval = T}

DotPlot(fibro,
        group.by = "Celltype_Final",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "Functional markers"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Functional markers")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 8, eval = T}

DotPlot(fibro,
        group.by = "Celltype_Final",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "Mesothelial"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Mesothelial")

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 8, eval = T}

DotPlot(fibro,
        group.by = "Celltype_Final",
        features = unique(markers_by_group[which(markers_by_group$annotation1 == "CAFs"),]$feature),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("CAFs")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 8, eval = T}

DotPlot(fibro,
        group.by = "Celltype_Final",
        features = grep("COL", rownames(fibro), value = T),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("Collagen")

```

### Cell cycle markers

```{r, message = F, warning = F, fig.width = 6, fig.height = 8, eval = T}

s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

DotPlot(fibro,
        group.by = "Celltype_Final",
        features = c(s.genes, g2m.genes),
        cols = c("azure", "tomato3")) +
  RotatedAxis() +
  ggtitle("S/G2M genes")

```

## Arm 3 analysis

```{r, message = F, warning = F, fig.width = 13, fig.height = 26}

seurat_arm3 <- subset(fibro, subset = Arm == "Arm3")
seurat_arm3 <- seurat_arm3 %>%
  NormalizeData() %>%
  ScaleData()

ncol(seurat_arm3)
table(seurat_arm3$Celltype_Final)

```

Keeping cell types with >500 cells in arm 3

```{r, message = F, warning = F, fig.width = 24, fig.height = 6}

keeps <- table(seurat_arm3$Celltype_Final) %>%
  as.data.frame() %>%
  filter(Freq> 500)

seurat_arm3_subset <- subset(seurat_arm3, subset = Celltype_Final %in% keeps$Var1)

table(seurat_arm3_subset$Celltype_Final)
table(seurat_arm3_subset$Celltype_Final, seurat_arm3_subset$Tissue)

```

## Cell type proportions by sample

```{r, message = F, warning = F, fig.width = 24, fig.height = 6}

create_barplot(seurat_arm3_subset,
               group_var = "Sample",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3_subset$Celltype_Final))),
               group_levels = sort(unique(seurat_arm3_subset$Sample)),
               plot_colors = celltype_final_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

## Cell type proportions by timepoint and sample type

```{r, message = F, warning = F, fig.width = 7, fig.height = 5}

seurat_arm3_subset$Tissue_Timepoint <- paste0(seurat_arm3_subset$Tissue, "_",
                                              seurat_arm3_subset$Timepoint)

create_barplot(seurat_arm3_subset,
               group_var = "Tissue_Timepoint",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3_subset$"Celltype_Final"))),
               group_levels = sort(unique(seurat_arm3_subset$Tissue_Timepoint)),
               plot_colors = celltype_final_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

## Cell type proportions by timepoint in tumors

```{r, message = F, warning = F, fig.width = 5, fig.height = 5}

create_barplot(subset(seurat_arm3_subset, subset = Tissue == "Tumor"),
               group_var = "Timepoint",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3_subset$"Celltype_Final"))),
               group_levels = sort(unique(seurat_arm3_subset$Timepoint)),
               plot_colors = celltype_final_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

prop_test <- sc_utils(subset(seurat_arm3_subset, subset = Tissue == "Tumor"))
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "0", sample_2 = "6",
	sample_identity = "Timepoint")

permutation_plot(prop_test)

```

## Cell type proportions by timepoint in tumor-adjacent samples

```{r, message = F, warning = F, fig.width = 5, fig.height = 5}

create_barplot(subset(seurat_arm3_subset, subset = Tissue == "Normal"),
               group_var = "Timepoint",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3$"Celltype_Final"))),
               group_levels = sort(unique(seurat_arm3$Timepoint)),
               plot_colors = celltype_final_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

prop_test <- sc_utils(subset(seurat_arm3_subset, subset = Tissue == "Normal"))
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "0", sample_2 = "6",
	sample_identity = "Timepoint")

permutation_plot(prop_test)

```

## Cell type proportions by sample type, baseline

```{r, message = F, warning = F, fig.width = 5, fig.height = 5}

create_barplot(subset(seurat_arm3_subset, subset = Timepoint == "0"),
               group_var = "Tissue",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3$"Celltype_Final"))),
               group_levels = sort(unique(seurat_arm3$Tissue)),
               plot_colors = celltype_final_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

prop_test <- sc_utils(subset(seurat_arm3_subset, subset = Timepoint == "0"))
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "Normal", sample_2 = "Tumor",
	sample_identity = "Tissue")

permutation_plot(prop_test)

```

## Cell type proportions by sample type, post-treatment

```{r, message = F, warning = F, fig.width = 5, fig.height = 5}

create_barplot(subset(seurat_arm3_subset, subset = Timepoint == "6"),
               group_var = "Tissue",
               plot_var = "Celltype_Final",
               plot_levels = sort((unique(seurat_arm3_subset$"Celltype_Final"))),
               group_levels = sort(unique(seurat_arm3_subset$Tissue)),
               plot_colors = celltype_final_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

prop_test <- sc_utils(subset(seurat_arm3_subset, subset = Timepoint == "6"))
prop_test <- permutation_test(
	prop_test, cluster_identity = "Celltype_Final",
	sample_1 = "Normal", sample_2 = "Tumor",
	sample_identity = "Tissue")

permutation_plot(prop_test)

```

## DEGs between pre- and post-treatment, tumor

```{r, message = F, warning = F, fig.width = 8, fig.height = 5}

Idents(seurat_arm3) <- seurat_arm3$Timepoint
deg_res <- FindMarkers(subset(seurat_arm3, subset = Tissue == "Tumor"),
                       ident.1 = "0",
                       ident.2 = "6",
                       assay = "RNA",
                       verbose = F)
deg_res$feature <- rownames(deg_res)

deg_res_sig <- deg_res %>%
  mutate_if(is.numeric, ~round(., 2)) %>%
  filter(abs(avg_log2FC) > 1 & p_val_adj < 0.05)

deg_res_sig

which(deg_res_sig$feature %in% unique(c(mycafs, icafs, abcafs, ficafs)))

deg_res %>% mutate_if(is.numeric, ~round(., 2)) %>%
  filter(feature %in% unique(c(mycafs, icafs, abcafs, ficafs)))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 10}

VlnPlot(subset(seurat_arm3, subset = Tissue == "Tumor"),
        group.by = "Timepoint",
        features = deg_res_sig$feature,
        pt.size = 0,
        ncol = 5) &
  theme_bw() &
  NoLegend() &
  xlab("")

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 6}

VlnPlot(subset(seurat_arm3, subset = Tissue == "Tumor"),
        group.by = "Timepoint",
        features = unique(c(mycafs, icafs, abcafs, ficafs)),
        pt.size = 0,
        ncol = 3) &
  theme_bw() &
  NoLegend() &
  xlab("")

```

## DEGs between pre- and post-treatment, tumor-adjacent

```{r, message = F, warning = F, fig.width = 8, fig.height = 5}

deg_res <- FindMarkers(subset(seurat_arm3, subset = Tissue == "Normal"),
                       ident.1 = "0",
                       ident.2 = "6",
                       assay = "RNA",
                       verbose = F)
deg_res$feature <- rownames(deg_res)

deg_res_sig <- deg_res %>%
  mutate_if(is.numeric, ~round(., 2)) %>%
  filter(abs(avg_log2FC) > 1 & p_val_adj < 0.05)

deg_res_sig

which(deg_res_sig$feature %in% unique(c(mycafs, icafs, abcafs, ficafs)))

deg_res %>% mutate_if(is.numeric, ~round(., 2)) %>%
  filter(feature %in% unique(c(mycafs, icafs, abcafs, ficafs)))

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 10}

VlnPlot(subset(seurat_arm3, subset = Tissue == "Normal"),
        group.by = "Timepoint",
        features = deg_res_sig$feature,
        pt.size = 0,
        ncol = 5) &
  theme_bw() &
  NoLegend() &
  xlab("")

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 6}

VlnPlot(subset(seurat_arm3, subset = Tissue == "Normal"),
        group.by = "Timepoint",
        features = unique(c(mycafs, icafs, abcafs, ficafs)),
        pt.size = 0,
        ncol = 3) &
  theme_bw() &
  NoLegend() &
  xlab("")

```

## QC metrics by cell type

```{r, message = F, warning = F, fig.width = 22, fig.height = 30, eval = T}

VlnPlot(fibro,
        features = c("nCount_RNA", "nCount_cell_RNA", "nFeature_RNA",
                     "nFeature_cell_RNA", "transcript_counts", "cell_area",
                     "nucleus_area", "nucleus_count"),
        group.by = "Celltype_Final",
        pt.size = 0,
        ncol = 1,
        cols = celltype_final_col) &
  theme_minimal() &
  RotatedAxis() &
  NoLegend()

```

## Numbers of cells per celltype

```{r, message = F, warning = F, fig.width = 16, fig.height = 3}

ct_table <- as.data.frame(table(fibro$Celltype_Final))
colnames(ct_table) <- c("Celltype_Final", "Freq")

min(ct_table$Freq)

ggplot(ct_table, aes(x = reorder(Celltype_Final, -Freq), y = Freq, fill = Celltype_Final)) +
       geom_bar(stat="identity", position='stack', width = 0.8) +
       scale_fill_manual(name = "Celltype_Final", values = celltype_final_col) +
       xlab("") +
       ylab("Cell count") +
       theme_classic() +
       theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
       NoLegend()

```


