---
title: "Combining and finalizing annotations"
author: "heinin"
date: "2025-11-11"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

### Packages and environment variables

```{r, warning=F, message=F}

suppressPackageStartupMessages({
  library(workflowr)
  library(arrow)
  library(dplyr)
  library(Seurat)
  library(SeuratObject)
  library(SeuratDisk)
  library(tidyverse)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(ggrepel)
  library(googlesheets4)
  library(workflowr)
  library(RColorBrewer)})

```

### Environment variables and helper functions

```{r, message = F, warning = F}

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(9999)
options(scipen = 99999)
options(ggrepel.max.overlaps = Inf)

source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

recluster_seurat <- function(seurat_subset, umap_pcs = 10){
    seurat_subset <- SCTransform(seurat_subset,
                                 vars.to.regress = c("nCount_RNA", "nFeature_RNA"),
                                 verbose = F)
      
    seurat_subset <- RunPCA(seurat_subset,
                            npcs = 50,
                            assay = "SCT",
                            #features = features,
                            approx = F,
                            verbose = F)
    
    integrated_data <- RunUMAP(seurat_subset,
                               reduction = "pca",
                               reduction.name = "umap",
                               dims = 1:umap_pcs,
                               return.model = TRUE,
                            verbose = F)

    integrated_data <- FindNeighbors(integrated_data,
                                     reduction = "pca",
                                     dims = 1:umap_pcs,
                                     graph.name = c("SCTnn",
                                                    "SCTsnn"),
                            verbose = F)

    integrated_data <- FindClusters(integrated_data,
                                    resolution = c(0.1,0.2,0.3,0.5,0.8,1),
                                    graph.name = "SCTsnn",
                            verbose = F)
    
    return(integrated_data)
}

```

### Marker information and metadata

```{r, message = F, warning = F, fig.width = 6, fig.height = 4}

# Cell type markers by group
gs4_deauth()
markers_by_group  <- gs4_get("https://docs.google.com/spreadsheets/d/1w0wrL-5KwNEWi_VpmriN7YmfwpBPlEYX0MKlwI2ZQu8/edit?usp=sharing")
markers_by_group <- read_sheet(markers_by_group, sheet = "Markers by group")

metadata  <- gs4_get("https://docs.google.com/spreadsheets/d/1sXXwOreLxjMSUoPt79c6jmaQpluWkaxA5P5HfDsed3I/edit?usp=sharing")
markers <- read_sheet(metadata, sheet = "Markers")
granular_annotations <- read_sheet(metadata, sheet = "Cell type annotations, all transcripts")
granular_annotations_immune <- read_sheet(metadata, sheet = "Cell type annotations, all transcripts, immune")

```

### Import data

```{r, message = F, warning = F, fig.width = 6, fig.height = 6}

nonimmune <- readRDS("/scratch/hnatri/PIPAC/nonimmune_annotated.rds")
immune <- readRDS("/scratch/hnatri/PIPAC/immune_annotated_complete.rds")
#lymphoid <- readRDS("/scratch/hnatri/PIPAC/lymphoid_annotated.rds")
#myeloid <- readRDS("/scratch/hnatri/PIPAC/myeloid_annotated.rds")

# Fixing preliminary non-immune annotations
nonimmune$combined_annotation <- as.character(nonimmune$combined_annotation)

nonimmune$combined_annotation <- ifelse(is.na(nonimmune$combined_annotation), "NOS",
                                        nonimmune$combined_annotation)

nonimmune$combined_annotation <- ifelse(nonimmune$combined_annotation == "B1", "B1_from_nonimmune",
                                        ifelse(nonimmune$combined_annotation == "L1", "L1_from_nonimmune", nonimmune$combined_annotation))

nonimmune$combined_annotation <- factor(nonimmune$combined_annotation,
                                        levels = sort(unique(nonimmune$combined_annotation)))

# Making some labels unique
immune$combined_annotation <- ifelse(immune$combined_annotation %in% c("F6", "F7", "F4", "B1", "F5", "F1", "F3", "L1"), paste0(immune$combined_annotation, "_from_immune"), immune$combined_annotation)

immune$combined_annotation  <- factor(immune$combined_annotation,
                                      levels = sort(unique(immune$combined_annotation)))

# Normalizing
DefaultAssay(nonimmune) <- "denoist_RNA"
nonimmune <- NormalizeData(nonimmune)

DefaultAssay(immune) <- "denoist_RNA"
immune <- NormalizeData(immune)

max(table(nonimmune$combined_annotation))
min(table(nonimmune$combined_annotation))
max(table(immune$combined_annotation))
min(table(immune$combined_annotation))

```

### Top cluster markers, non-immune

```{r, message = F, warning = F, fig.width = 5, fig.height = 4}

Idents(nonimmune) <- nonimmune$combined_annotation
cluster_markers <- FindAllMarkers(nonimmune,
                                  return.thresh = 0.01,
                                  logfc.threshold = 0.5,
                                  min.pct = 0.20,
                                  verbose = F)

top_cluster_markers <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:15)

write.table(top_cluster_markers, "/home/hnatri/PIPAC_spatial/nonimmune_prelim_annot_topmarkers.tsv",
            sep = "\t", quote = F, row.names = T)

```

```{r, message = F, warning = F, fig.width = 14, fig.height = 14}

plot_features <- c("PTPRC",
                   "CD3D", "CD3E", "CD4", "CD8A", # T cells
                   "STAT4", "STAT3", "TIGIT", "GZMB",
                   "FOXP3", # Tregs
                   "SELL", "CD19", # B cells
                   "CD68", "CD44", "MARCO", "APOE", # Macrophages
                   "C1QB", "C1QBP", "CD163",
                   "SPP1", "VCAN",
                   "MUC5AC", "NOTCH3", "RGS5", "MS4A1", # Lineage markers
                   "PGA5", "PLVAP", "CCL21", "LYVE",
                   "SPARCL1", # Endo/peri
                   "FN1", "DCN", "LUM", # Fibroblasts
                   "COL1A2", "PDPN", "ACTA2", "LMNA",
                   "EGR3", "TP53", "JUN", "KIT", # Tumor
                   "SOX9", "RNF43", "EPCAM", "PECAM", "CEACAM")

DotPlot(nonimmune,
        group.by = "combined_annotation",
        features = plot_features,
        cols = c("gray98", "tomato3")) +
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 22, fig.height = 33}

nonimmune_prelim <- as.factor(sort(unique(nonimmune$combined_annotation)))
nonimmune_prelim_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(nonimmune_prelim))
names(nonimmune_prelim_col) <- levels(nonimmune_prelim)

create_dotplot_heatmap(seurat_object = nonimmune,
                       plot_features = unique(top_cluster_markers$gene),
                       group_var = "combined_annotation",
                       group_colors = nonimmune_prelim_col,
                       column_title = "",
                       row_km = 20,
                       col_km = 20,
                       row.order = NULL,
                       col.order = NULL)

```

### Top cluster markers, immune

```{r, message = F, warning = F, fig.width = 5, fig.height = 4}

immune <- JoinLayers(immune)
Idents(immune) <- immune$combined_annotation
cluster_markers <- FindAllMarkers(immune,
                                  return.thresh = 0.01,
                                  logfc.threshold = 0.5,
                                  min.pct = 0.20,
                                  verbose = F)

top_cluster_markers <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:15)

write.table(top_cluster_markers, "/home/hnatri/PIPAC_spatial/immune_prelim_annot_topmarkers.tsv",
            sep = "\t", quote = F, row.names = T)

```

```{r, message = F, warning = F, fig.width = 20, fig.height = 34}

DotPlot(immune,
        group.by = "combined_annotation",
        features = plot_features,
        cols = c("gray98", "tomato3")) +
  RotatedAxis()

```

```{r, message = F, warning = F, fig.width = 30, fig.height = 33}

immune_prelim <- as.factor(sort(unique(immune$combined_annotation)))
immune_prelim_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(immune_prelim))
names(immune_prelim_col) <- levels(immune_prelim)

create_dotplot_heatmap(seurat_object = immune,
                       plot_features = unique(top_cluster_markers$gene),
                       group_var = "combined_annotation",
                       group_colors = immune_prelim_col,
                       column_title = "",
                       row_km = 24,
                       col_km = 24,
                       row.order = NULL,
                       col.order = NULL)

```

### Adding new granular annotations

```{r, message = F, warning = F, fig.width = 22, fig.height = 22, eval = T}

nonimmune$Annotation_Cell <- mapvalues(nonimmune$combined_annotation,
                                       from = granular_annotations$combined_annotation,
                                       to = granular_annotations$Annotation)

immune$Annotation_Cell <- mapvalues(immune$combined_annotation,
                                    from = granular_annotations_immune$combined_annotation,
                                    to = granular_annotations_immune$Annotation)

```

### Moving immune clusters from nonimmune object

```{r, message = F, warning = F, fig.width = 22, fig.height = 22, eval = F}

move_clusters <- c("Myel3_stromal_ambig",
                   "reccl11_0_F1_from_stromal",
                   "reccl11_3_NOS_from_stromal",
                   "reccl11_5_Myel_from_stromal",
                   "reccl11_6_Myel_from_stromal",
                   "reccl13_0_Myel_from_stromal",
                   "reccl13_1_F1_from_stromal",
                   "reccl13_3_Myel_from_stromal",
                   "reccl13_4_Myel_from_stromal",
                   "reccl13_5_Myel_from_stromal",
                   "reccl13_6_Myel_from_stromal",
                   "Plasma")

immune_from_nonimmune <- subset(nonimmune, subset = combined_annotation %in% move_clusters)

immune@reductions$pca <- NULL
immune@reductions$umap <- NULL
#immune@reductions$sp <- NULL

immune_from_nonimmune@reductions$pca <- NULL
immune_from_nonimmune@reductions$umap <- NULL
#immune_from_nonimmune@reductions$sp <- NULL

immune@meta.data <- immune@meta.data[,c("cell_id", "combined_annotation", "nCount_RNA", "nFeature_RNA", "total_counts", "cell_area", "nucleus_area", "nucleus_count", "Annotation_Cell")]
immune_from_nonimmune@meta.data <- immune_from_nonimmune@meta.data[,c("cell_id", "combined_annotation", "nCount_RNA", "nFeature_RNA", "total_counts", "cell_area", "nucleus_area", "nucleus_count", "Annotation_Cell")]

immune <- merge(immune, immune_from_nonimmune)
nonimmune <- subset(nonimmune, subset = combined_annotation %in% move_clusters, invert = T)

saveRDS(nonimmune, "/scratch/hnatri/PIPAC/nonimmune_annotated_modified.rds")
saveRDS(immune, "/scratch/hnatri/PIPAC/immune_annotated_modified.rds")

```

### Selected markers

```{r, message = T, warning = F, fig.width = 22, fig.height = 22, eval = T}

nonimmune <- readRDS("/scratch/hnatri/PIPAC/nonimmune_annotated_modified.rds")
immune <- readRDS("/scratch/hnatri/PIPAC/immune_annotated_modified.rds")

main_plot_features <- c("PTPRC", "CD3D", "CD3E", "CD4", "CD8A", "JAK1", "JAK2", "STAT4", "STAT3", "TIGIT", "GZMB", "SELL", "CD19", "CD68", "CD44", "MARCO", "APOE", "C1QB", "C1QBP", "CD163", "SPP1", "VCAN", "MUC5AC", "NOTCH3", "RGS5", "MS4A1", "PGA5", "PLVAP", "CCL21", "LYVE", "SPARCL1", "FN1", "DCN", "LUM", "COL1A2", "PDPN", "ACTA2", "LMNA", "EGR3", "TP53", "JUN", "FOS", "BRAF", "KIT", "SOX9", "RNF43", "EPCAM", "PECAM", "CEACAM", "TGFB1", "JUNB", "TCF7", "GZMA", "GZMK", "CD47", "CD74", "LYZ", "TREM2", "CD247", "COL5A1", "SPARC", "CD36", "CTLA4", "RGS1", "CCL2", "SELENOP", "LYVE1", "CTSD", "MMP9", "CDK1", "FLT1", "RAMP2", "CCL14", "MCAM", "NRP1", "CEACAM6", "IL32", "TGFBI", "VEGFA", "LAMP1", "IGFBP7", "POSTN", "C1S", "C1R", "CSF1", "NOTCH1", "FOXP3", "IL2RB", "TUBA1B", "S100A9", "C3", "CXCR2", "SDC1", "CD38", "IGHG1", "PRDX4", "ICAM1", "GATA2", "CXCL3")

mycafs <- c("TGFB1", "ACTA2", "TAGLN", "HOPX", "MYL9")
icafs <- c("IL6", "CXCL12", "CXCL13", "CXCL14")
abcafs <- c("CD74", "HLA-DPB1", "HLA-DPB2", "HLA-DQA1", "HLA-DRA", "HLA-DRB1")
ficafs <- c("SFRP2", "SFRP4", "COMP", "CXCL14", "PLA2G2A", "ASPN")

main_plot_features <- c(main_plot_features, mycafs, icafs, abcafs, ficafs)
main_plot_features <- unique(main_plot_features)

nonimmune_prelim <- factor(sort(unique(nonimmune$Annotation_Cell)),
                           levels = unique(nonimmune$Annotation_Cell))
nonimmune_prelim_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(nonimmune_prelim))
names(nonimmune_prelim_col) <- levels(nonimmune_prelim)

hp <- create_dotplot_heatmap(seurat_object = nonimmune,
                             plot_features = main_plot_features,
                             group_var = "Annotation_Cell",
                             group_colors = nonimmune_prelim_col,
                             column_title = "",
                             row_km = 20,
                             col_km = 20,
                             row.order = NULL,
                             col.order = NULL)

#for (cl in names(column_order(hp))){
#  message(paste0("Cluster "), cl)
#  message(paste0(as.character(unique(nonimmune$Annotation_Cell)[column_order(hp)[[as.character(cl)]]]), #sep = ", "))
#}

```

```{r, message = T, warning = F, fig.width = 26, fig.height = 22, eval = T}

immune$Annotation_Cell <- factor(immune$Annotation_Cell,
                                 levels = sort(unique(immune$Annotation_Cell)))
immune_prelim <- as.factor(sort(unique(immune$Annotation_Cell)))
immune_prelim_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(immune_prelim))
names(immune_prelim_col) <- levels(immune_prelim)

hp <- create_dotplot_heatmap(seurat_object = immune,
                             plot_features = main_plot_features,
                             group_var = "Annotation_Cell",
                             group_colors = immune_prelim_col,
                             column_title = "",
                             row_km = 24,
                             col_km = 24,
                             row.order = NULL,
                             col.order = NULL)

#for (cl in names(column_order(hp))){
#  message(paste0("Cluster "), cl)
#  message(paste0(as.character(unique(immune$Annotation_Cell)[column_order(hp)[[as.character(cl)]]]), sep = #", "))
#}

```

#### Top markers with cell type name

Non-immune

```{r, message = T, warning = F, fig.width = 22, fig.height = 22, eval = T}

nonimmune <- JoinLayers(nonimmune)
Idents(nonimmune) <- nonimmune$Annotation_Cell
cluster_markers <- FindAllMarkers(nonimmune,
                                  return.thresh = 0.01,
                                  logfc.threshold = 0.5,
                                  min.pct = 0.20,
                                  verbose = F)

top_cluster_markers <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:15)

```

```{r, message = T, warning = F, fig.width = 26, fig.height = 30, eval = T}

nonimmune$Annotation_Cell <- factor(nonimmune$Annotation_Cell,
                                    levels = sort(unique(nonimmune$Annotation_Cell)))
nonimmune_prelim <- as.factor(sort(unique(nonimmune$Annotation_Cell)))
nonimmune_prelim_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(nonimmune_prelim))
names(nonimmune_prelim_col) <- levels(nonimmune_prelim)

hp <- create_dotplot_heatmap(seurat_object = nonimmune,
                             plot_features = unique(c(top_cluster_markers$gene, main_plot_features)),
                             group_var = "Annotation_Cell",
                             group_colors = nonimmune_prelim_col,
                             column_title = "",
                             row_km = 24,
                             col_km = 24,
                             row.order = NULL,
                             col.order = NULL)

#for (cl in names(column_order(hp))){
#  message(paste0("Cluster "), cl)
#  message(paste0(as.character(unique(nonimmune$Annotation_Cell)[column_order(hp)[[as.character(cl)]]]), #sep = ", "))
#}

```

Immune

```{r, message = T, warning = F, fig.width = 22, fig.height = 22, eval = T}

immune <- JoinLayers(immune)
Idents(immune) <- immune$Annotation_Cell
cluster_markers <- FindAllMarkers(immune,
                                  return.thresh = 0.01,
                                  logfc.threshold = 0.5,
                                  min.pct = 0.20,
                                  verbose = F)

top_cluster_markers <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:15)

```

```{r, message = T, warning = F, fig.width = 26, fig.height = 30, eval = T}

immune$Annotation_Cell <- factor(immune$Annotation_Cell,
                                 levels = sort(unique(immune$Annotation_Cell)))
immune_prelim <- as.factor(sort(unique(immune$Annotation_Cell)))
immune_prelim_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(immune_prelim))
names(immune_prelim_col) <- levels(immune_prelim)

hp <- create_dotplot_heatmap(seurat_object = immune,
                             plot_features = unique(c(top_cluster_markers$gene, main_plot_features)),
                             group_var = "Annotation_Cell",
                             group_colors = immune_prelim_col,
                             column_title = "",
                             row_km = 24,
                             col_km = 24,
                             row.order = NULL,
                             col.order = NULL)

#for (cl in names(column_order(hp))){
#  message(paste0("Cluster "), cl)
#  message(paste0(as.character(unique(immune$Annotation_Cell)[column_order(hp)[[as.character(cl)]]]), sep = #", "))
#}

```

### QC metrics by cell type

```{r, message = F, warning = F, fig.width = 20, fig.height = 30, eval = T}

VlnPlot(nonimmune,
        features = c("nCount_RNA", "nCount_cell_RNA", "nFeature_RNA",
                     "nFeature_cell_RNA", "transcript_counts", "cell_area",
                     "nucleus_area", "nucleus_count"),
        group.by = "Annotation_Cell",
        pt.size = 0,
        ncol = 1,
        cols = nonimmune_prelim_col) &
  theme_minimal() &
  RotatedAxis() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 30, fig.height = 30, eval = T}

VlnPlot(immune,
        features = c("nCount_RNA", "nCount_cell_RNA", "nFeature_RNA",
                     "nFeature_cell_RNA", "transcript_counts", "cell_area",
                     "nucleus_area", "nucleus_count"),
        group.by = "Annotation_Cell",
        pt.size = 0,
        ncol = 1,
        cols = immune_prelim_col) &
  theme_minimal() &
  RotatedAxis() &
  NoLegend()

```

### Cells per cell type

```{r, message = F, warning = F, fig.width = 14, fig.height = 5, eval = T}

ct_table <- as.data.frame(table(nonimmune$Annotation_Cell))
colnames(ct_table) <- c("Annotation_Cell", "Freq")

min(ct_table$Freq)

ggplot(ct_table, aes(x = reorder(Annotation_Cell, -Freq), y = Freq, fill = Annotation_Cell)) +
       geom_bar(stat="identity", position='stack', width = 0.8) +
       scale_fill_manual(name = "Celltype", values = nonimmune_prelim_col) +
       xlab("") +
       ylab("Cell count") +
       theme_classic() +
       theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
       NoLegend()

```

```{r, message = F, warning = F, fig.width = 20, fig.height = 5, eval = T}

ct_table <- as.data.frame(table(immune$Annotation_Cell))
colnames(ct_table) <- c("Annotation_Cell", "Freq")

min(ct_table$Freq)

ggplot(ct_table, aes(x = reorder(Annotation_Cell, -Freq), y = Freq, fill = Annotation_Cell)) +
       geom_bar(stat="identity", position='stack', width = 0.8) +
       scale_fill_manual(name = "Celltype", values = immune_prelim_col) +
       xlab("") +
       ylab("Cell count") +
       theme_classic() +
       theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
       NoLegend()

ct_table %>% filter(Freq>1000) %>%
  ggplot(aes(x = reorder(Annotation_Cell, -Freq), y = Freq, fill = Annotation_Cell)) +
       geom_bar(stat="identity", position='stack', width = 0.8) +
       scale_fill_manual(name = "Celltype", values = immune_prelim_col) +
       xlab("") +
       ylab("Cell count") +
       theme_classic() +
       theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
       NoLegend()

```

### Combining similar cluster to less granular annotations

Non-immune

```{r, message = F, warning = F, fig.width = 22, fig.height = 22}

checked <- c("F38", "F27", "TumorEpi5", "TumorEpi4", "TumorEpi3", "TumorEpi6", "TumorEpi1", "F22", "Endo", "Peri", "F13", "F4", "F7", "F3", "F32", "F23", "F25", "F24", "F1", "F15", "F5", "F16", "F17", "F18", "Plasma", "F30", "F31", "F34", "F37", "F33", "F36", "F35", "F28", "F26", "F29", "F11", "F10", "F19", "F20", "F21", "F2", "F12", "F6", "F9", "F8", "F24_LowExp", "F23_LowExp", "TumorEpi2", "F14", "F2")

setdiff(checked, unique(nonimmune$Annotation_Cell))
setdiff(unique(nonimmune$Annotation_Cell), checked)

checked[duplicated(checked)]

g1 <- c("F16", "F17", "F18")
g2 <- c("F23_LowExp", "F24_LowExp")
g3 <- c("Plasma", "F30", "F31", "F32")
g4 <- c("F34", "F37", "F33", "F36", "F35")
g5 <- c("F28", "F26", "F29", "F11")
g6 <- c("F10", "F19", "F20", "F21")
g7 <- c("F2", "F12")
g8 <- c("F6", "F9")
g9 <- c("F8", "F4")
g10 <- c("F27", "F38")
g11 <- c("F22", "Fcombined5")
g12 <- c("F25", "Fcombined3")

nonimmune$Annotation_Cell2 <- ifelse(nonimmune$Annotation_Cell %in% g1, "Fcombined1",
                              ifelse(nonimmune$Annotation_Cell %in% g2, "Fcombined2",
                              ifelse(nonimmune$Annotation_Cell %in% g3, "Plasma2",
                              ifelse(nonimmune$Annotation_Cell %in% g4, "Fcombined3",
                              ifelse(nonimmune$Annotation_Cell %in% g5, "Fcombined4",
                              ifelse(nonimmune$Annotation_Cell %in% g6, "Fcombined5",
                              ifelse(nonimmune$Annotation_Cell %in% g7, "Fcombined6",
                              ifelse(nonimmune$Annotation_Cell %in% g8, "Fcombined7",
                              ifelse(nonimmune$Annotation_Cell %in% g9, "Fcombined8",
                              ifelse(nonimmune$Annotation_Cell %in% g10, "Fcombined10",
                              ifelse(nonimmune$Annotation_Cell %in% g11, "Fcombined11",
                              ifelse(nonimmune$Annotation_Cell %in% g12, "Fcombined12",
                                     as.character(nonimmune$Annotation_Cell)))))))))))))

g13 <- c("Fcombined1", "Fcombined2")
g14 <- c("Fcombined3", "F25")
g15 <- c("TumorEpi2", "TumorEpi4")
g16 <- c("Fcombined7", "F7")
g17 <- c("Fcombined14", "Fcombined13")

nonimmune$Annotation_Cell2 <- ifelse(nonimmune$Annotation_Cell2 %in% g13, "Fcombined13",
                              ifelse(nonimmune$Annotation_Cell2 %in% g14, "Fcombined14",
                              ifelse(nonimmune$Annotation_Cell2 %in% g15, "TumorEpiCombined1",
                              ifelse(nonimmune$Annotation_Cell2 %in% g16, "Fcombined15",
                              ifelse(nonimmune$Annotation_Cell2 %in% g17, "Fcombined16",
                              as.character(nonimmune$Annotation_Cell2))))))

```

Immune

```{r, message = F, warning = F, fig.width = 22, fig.height = 22}

checked <- c("LFambiguous18", "L38", "L17", "L41", "MFambiguous47", "M37", "M34", "M33", "M2", "M6", "MFambiguous46", "M39", "L1", "Treg8", "Treg2", "Treg3", "Treg5", "Treg6", "Treg4", "L2", "Treg7", "Treg15", "LFambiguous10", "Treg13", "Treg10", "Treg11", "Treg16", "Treg14", "Treg17", "Treg18", "Treg9", "Treg12", "L29", "L45", "L26", "L25", "L20", "L28", "L23", "L22", "L24", "L21", "L27", "L16", "L18", "L39", "L37", "MLFambiguous3", "MLFambiguous1", "MLFambiguous2", "MLFambiguous5", "MLFambiguous6", "MLFambiguous7", "MFambiguous19", "MFambiguous21", "MFambiguous22", "MFambiguous23", "MFambiguous20", "MFambiguous26", "MFambiguous14", "MFambiguous18", "M15", "MFambiguous76", "MFambiguous25", "MFambiguous24", "MFambiguous1", "M44", "M51", "M43", "M48", "M50", "M52", "M46", "M49", "M47", "M45", "M40", "M30", "M31", "M3", "M13", "M41", "M32", "M1", "M53", "MFambiguous80", "MFambiguous64", "MFambiguous63", "MFambiguous66", "MFambiguous73", "MFambiguous53", "MFambiguous77", "MFambiguous78", "MFambiguous52", "MFambiguous61", "MFambiguous49", "MFambiguous55", "MFambiguous48", "MFambiguous82", "MFambiguous54", "MFambiguous45", "MFambiguous75", "MFambiguous59", "MFambiguous57", "MFambiguous62", "MFambiguous60", "MFambiguous68", "MFambiguous65", "MFambiguous74", "MFambiguous70", "MFambiguous69", "MFambiguous67", "MFambiguous71", "MFambiguous72", "M36", "MFambiguous51", "M11", "M7", "MFambiguous50", "M38", "MFambiguous56", "MFambiguous58", "L34", "L33", "L31", "L32", "MFambiguous7", "MFambiguous2", "MFambiguous16", "MFambiguous4", "MFambiguous9", "MFambiguous10", "MFambiguous5", "MFambiguous81", "MFambiguous6", "MFambiguous11", "M29", "M12", "M5", "M28", "MFambiguous33", "M25", "M21", "M26", "MFambiguous12", "MFambiguous17", "MFambiguous15", "MFambiguous13", "MFambiguous8", "MFambiguous3", "MFambiguous30", "M14", "M8", "MFambiguous32", "MFambiguous27", "MFambiguous31", "M10", "M24", "M22", "MFambiguous34", "MFambiguous29", "M23", "M27", "M18", "MFambiguous35", "M20", "M17", "M16", "M4", "M9", "MFambiguous28", "M19", "L19", "L15", "L36", "L30", "L35", "L44", "MFambiguous36", "MFambiguous40", "MFambiguous42", "MFambiguous38", "MFambiguous39", "MFambiguous44", "MFambiguous37", "MFambiguous43", "M42", "MFambiguous41", "LFambiguous15", "LFambiguous11", "LFambiguous17", "LFambiguous13", "LFambiguous12", "L40", "LFambiguous14", "LFambiguous8", "LFambiguous16", "LFambiguous7", "LFambiguous9", "LFambiguous6", "LFambiguous5", "LFambiguous4", "LFambiguous1", "LFambiguous3", "LFambiguous2", "Treg1", "Treg19", "MLFambiguous8", "Treg20", "MLFambiguous4", "M35")

setdiff(unique(immune$Annotation_Cell), checked)
setdiff(checked, unique(immune$Annotation_Cell))

checked[duplicated(checked)]

g1 <- c("L1", "Treg8", "Treg2", "Treg3", "Treg5", "Treg6", "Treg4", "L2", "Treg7", "Treg15", "LFambiguous10", "Treg13", "Treg10", "Treg11", "Treg16", "Treg14", "Treg17", "Treg18", "Treg9", "Treg12")
g2 <- c("L29", "L45", "L26", "L25", "L20", "L28", "L23", "L22", "L24", "L21", "L27
L16", "L18")
g3 <- c("L39", "L37")
g4 <- c("MLFambiguous3", "MLFambiguous1", "MLFambiguous2", "MLFambiguous5", "MLFambiguous6")
g5 <- c("MFambiguous19", "MFambiguous21", "MFambiguous22", "MFambiguous23", "MFambiguous20", "MFambiguous26", "MFambiguous14", "MFambiguous18", "M15", "MFambiguous76", "MFambiguous25", "MFambiguous24", "MFambiguous1", "MLFambiguous7")
g6 <- c("M44", "M51", "M43", "M48", "M50", "M52", "M46", "M49", "M47", "M45")
g7 <- c("M40", "M30", "M31")
g8 <- c("M3", "M13", "M41", "M32", "M1", "M53")
g9 <- c("MFambiguous80", "MFambiguous64", "MFambiguous63", "MFambiguous66", "MFambiguous73", "MFambiguous53", "MFambiguous77", "MFambiguous78", "MFambiguous52", "MFambiguous61", "MFambiguous49", "MFambiguous55", "MFambiguous48", "MFambiguous82", "MFambiguous54", "MFambiguous45", "MFambiguous75", "MFambiguous59", "MFambiguous57", "MFambiguous62", "MFambiguous60", "MFambiguous68", "MFambiguous65", "MFambiguous74", "MFambiguous70", "MFambiguous69", "MFambiguous67", "MFambiguous71", "MFambiguous72")
g10 <- c("M36", "MFambiguous51", "M11", "M7", "MFambiguous50", "M38", "MFambiguous56", "MFambiguous58")
g11 <- c("L34", "L33", "L31", "L32")
g12 <- c("MFambiguous7", "MFambiguous2", "MFambiguous16", "MFambiguous4", "MFambiguous9", "MFambiguous10", "MFambiguous5", "MFambiguous81", "MFambiguous6", "MFambiguous11", "M29", "M12", "M5", "M28", "MFambiguous33", "M25", "M21", "M26", "MFambiguous12", "MFambiguous17", "MFambiguous15", "MFambiguous13", "MFambiguous8", "MFambiguous3")
g13 <- c("MFambiguous30", "M14", "M8", "MFambiguous32", "MFambiguous27", "MFambiguous31")
g14 <- c("M10", "M24", "M22", "MFambiguous34", "MFambiguous29", "M23", "M27", "M18", "MFambiguous35")
g15 <- c("M20", "M17", "M16")
g16 <- c("M4", "M9", "MFambiguous28", "M19")
g17 <- c("L19", "L15", "L36", "L30", "L35", "L44")
g18 <- c("MFambiguous36", "MFambiguous40", "MFambiguous42", "MFambiguous38", "MFambiguous39", "MFambiguous44", "MFambiguous37", "MFambiguous43", "M42", "MFambiguous41", "LFambiguous15", "LFambiguous11", "LFambiguous17", "LFambiguous13", "LFambiguous12", "L40", "LFambiguous14", "LFambiguous8", "LFambiguous16")
g19 <- c("LFambiguous7", "LFambiguous9", "LFambiguous6", "LFambiguous5", "LFambiguous4", "LFambiguous1", "LFambiguous3", "LFambiguous2", "Treg1", "Treg19", "MLFambiguous8", "Treg20", "MLFambiguous4")
g20 <- c("L14", "LFambiguous18")

immune$Annotation_Cell2 <- ifelse(immune$Annotation_Cell %in% g1, "Treg1",
                           ifelse(immune$Annotation_Cell %in% g2, "Lcombined1",
                           ifelse(immune$Annotation_Cell %in% g3, "Lcombined2",
                           ifelse(immune$Annotation_Cell %in% g4, "Unclassified1",
                           ifelse(immune$Annotation_Cell %in% g5, "Unclassified2",
                           ifelse(immune$Annotation_Cell %in% g6, "Mcombined1",
                           ifelse(immune$Annotation_Cell %in% g7, "Mcombined2",
                           ifelse(immune$Annotation_Cell %in% g8, "Mcombined3",
                           ifelse(immune$Annotation_Cell %in% g9, "Unclassified3",
                           ifelse(immune$Annotation_Cell %in% g10, "Mcombined4",
                           ifelse(immune$Annotation_Cell %in% g11, "Lcombined3",
                           ifelse(immune$Annotation_Cell %in% g12, "Mcombined4",
                           ifelse(immune$Annotation_Cell %in% g13, "Mcombined5",
                           ifelse(immune$Annotation_Cell %in% g14, "Mcombined6",
                           ifelse(immune$Annotation_Cell %in% g15, "Mcombined7",
                           ifelse(immune$Annotation_Cell %in% g16, "Mcombined8",
                           ifelse(immune$Annotation_Cell %in% g17, "Lcombined4",
                           ifelse(immune$Annotation_Cell %in% g18, "Unclassified4",
                           ifelse(immune$Annotation_Cell %in% g19, "Treg2",
                                  as.character(immune$Annotation_Cell))))))))))))))))))))

g21 <- c("Mcombined6", "Mcombined7")
g22 <- c("L27", "Lcombined1", "L38")
g23 <- c("M37", "L16")
g24 <- c("M2", "Mcombined5")
g25 <- c("LFambiguous18", "L41")
g26 <- c("Lcombined2", "Treg2")

immune$Annotation_Cell2 <- ifelse(immune$Annotation_Cell2 %in% g21, "Lcombined5",
                           ifelse(immune$Annotation_Cell2 %in% g22, "Mcombined9",
                           ifelse(immune$Annotation_Cell2 %in% g23, "Plasma1",
                           ifelse(immune$Annotation_Cell2 %in% g24, "Mcombined10",
                           ifelse(immune$Annotation_Cell2 %in% g25, "Lcombined6",
                           ifelse(immune$Annotation_Cell2 %in% g26, "Lcombined7",
                                  as.character(immune$Annotation_Cell2)))))))

saveRDS(nonimmune, "/scratch/hnatri/PIPAC/nonimmune_annotated_modified_combined.rds")
saveRDS(immune, "/scratch/hnatri/PIPAC/immune_annotated_modified_combined.rds")

```

### Moving around cells in wrong objects

```{r, message = F, warning = F, fig.width = 22, fig.height = 22, eval = T}

#nonimmune <- readRDS("/scratch/hnatri/PIPAC/nonimmune_annotated_modified_combined.rds")
#immune <- readRDS("/scratch/hnatri/PIPAC/immune_annotated_modified_combined.rds")

immune_from_nonimmune <- subset(nonimmune, subset = Annotation_Cell2 == "Plasma2")

immune@reductions$pca <- NULL
immune@reductions$umap <- NULL
#immune@reductions$sp <- NULL

immune_from_nonimmune@reductions$pca <- NULL
immune_from_nonimmune@reductions$umap <- NULL
#immune_from_nonimmune@reductions$sp <- NULL

immune@meta.data <- immune@meta.data[,c("cell_id", "combined_annotation", "Annotation_Cell2", "Annotation_Cell")]
immune_from_nonimmune@meta.data <- immune_from_nonimmune@meta.data[,c("cell_id", "combined_annotation", "Annotation_Cell2", "Annotation_Cell")]
nonimmune@meta.data <- nonimmune@meta.data[,c("cell_id", "combined_annotation", "Annotation_Cell2", "Annotation_Cell")]

immune <- merge(immune, immune_from_nonimmune)
nonimmune <- subset(nonimmune, subset = Annotation_Cell2 == "Plasma2", invert = T)

saveRDS(nonimmune, "/scratch/hnatri/PIPAC/nonimmune_annotated_modified_combined.rds")
saveRDS(immune, "/scratch/hnatri/PIPAC/immune_annotated_modified_combined.rds")

```

### Plotting after combining clusters

Non-immune

```{r, message = T, warning = F, fig.width = 22, fig.height = 22, eval = T}

nonimmune <- JoinLayers(nonimmune)
Idents(nonimmune) <- nonimmune$Annotation_Cell2
cluster_markers <- FindAllMarkers(nonimmune,
                                  return.thresh = 0.01,
                                  logfc.threshold = 0.5,
                                  min.pct = 0.20,
                                  verbose = F)

top_cluster_markers <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:15)

```

```{r, message = T, warning = F, fig.width = 26, fig.height = 30, eval = T}

nonimmune$Annotation_Cell2 <- factor(nonimmune$Annotation_Cell2,
                                    levels = sort(unique(nonimmune$Annotation_Cell2)))
nonimmune_prelim <- as.factor(sort(unique(nonimmune$Annotation_Cell2)))
nonimmune_prelim_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(nonimmune_prelim))
names(nonimmune_prelim_col) <- levels(nonimmune_prelim)

hp <- create_dotplot_heatmap(seurat_object = nonimmune,
                             plot_features = unique(c(top_cluster_markers$gene, main_plot_features)),
                             group_var = "Annotation_Cell2",
                             group_colors = nonimmune_prelim_col,
                             column_title = "",
                             row_km = 6,
                             col_km = 6,
                             row.order = NULL,
                             col.order = NULL)

#for (cl in names(column_order(hp))){
#  message(paste0("Cluster "), cl)
#  message(paste0(as.character(unique(nonimmune$Annotation_Cell2)[column_order(hp)[[as.character(cl)]]]), #sep = ", "))
#}

```

Immune

```{r, message = T, warning = F, fig.width = 8, fig.height = 22, eval = T}

immune <- JoinLayers(immune)
Idents(immune) <- immune$Annotation_Cell2
cluster_markers <- FindAllMarkers(immune,
                                  return.thresh = 0.01,
                                  logfc.threshold = 0.5,
                                  min.pct = 0.20,
                                  verbose = F)

top_cluster_markers <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:15)

```

```{r, message = T, warning = F, fig.width = 24, fig.height = 30, eval = T}

immune$Annotation_Cell2 <- factor(immune$Annotation_Cell2,
                                 levels = sort(unique(immune$Annotation_Cell2)))
immune_prelim <- as.factor(sort(unique(immune$Annotation_Cell2)))
immune_prelim_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(immune_prelim))
names(immune_prelim_col) <- levels(immune_prelim)

hp <- create_dotplot_heatmap(seurat_object = immune,
                             plot_features = unique(c(top_cluster_markers$gene, main_plot_features)),
                             group_var = "Annotation_Cell2",
                             group_colors = immune_prelim_col,
                             column_title = "",
                             row_km = 12,
                             col_km = 12,
                             row.order = NULL,
                             col.order = NULL)

#for (cl in names(column_order(hp))){
#  message(paste0("Cluster "), cl)
#  message(paste0(as.character(unique(immune$Annotation_Cell2)[column_order(hp)[[as.character(cl)]]]), sep #= ", "))
#}

```

#### Cells per cell type

```{r, message = F, warning = F, fig.width = 14, fig.height = 5, eval = T}

ct_table <- as.data.frame(table(nonimmune$Annotation_Cell2))
colnames(ct_table) <- c("Annotation_Cell2", "Freq")

min(ct_table$Freq)

ggplot(ct_table, aes(x = reorder(Annotation_Cell2, -Freq), y = Freq, fill = Annotation_Cell2)) +
       geom_bar(stat="identity", position='stack', width = 0.8) +
       scale_fill_manual(name = "Celltype", values = nonimmune_prelim_col) +
       xlab("") +
       ylab("Cell count") +
       theme_classic() +
       theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
       NoLegend()

```

```{r, message = F, warning = F, fig.width = 20, fig.height = 5, eval = T}

ct_table <- as.data.frame(table(immune$Annotation_Cell2))
colnames(ct_table) <- c("Annotation_Cell2", "Freq")

min(ct_table$Freq)

ggplot(ct_table, aes(x = reorder(Annotation_Cell2, -Freq), y = Freq, fill = Annotation_Cell2)) +
       geom_bar(stat="identity", position='stack', width = 0.8) +
       scale_fill_manual(name = "Celltype", values = immune_prelim_col) +
       xlab("") +
       ylab("Cell count") +
       theme_classic() +
       theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
       NoLegend()

```

### Reclustering pericytes

```{r, message = F, warning = F, fig.width = 4, fig.height = 4, eval = F}

Peri <- subset(nonimmune, subset = Annotation_Cell2 == "Peri")
Peri <- recluster_seurat(Peri, umap_pcs = 20)

DimPlot(Peri,
        group.by = "SCTsnn_res.0.8",
        reduction = "umap",
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

table(Peri$SCTsnn_res.0.8)

```

```{r, message = F, warning = F, fig.width = 22, fig.height = 6, eval = F}

DotPlot(Peri,
        group.by = "SCTsnn_res.0.8",
        features = main_plot_features,
        cols = c("gray98", "tomato3")) +
  RotatedAxis()
  
```

```{r, message = F, warning = F, fig.width = 9, fig.height = 6, eval = F}

FeaturePlot(Peri,
            features = c("CD68", "CD163", "COL1A2", "DCN", "FN1", "EPCAM"),
            cols = c("gray89", "tomato3"),
            reduction = "umap",
            order = T,
            raster = T,
            ncol = 3) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 22, fig.height = 36, eval = F}

FeaturePlot(Peri,
            features = main_plot_features,
            cols = c("gray98", "tomato3"),
            reduction = "umap",
            raster = T,
            ncol = 5) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

### Saving annotated data

```{r, message = F, warning = F, fig.width = 22, fig.height = 22}

#nonimmune <- readRDS("/scratch/hnatri/PIPAC/nonimmune_annotated_modified_combined.rds")
#immune <- readRDS("/scratch/hnatri/PIPAC/immune_annotated_modified_combined.rds")

# Adding final annotations to the complete clustered object
seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/Seurat/cell_merged_spatial_filtered_splitsamples_clustered_NN30_PC50_Seurat_denoIST.rds")

merged_data <- merge(nonimmune, immune)

seurat_data$Celltype <- mapvalues(seurat_data$cell_id,
                                  from = merged_data$cell_id,
                                  to = merged_data$Annotation_Cell2)

saveRDS(seurat_data, "/tgen_labs/banovich/PIPAC/Seurat/cell_merged_spatial_filtered_splitsamples_clustered_NN30_PC50_Seurat_denoIST_annotated.rds")

# Rscript -e "rmarkdown::render('finalize_annotations.Rmd')"

```
