---
title: "Subclustering of immune and non-immune subsets"
author: "heinin"
date: "2026-04-03"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Libraries and environment variables

```{r, message = F, warning = F}

library(workflowr)
library(Seurat)
library(googlesheets4)
library(tidyverse)
library(plyr)
library(SoupX)
library(scProportionTest)
library(clusterProfiler)
library(org.Hs.eg.db)
library(org.Mm.eg.db)
library(biomaRt)
library(gprofiler2)

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(1234)
options(future.globals.maxSize = 30000 * 1024^2)
reduction <- "umap"

source("/home/hnatri/SingleCellBestPractices/scripts/preprocessing_qc_module.R")
source("/home/hnatri/SingleCellBestPractices/scripts/integration_module.R")
source("/home/hnatri/Utilities/utilities.R")
source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

# Markers
main_plot_features <- c("PTPRC", "CD3D", "CD3E", "CD4", "CD8A", "JAK1", "JAK2", "STAT4", "STAT3", "TIGIT", "GZMB", "SELL", "CD19", "CD68", "CD44", "MARCO", "APOE", "C1QB", "C1QBP", "CD163", "SPP1", "VCAN", "MUC5AC", "NOTCH3", "RGS5", "MS4A1", "PGA5", "PLVAP", "CCL21", "LYVE", "SPARCL1", "FN1", "DCN", "LUM", "COL1A2", "PDPN", "ACTA2", "LMNA", "EGR3", "TP53", "JUN", "FOS", "BRAF", "KIT", "SOX9", "RNF43", "EPCAM", "PECAM", "CEACAM", "TGFB1", "JUNB", "TCF7", "GZMA", "GZMK", "CD47", "CD74", "LYZ", "TREM2", "CD247", "COL5A1", "SPARC", "CD36", "CTLA4", "RGS1", "CCL2", "SELENOP", "LYVE1", "CTSD", "MMP9", "CDK1", "FLT1", "RAMP2", "CCL14", "MCAM", "NRP1", "CEACAM6", "IL32", "TGFBI", "VEGFA", "LAMP1", "IGFBP7", "POSTN", "C1S", "C1R", "CSF1", "NOTCH1", "FOXP3", "IL2RB", "TUBA1B", "S100A9", "C3", "CXCR2", "SDC1", "CD38", "IGHG1", "PRDX4", "ICAM1", "GATA2", "CXCL3", "JCHAIN")

mycafs <- c("TGFB1", "ACTA2", "TAGLN") #"HOPX", "MYL9"
icafs <- c("IL6", "CXCL12", "CXCL14") #"CXCL13", 
apcafs <- c("CD74", "HLA-DPB1", "HLA-DPB2", "HLA-DQA1", "HLA-DRA", "HLA-DRB1")
ficafs <- c("SFRP2", "SFRP4", "COMP", "CXCL14", "PLA2G2A", "ASPN")

# F-apCAFs (C1QC, APOC1, RARRES1) and M-apCAFs (CD24, CA9, EGLN3)
more_apcafs <- c("C1QC", "APOC1", "RARRES1", "CD24", "CA9", "EGLN3")

# Matrix CAFs
mcafs <- c("FBLN1", "PDGFRA", "POSTN")

# Vascular CAFs
vcafs <- c("KCNJ8", "RGS5", "MCAM", "CD146", "NOTCH3", "NID2", "DES", "PDGFRB", "ACTA2")

main_plot_features <- c(main_plot_features, mycafs, icafs, apcafs, more_apcafs, ficafs, mcafs, vcafs)
main_plot_features <- unique(main_plot_features)

main_plot_features <- gorth(main_plot_features, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

caf_markers <- c("Ly6c", "Il6", "Ccl2", "Cxcl1", "Cxcl2", "Cxcl12",
                 "Pdgfra", "Agtr1", "Cfd", "Il1a", "Il1b", "Ccl8",
                 "Ccl27", "Acta2", "Myl9", "Tagln", "Tpm2", "Col1a1",
                 "Col1a2", "Col3a1")
caf_markers <- unique(caf_markers)

more_caf_markers <- c(mycafs, icafs, apcafs, more_apcafs, ficafs, mcafs, vcafs)
more_caf_markers <- unique(more_caf_markers)

more_caf_markers <- gorth(more_caf_markers, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

caf_markers <- unique(c(caf_markers, more_caf_markers))

mdsc <- c("Cd11b", "Gr-1", "Ly6G", "Ly6C", "Cd244", "Arg2", "Stat3", "Jaml",
          "Ccr2", "Arg1", "Csf1r", "Nos2", "S100a8", "S100a9", "Ros")

m2 <- c("Arg1", "Mrc1", "Cd206", "Chil3", "Ym1", "Retnla", "Fizz1",
        "Cd163", "Igf1", "Pparg", "Irf4", "Stat6", "Ccl17", "Ccl22", "Il10",
        "Tgfb1", "Egr2", "Fpr2")

m1 <- c("Nos2", "Il1b", "Il6", "Tnfa", "Il12a", "Il12b", "Il23", "Ccl5",
        "Cxcl2", "Cxcl10", "Cd80", "Cd38", "Cd68", "Cd64", "Fcgr1", "H2-Aa",
        "Cd11c", "Fpr2", "Gpr18")

mouse_immune <- c("Ptprc", "Cd3e", "Cd3g", "Cd247", "Cd4", "Cd8a", "Cd8b1", 
                  "Foxp3", "Cd25", "Il2ra", "Ctla4", "Cd19", "Ms4a1", "Cd79a",
                  "Ncr1", "NKp46", "Klrb1c", "Cd226",
                  "Adgre1", "Cd68", "Cd11b", "Itgam",
                  "Nos2", "Cd80", "Cd86",
                  "Cd163", "Mrc1", "Arg1", "Itgax", "Cd11c",
                  "Siglech", "Bst2", "Ly6g", "Ly6c", "S100a8",
                  "Ly6c", "Cd34", "Kit", "Sell", "Ccr7", "Cd44",
                  "Pdcd1")

emt_markers <- c("Cdh1", "Epcam", "Vim", "Fn1", "Cdh2", "Itga5",
                 "Snai1", "Snai2", "Twist1", "Zeb1", "Zeb2", "Mmp2",
                 "Mmp9", "Col1a1")

all_plot_features <- unique(c(main_plot_features, caf_markers, mdsc, m2, m1,
                              mouse_immune, emt_markers))

gs4_deauth()
metadata  <- gs4_get("https://docs.google.com/spreadsheets/d/1dRWnmQrzN8Q0EFgccnwSGIZ5GtnFyPn5SMYIS3pQCBg/edit?usp=sharing")
immune_annotations <- read_sheet(metadata, sheet = "Immune annotations")
nonimmune_annotations <- read_sheet(metadata, sheet = "Non-immune annotations")

```

## Import data

```{r, message = F, warning = F, eval = T, eval = T}

seurat_integrated <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_annotated.rds")

sort(as.character(unique(seurat_integrated$Updated_Annotation)))

seurat_integrated$Updated_Annotation <- factor(seurat_integrated$Updated_Annotation,
                                               levels = sort(as.character(unique(seurat_integrated$Updated_Annotation))))

immune_cts <- c("Lymph1_Tc_Tex", "Lymph2_Tc", "Lymph3_B", "Lymph4_NK", "Plasma1",
                "Myel1_MDSC_M2", "Myel2_pDC", "Myel3_MDSC_M2", "Myel4_mMDSC")

```

## Immune subset

### Integration

```{r, message = F, warning = F, eval = F}

immune <- subset(seurat_integrated, subset = Updated_Annotation %in% immune_cts)

immune[["RNA"]] <- split(immune[["RNA"]], f = immune$Sample )

DefaultAssay(immune) <- "RNA"
immune <-  SCTransform(immune, 
                       vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA"),
                       verbose = F)

s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

mmus_s <- gorth(s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m <- gorth(g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

immune <- CellCycleScoring(immune,
                           s.features = mmus_s,
                           g2m.features = mmus_g2m,
                           set.ident = F,
                           verbose = F) 

immune <-  SCTransform(immune, 
                       vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA",
                                           "S.Score", "G2M.Score", "Phase"),
                       verbose = F)

immune <- RunPCA(immune, assay = "SCT",
                 verbose = F)

immune <- IntegrateLayers(object = immune,
                          method = HarmonyIntegration,
                          orig.reduction = "pca",
                          new.reduction = 'harmony',
                          normalization.method = "SCT",
                          k.weight = 20, 
                          verbose = F)

#saveRDS(immune, "/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_immune_integrated.rds")

```

### Dimensionality reduction and clustering

```{r, message = F, warning = F, fig.width = 5, fig.height = 4, eval = F}

#immune <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_immune_integrated.rds")

immune <-  SCTransform(immune, 
                       vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA",
                                          "S.Score", "G2M.Score", "Phase"),
                       verbose = F)

immune <- RunPCA(immune,
                 assay = "SCT",
                 verbose = F)

umap_pcs <- get_pcs(immune, reduction_name = "pca")

immune <- RunUMAP(object = immune,
                  assay = "SCT",
                  reduction = "harmony",
                  reduction.name = "harmony_umap",
                  reduction.key = "harmony_umap_",
                  dims = 1:min(umap_pcs),
                  return.model = TRUE,
                  verbose = F)

immune <- FindNeighbors(object = immune,
                        assay = "SCT",
                        reduction = "harmony",
                        dims = 1:min(umap_pcs),
                        graph.name = c("harmony_nn",
                                       "harmony_snn"),
                        verbose = F)

immune <- FindClusters(object = immune,
                       resolution = c(0.1,0.2,0.3,0.5,0.8,1),
                       graph.name = "harmony_snn",
                       verbose = F)

#saveRDS(immune, "/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_immune_integrated.rds")

```

```{r, message = F, warning = F, fig.width = 5, fig.height = 5, eval = T}

immune <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_immune_integrated.rds")

DimPlot(immune,
        group.by = "harmony_snn_res.0.5",
        reduction = "harmony_umap",
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 5, eval = T}

DimPlot(immune,
        group.by = "Updated_Annotation",
        reduction = "harmony_umap",
        label = T,
        repel = T) +
  coord_fixed(ratio = 1) +
  theme_bw()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8, eval = T}

DimPlot(immune,
        group.by = "Updated_Annotation",
        split.by = "harmony_snn_res.0.5",
        reduction = "harmony_umap",
        ncol = 5) +
  coord_fixed(ratio = 1) +
  theme_bw()

DimPlot(immune,
        group.by = "harmony_snn_res.0.5",
        split.by = "Updated_Annotation",
        reduction = "harmony_umap",
        ncol = 4) +
  coord_fixed(ratio = 1) +
  theme_bw()

```

### Cluster/celltype stacked barplots

```{r, message = F, warning = F, fig.width = 10, fig.height = 5, eval = T}

table(immune$harmony_snn_res.0.5)

cts <- as.factor(sort(unique(as.character(immune$Updated_Annotation))))
ct_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(cts))
names(ct_col) <- levels(cts)

clusters <- as.factor(sort(unique(immune$harmony_snn_res.0.5)))
cluster_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(clusters))
names(cluster_col) <- levels(clusters)

create_barplot(immune,
               group_var = "harmony_snn_res.0.5",
               plot_var = "Updated_Annotation",
               plot_levels = sort(unique(as.character(immune$Updated_Annotation))),
               group_levels = sort(unique(immune$harmony_snn_res.0.5)),
               plot_colors = ct_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 5, eval = T}

create_barplot(immune,
               group_var = "Updated_Annotation",
               plot_var = "harmony_snn_res.0.5",
               plot_levels = sort(unique(as.character(immune$harmony_snn_res.0.5))),
               group_levels = sort(unique(immune$Updated_Annotation)),
               plot_colors = cluster_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "harmony_snn_res.0.5")

```

### Marker DotPlot

```{r, message = F, warning = F, fig.width = 8, fig.height = 18}

create_dotplot_heatmap(seurat_object = immune,
                       plot_features = sort(all_plot_features),
                       group_var = "harmony_snn_res.0.5",
                       group_colors = cluster_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

```{r, message = F, warning = F, fig.width = 14, fig.height = 60, eval = F}

FeaturePlot(immune,
            features = sort(all_plot_features),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 6) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

### Marker FeaturePlots

```{r, message = F, warning = F, fig.width = 12, fig.height = 12}

small_markers <- c("EPCAM", "VIM", "FN1",
                   "COL1A1", "ASPN", "PTPRC",
                   "CD3E", "CD8A", "CD4",
                   "C1QC", "SPP1", "TRP53",
                   "CDK1", "CD14", "JCHAIN")
small_markers <- gorth(small_markers, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

FeaturePlot(immune,
            features = sort(small_markers),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 4) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

### QC metrics by cluster

```{r, message = F, warning = F, fig.width = 12, fig.height = 9, eval = T}

VlnPlot(immune,
        features = c("nCount_RNA", "nFeature_RNA", "percent.mt_RNA"),
        group.by = "harmony_snn_res.0.5",
        pt.size = 0,
        ncol = 1) &
  NoLegend()

VlnPlot(immune,
        features = c("G2M.Score", "S.Score"),
        group.by = "Updated_Annotation",
        pt.size = 0,
        ncol = 1) &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 4}

FeaturePlot(immune,
            features = c("nCount_RNA", "nFeature_RNA", "percent.mt_RNA"),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 3) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 4}

FeaturePlot(immune,
            features = c("G2M.Score", "S.Score"),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 2) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

## Non-immune subset

### Integration

```{r, message = F, warning = F, eval = F}

nonimmune <- subset(seurat_integrated, subset = Updated_Annotation %in% immune_cts, invert = T)

nonimmune[["RNA"]] <- split(nonimmune[["RNA"]], f = nonimmune$Sample )

DefaultAssay(nonimmune) <- "RNA"
nonimmune <-  SCTransform(nonimmune, 
                       vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA"),
                       verbose = F)

s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

mmus_s <- gorth(s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m <- gorth(g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

nonimmune <- CellCycleScoring(nonimmune,
                              s.features = mmus_s,
                              g2m.features = mmus_g2m,
                              set.ident = F,
                              verbose = F) 

nonimmune <-  SCTransform(nonimmune, 
                          vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA",
                                              "S.Score", "G2M.Score", "Phase"),
                          verbose = F)

nonimmune <- RunPCA(nonimmune, assay = "SCT",
                    verbose = F)

nonimmune <- IntegrateLayers(object = nonimmune,
                             method = HarmonyIntegration,
                             orig.reduction = "pca",
                             new.reduction = 'harmony',
                             normalization.method = "SCT",
                             k.weight = 20, 
                             verbose = F)

saveRDS(nonimmune, "/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_nonimmune_integrated.rds")

```

### Dimensionality reduction and clustering

```{r, message = F, warning = F, fig.width = 5, fig.height = 4, eval = F}

#nonimmune <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_nonimmune_integrated.rds")

nonimmune <-  SCTransform(nonimmune, 
                       vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA",
                                          "S.Score", "G2M.Score", "Phase"),
                       verbose = F)

nonimmune <- RunPCA(nonimmune,
                 assay = "SCT",
                 verbose = F)

umap_pcs <- get_pcs(nonimmune, reduction_name = "pca")

nonimmune <- RunUMAP(object = nonimmune,
                  assay = "SCT",
                  reduction = "harmony",
                  reduction.name = "harmony_umap",
                  reduction.key = "harmony_umap_",
                  dims = 1:min(umap_pcs),
                  return.model = TRUE,
                  verbose = F)

nonimmune <- FindNeighbors(object = nonimmune,
                        assay = "SCT",
                        reduction = "harmony",
                        dims = 1:min(umap_pcs),
                        graph.name = c("harmony_nn",
                                       "harmony_snn"),
                        verbose = F)

nonimmune <- FindClusters(object = nonimmune,
                       resolution = c(0.1,0.2,0.3,0.5,0.8,1),
                       graph.name = "harmony_snn",
                       verbose = F)

#saveRDS(nonimmune, "/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_nonimmune_integrated.rds")

```

### Subclusteering subclusters

```{r, message = F, warning = F, fig.width = 5, fig.height = 5, eval = F}

nonimmune <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_nonimmune_integrated.rds")

Idents(nonimmune) <- nonimmune$harmony_snn_res.0.5
nonimmune <- FindSubCluster(nonimmune,
                            graph.name = "harmony_snn",
                            cluster = "9",
                            subcluster.name = "9_subcluster",
                            resolution = 0.5,
                            algorithm = 1)

Idents(nonimmune) <- nonimmune$`9_subcluster`
nonimmune <- FindSubCluster(nonimmune,
                            graph.name = "harmony_snn",
                            cluster = "11",
                            subcluster.name = "11_subcluster",
                            resolution = 0.5,
                            algorithm = 1)

Idents(nonimmune) <- nonimmune$`11_subcluster`
nonimmune <- FindSubCluster(nonimmune,
                            graph.name = "harmony_snn",
                            cluster = "12",
                            subcluster.name = "12_subcluster",
                            resolution = 0.5,
                            algorithm = 1)

Idents(nonimmune) <- nonimmune$`12_subcluster`
nonimmune <- FindSubCluster(nonimmune,
                            graph.name = "harmony_snn",
                            cluster = "3",
                            subcluster.name = "3_subcluster",
                            resolution = 0.5,
                            algorithm = 1)

Idents(nonimmune) <- nonimmune$`3_subcluster`
nonimmune <- FindSubCluster(nonimmune,
                            graph.name = "harmony_snn",
                            cluster = "5",
                            subcluster.name = "Subcluster",
                            resolution = 0.2,
                            algorithm = 1)

sort(unique(nonimmune$Subcluster))

DimPlot(nonimmune,
        group.by = "Subcluster",
        label = T) +
  theme_classic() +
  NoLegend()

saveRDS(nonimmune, "/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_nonimmune_subclustered.rds")

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 8, eval = T}

nonimmune <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_nonimmune_subclustered.rds")

DimPlot(nonimmune,
        group.by = "Subcluster",
        reduction = "harmony_umap",
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 5, eval = T}

DimPlot(nonimmune,
        group.by = "Updated_Annotation",
        reduction = "harmony_umap",
        label = T,
        repel = T) +
  coord_fixed(ratio = 1) +
  theme_bw()

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 12, eval = T}

DimPlot(nonimmune,
        group.by = "Updated_Annotation",
        split.by = "Subcluster",
        reduction = "harmony_umap",
        ncol = 6) +
  coord_fixed(ratio = 1) +
  theme_bw()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8, eval = T}

DimPlot(nonimmune,
        group.by = "Subcluster",
        split.by = "Updated_Annotation",
        reduction = "harmony_umap",
        ncol = 6) +
  coord_fixed(ratio = 1) +
  theme_bw()

```

### Cluster/celltype stacked barplots

```{r, message = F, warning = F, fig.width = 8, fig.height = 5, eval = T}

cts <- as.factor(sort(unique(as.character(nonimmune$Updated_Annotation))))
ct_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(cts))
names(ct_col) <- levels(cts)

clusters <- as.factor(sort(unique(nonimmune$Subcluster)))
cluster_col <- colorRampPalette(brewer.pal(12, "Paired"))(nb.cols <- length(clusters))
names(cluster_col) <- levels(clusters)

create_barplot(nonimmune,
               group_var = "Subcluster",
               plot_var = "Updated_Annotation",
               plot_levels = sort(unique(as.character(nonimmune$Updated_Annotation))),
               group_levels = sort(unique(nonimmune$Subcluster)),
               plot_colors = ct_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 5, eval = T}

create_barplot(nonimmune,
               group_var = "Updated_Annotation",
               plot_var = "Subcluster",
               plot_levels = sort(unique(as.character(nonimmune$Subcluster))),
               group_levels = sort(unique(nonimmune$Updated_Annotation)),
               plot_colors = cluster_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Cluster")

```

### Marker DotPlot

```{r, message = F, warning = F, fig.width = 8, fig.height = 18}

create_dotplot_heatmap(seurat_object = nonimmune,
                       plot_features = sort(all_plot_features),
                       group_var = "Subcluster",
                       group_colors = cluster_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

```{r, message = F, warning = F, fig.width = 14, fig.height = 60, eval = F}

FeaturePlot(nonimmune,
            features = sort(all_plot_features),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 6) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

### Marker FeaturePlots

```{r, message = F, warning = F, fig.width = 12, fig.height = 12}

small_markers <- c("EPCAM", "VIM", "FN1", "COL1A1", "ASPN", "PTPRC", "TGFB1",
                   "CD74", "EGFR", "VEGF")
small_markers <- gorth(small_markers, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

FeaturePlot(nonimmune,
            features = sort(small_markers),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 4) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

### QC metrics by cluster

```{r, message = F, warning = F, fig.width = 12, fig.height = 9, eval = T}

VlnPlot(nonimmune,
        features = c("nCount_RNA", "nFeature_RNA", "percent.mt_RNA"),
        group.by = "Subcluster",
        pt.size = 0,
        ncol = 1) &
  NoLegend()

VlnPlot(nonimmune,
        features = c("G2M.Score", "S.Score"),
        group.by = "Subcluster",
        pt.size = 0,
        ncol = 1) &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 10, fig.height = 4}

FeaturePlot(nonimmune,
            features = c("nCount_RNA", "nFeature_RNA", "percent.mt_RNA"),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 3) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 4}

FeaturePlot(nonimmune,
            features = c("G2M.Score", "S.Score"),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 2) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

## Removing doublets, reconstructing the UMAP, adding new annotations

```{r, message = F, warning = F, fig.width = 6, fig.height = 4, eval = F}

immune$Final_Celltype <- mapvalues(immune$harmony_snn_res.0.5,
                                   immune_annotations$harmony_snn_res.0.5,
                                   immune_annotations$`Final annotation`)

immune <- subset(immune, Final_Celltype == "Doublets", invert = T)

nonimmune$Final_Celltype <- mapvalues(nonimmune$Subcluster,
                                   nonimmune_annotations$Subcluster,
                                   nonimmune_annotations$`Final annotation`)

nonimmune <- subset(nonimmune, Final_Celltype %in% c("Doublets", "Remove"), invert = T)

immune_nonimmune <- list("immune" = immune,
                         "nonimmune" = nonimmune)

immune_nonimmune <- lapply(immune_nonimmune, function(xx){
  xx <-  SCTransform(xx, 
                     vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA",
                                        "S.Score", "G2M.Score", "Phase"),
                     verbose = F)

  xx <- RunPCA(xx,
                   assay = "SCT",
                   verbose = F)
  
  umap_pcs <- get_pcs(xx, reduction_name = "pca")
  
  xx <- RunUMAP(object = xx,
                    assay = "SCT",
                    reduction = "pca",
                    reduction.name = "umap",
                    reduction.key = "umap_",
                    dims = 1:min(umap_pcs),
                    return.model = TRUE,
                    verbose = F)
  
  xx <- FindNeighbors(object = xx,
                          assay = "SCT",
                          reduction = "pca",
                          dims = 1:min(umap_pcs),
                          graph.name = c("nn",
                                         "snn"),
                          verbose = F)
  
  xx <- FindClusters(object = xx,
                         resolution = c(0.1,0.2,0.3,0.5,0.8,1),
                         graph.name = "snn",
                         verbose = F)
  
  return(xx)
})

saveRDS(immune_nonimmune[["immune"]], "/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_immune_annotated_reUMAP.rds")
saveRDS(immune_nonimmune[["nonimmune"]], "/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_nonimmune_annotated_reUMAP.rds")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 4, eval = F}

# Merging
DefaultAssay(immune_nonimmune[["immune"]]) <- "RNA"
DefaultAssay(immune_nonimmune[["nonimmune"]]) <- "RNA"
immune_nonimmune[["immune"]]@assays$SCT <- NULL
immune_nonimmune[["nonimmune"]]@assays$SCT <- NULL
immune_nonimmune_seurat <- merge(immune_nonimmune[[1]], immune_nonimmune[[2]])

immune_nonimmune_seurat <-  SCTransform(immune_nonimmune_seurat, 
                                        vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA",
                                                           "S.Score", "G2M.Score", "Phase"),
                                        verbose = F)

immune_nonimmune_seurat <- RunPCA(immune_nonimmune_seurat,
                 assay = "SCT",
                 verbose = F)

umap_pcs <- get_pcs(immune_nonimmune_seurat, reduction_name = "pca")

immune_nonimmune_seurat <- RunUMAP(object = immune_nonimmune_seurat,
                  assay = "SCT",
                  reduction = "pca",
                  reduction.name = "umap",
                  reduction.key = "map_",
                  dims = 1:min(umap_pcs),
                  return.model = TRUE,
                  verbose = F)

immune_nonimmune_seurat <- FindNeighbors(object = immune_nonimmune_seurat,
                        assay = "SCT",
                        reduction = "pca",
                        dims = 1:min(umap_pcs),
                        graph.name = c("nn",
                                       "snn"),
                        verbose = F)

immune_nonimmune_seurat <- FindClusters(object = immune_nonimmune_seurat,
                       resolution = c(0.1,0.2,0.3,0.5,0.8,1),
                       graph.name = "snn",
                       verbose = F)
                         
saveRDS(immune_nonimmune_seurat, "/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_fulldata_annotated.rds")

```

```{r, message = F, warning = F, fig.width = 5, fig.height = 5, eval = T}

immune_nonimmune_seurat <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_fulldata_annotated.rds")
immune <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_immune_annotated_reUMAP.rds")
nonimmune <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_nonimmune_annotated_reUMAP.rds")

cts <- as.factor(sort(unique(as.character(immune_nonimmune_seurat$Final_Celltype))))
ct_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(cts))
names(ct_col) <- levels(cts)

DimPlot(immune_nonimmune_seurat,
        group.by = "Final_Celltype",
        reduction = "umap",
        label = T,
        repel = T,
        cols = ct_col) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

DimPlot(immune,
        group.by = "Final_Celltype",
        reduction = "umap",
        label = T,
        repel = T,
        cols = ct_col) +
  coord_fixed(ratio = 1) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  theme_bw() +
  NoLegend()

DimPlot(immune,
        group.by = "snn_res.1",
        reduction = "umap",
        label = T,
        repel = T) +
  coord_fixed(ratio = 1) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  theme_bw() +
  NoLegend()

DimPlot(nonimmune,
        group.by = "Final_Celltype",
        reduction = "umap",
        label = T,
        repel = T,
        cols = ct_col) +
  coord_fixed(ratio = 1) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  theme_bw() +
  NoLegend()


# Rscript -e "rmarkdown::render('PDX_CAF_subclustering.Rmd')"

```


