---
title: "Integration, annotation, and comparative analysis of PDX CAF scRNAseq data"
author: "heinin"
date: "2026-01-31"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Libraries and environment variables

```{r, message = F, warning = F}

library(workflowr)
library(Seurat)
library(googlesheets4)
library(tidyverse)
library(plyr)
library(SoupX)
library(scProportionTest)
library(clusterProfiler)
library(org.Hs.eg.db)
library(org.Mm.eg.db)
library(biomaRt)
library(gprofiler2)

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(1234)
options(future.globals.maxSize = 30000 * 1024^2)
reduction <- "umap"

source("/home/hnatri/SingleCellBestPractices/scripts/preprocessing_qc_module.R")
source("/home/hnatri/SingleCellBestPractices/scripts/integration_module.R")
source("/home/hnatri/Utilities/utilities.R")
source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

# Markers
main_plot_features <- c("PTPRC", "CD3D", "CD3E", "CD4", "CD8A", "JAK1", "JAK2", "STAT4", "STAT3", "TIGIT", "GZMB", "SELL", "CD19", "CD68", "CD44", "MARCO", "APOE", "C1QB", "C1QBP", "CD163", "SPP1", "VCAN", "MUC5AC", "NOTCH3", "RGS5", "MS4A1", "PGA5", "PLVAP", "CCL21", "LYVE", "SPARCL1", "FN1", "DCN", "LUM", "COL1A2", "PDPN", "ACTA2", "LMNA", "EGR3", "TP53", "JUN", "FOS", "BRAF", "KIT", "SOX9", "RNF43", "EPCAM", "PECAM", "CEACAM", "TGFB1", "JUNB", "TCF7", "GZMA", "GZMK", "CD47", "CD74", "LYZ", "TREM2", "CD247", "COL5A1", "SPARC", "CD36", "CTLA4", "RGS1", "CCL2", "SELENOP", "LYVE1", "CTSD", "MMP9", "CDK1", "FLT1", "RAMP2", "CCL14", "MCAM", "NRP1", "CEACAM6", "IL32", "TGFBI", "VEGFA", "LAMP1", "IGFBP7", "POSTN", "C1S", "C1R", "CSF1", "NOTCH1", "FOXP3", "IL2RB", "TUBA1B", "S100A9", "C3", "CXCR2", "SDC1", "CD38", "IGHG1", "PRDX4", "ICAM1", "GATA2", "CXCL3")

mycafs <- c("TGFB1", "ACTA2", "TAGLN") #"HOPX", "MYL9"
icafs <- c("IL6", "CXCL12", "CXCL14") #"CXCL13", 
abcafs <- c("CD74", "HLA-DPB1", "HLA-DPB2", "HLA-DQA1", "HLA-DRA", "HLA-DRB1")
ficafs <- c("SFRP2", "SFRP4", "COMP", "CXCL14", "PLA2G2A", "ASPN")

main_plot_features <- c(main_plot_features, mycafs, icafs, abcafs, ficafs)
main_plot_features <- unique(main_plot_features)

main_plot_features <- gorth(main_plot_features, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

```

## Import data

```{r, message = F, warning = F, eval = T, eval = F}

seurat_data <- readRDS("/scratch/hnatri/PIPAC/PDX_CAF_Seurat_filtered.rds")

table(seurat_data$Sample)

```

## Integration

```{r, message = F, warning = F, eval = F}

DefaultAssay(seurat_data) <- "RNA"
seurat_data <-  SCTransform(seurat_data, 
                            vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA"),
                            verbose = F)

s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

mmus_s <- gorth(s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m <- gorth(g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

seurat_data <- CellCycleScoring(seurat_data,
                                s.features = mmus_s,
                                g2m.features = mmus_g2m,
                                set.ident = F,
                                verbose = F) 

seurat_data <-  SCTransform(seurat_data, 
                            vars.to.regress = c("S.Score", "G2M.Score", "Phase"),
                            verbose = F)

seurat_data <- RunPCA(seurat_data, assay = "SCT",
                      verbose = F)

seurat_integrated <- IntegrateLayers(object = seurat_data,
                                     method = HarmonyIntegration,
                                     orig.reduction = "pca",
                                     new.reduction = 'harmony',
                                     normalization.method = "SCT",
                                     k.weight = 20,
                                     verbose = FALSE)

#saveRDS(seurat_integrated, "/scratch/hnatri/PIPAC/PDX_CAF_Seurat_integrated.rds")

```

## Dimensionality reduction and clustering

```{r, message = F, warning = F, fig.width = 5, fig.height = 4, eval = F}

seurat_integrated <- readRDS("/scratch/hnatri/PIPAC/PDX_CAF_Seurat_integrated.rds")

seurat_integrated <-  SCTransform(seurat_integrated, 
                                  vars.to.regress = c("S.Score", "G2M.Score", "Phase"),
                                  verbose = F)

seurat_integrated <- RunPCA(seurat_integrated,
                            assay = "SCT",
                            verbose = F)

umap_pcs <- get_pcs(seurat_integrated, reduction_name = "pca")

seurat_integrated <- RunUMAP(object = seurat_integrated,
                             assay = "SCT",
                             reduction = "harmony",
                             reduction.name = "harmony_umap",
                             reduction.key = "harmony_umap_",
                             dims = 1:min(umap_pcs),
                             return.model = TRUE,
                             verbose = F)

seurat_integrated <- FindNeighbors(object = seurat_integrated,
                                 assay = "SCT",
                                 reduction = "harmony",
                                 dims = 1:min(umap_pcs),
                                 graph.name = c("harmony_nn",
                                                "harmony_snn"),
                                 verbose = F)

seurat_integrated <- FindClusters(object = seurat_integrated,
                                resolution = c(0.1,0.2,0.3,0.5,0.8,1),
                                graph.name = "harmony_snn",
                                 verbose = F)

DimPlot(seurat_integrated,
        group.by = "harmony_snn_res.0.8",
        reduction = "harmony_umap",
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

table(seurat_integrated$harmony_snn_res.0.8)

#saveRDS(seurat_integrated, "/scratch/hnatri/PIPAC/PDX_CAF_Seurat_integrated.rds")

```

## Annotation

SoupX RNA

```{r, message = F, warning = F, fig.width = 22, fig.height = 56, eval = F}

DefaultAssay(seurat_integrated) <- "SoupX_RNA"
seurat_integrated <- JoinLayers(seurat_integrated)
seurat_integrated <- NormalizeData(seurat_integrated)

FeaturePlot(seurat_integrated,
            features = main_plot_features,
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 5) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

RNA

```{r, message = F, warning = F, fig.width = 4, fig.height = 4}

seurat_integrated <- readRDS("/scratch/hnatri/PIPAC/PDX_CAF_Seurat_integrated.rds")

DefaultAssay(seurat_integrated) <- "RNA"
seurat_integrated <- JoinLayers(seurat_integrated)
seurat_integrated <- NormalizeData(seurat_integrated)

DimPlot(seurat_integrated,
        group.by = "harmony_snn_res.0.8",
        reduction = "harmony_umap",
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

table(seurat_integrated$harmony_snn_res.0.8)

```

```{r, message = F, warning = F, fig.width = 22, fig.height = 56}

FeaturePlot(seurat_integrated,
            features = main_plot_features,
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 5) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

mini_features <- gorth(c("CD3E", "CD8", "SPP1", "CD68",
                         "FN", "COL1A1", "EPCAM"), source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

FeaturePlot(seurat_integrated,
            features = mini_features,
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 3) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 16}

clusters <- as.factor(sort(unique(seurat_integrated$harmony_snn_res.0.8)))
cluster_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(clusters))
names(cluster_col) <- levels(clusters)

create_dotplot_heatmap(seurat_object = seurat_integrated,
                       plot_features = main_plot_features,
                       group_var = "harmony_snn_res.0.8",
                       group_colors = cluster_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

### Top cluster markers

```{r, message = F, warning = F, fig.width = 5, fig.height = 4}

Idents(seurat_integrated) <- seurat_integrated$harmony_snn_res.0.8
cluster_markers <- FindAllMarkers(seurat_integrated,
                                  return.thresh = 0.01,
                                  logfc.threshold = 0.5,
                                  min.pct = 0.20,
                                  verbose = F)

table(cluster_markers$cluster)

top_cluster_markers <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:10)

top_cluster_markers_save <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:30)

#write.table(top_cluster_markers_save, "/home/hnatri/PIPAC_spatial/PDX_CAF_cluster_markers.tsv",
#            quote = F, sep = "\t")

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 21}

create_dotplot_heatmap(seurat_object = seurat_integrated,
                       plot_features = unique(top_cluster_markers$gene),
                       group_var = "harmony_snn_res.0.8",
                       group_colors = cluster_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

```{r, message = F, warning = F, fig.width = 22, fig.height = 66}

FeaturePlot(seurat_integrated,
            features = unique(top_cluster_markers$gene),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 8) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8, eval = T}

# Annotation df
# Non-immune c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 14, 15, 16, 18, 19)
annotations <- data.frame("cluster" = c(0, seq(1, 24)),
                          "annotation" = c("Tumor1", # 0
                                           "Proliferating1",
                                           "Tumor2",
                                           "CAF1", # 3
                                           "Tumor3", # 4
                                           "CAF2",
                                           "Proliferating2",
                                           "Tumor4",
                                           "CAF3", #8
                                           "CAF4",
                                           "Myeloid1",
                                           "Myeloid2",
                                           "Lymphoid1", # 12
                                           "Lymphoid2",
                                           "CAF5",
                                           "CAF6", # 15
                                           "CAF7",
                                           "Immune_Unclassified1",
                                           "Tumor5",
                                           "Proliferating3",
                                           "Immune_Unclassified2",
                                           "Immune_Unclassified3",
                                           "Immune_Unclassified4",
                                           "Immune_Unclassified5",
                                           "Immune_Unclassified6"))
seurat_integrated$Annotation <- mapvalues(seurat_integrated$harmony_snn_res.0.8,
                                from = annotations$cluster,
                                to = annotations$annotation)

seurat_integrated$Annotation <- factor(seurat_integrated$Annotation,
                                       levels = sort(unique(as.character(seurat_integrated$Annotation))))

# Updating annotations
seurat_integrated$Updated_Annotation <- seurat_integrated$Annotation
seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified1", "Lymphoid3", seurat_integrated$Updated_Annotation)
seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified2", "Lymphoid4", seurat_integrated$Updated_Annotation)
seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified6", "Lymphoid5", seurat_integrated$Updated_Annotation)
seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified3", "Myeloid3", seurat_integrated$Updated_Annotation)
seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified4", "Myeloid4", seurat_integrated$Updated_Annotation)
seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified5", "Unclassified", seurat_integrated$Updated_Annotation)

seurat_integrated$Updated_Annotation <- factor(seurat_integrated$Updated_Annotation,
                                       levels = sort(unique(as.character(seurat_integrated$Updated_Annotation))))

table(seurat_integrated$Updated_Annotation)

```


```{r, message = F, warning = F, fig.width = 6, fig.height = 6}

cts <- as.factor(sort(unique(as.character(seurat_integrated$Updated_Annotation))))
ct_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(cts))
names(ct_col) <- levels(cts)

DimPlot(seurat_integrated,
        group.by = "Updated_Annotation",
        reduction = "harmony_umap",
        cols = ct_col,
        label = T,
        label.box = T,
        repel = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

```


```{r, message = F, warning = F, fig.width = 8, fig.height = 23}

create_dotplot_heatmap(seurat_object = seurat_integrated,
                       plot_features = unique(top_cluster_markers$gene),
                       group_var = "Updated_Annotation",
                       group_colors = ct_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

## Comparative analysis

### Cell type proportions

```{r, message = F, warning = F, fig.width = 7, fig.height = 5, eval = T}

table(seurat_integrated$Sample)
table(seurat_integrated$Sample, seurat_integrated$Updated_Annotation)

create_barplot(seurat_integrated,
               group_var = "Sample",
               plot_var = "Updated_Annotation",
               plot_levels = sort(unique(as.character(seurat_integrated$Updated_Annotation))),
               group_levels = sort(unique(seurat_integrated$Sample)),
               plot_colors = ct_col,
               var_names =  c("Frequency (%)", ""),
               legend_title = "Celltype")

```

```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

prop_test <- sc_utils(seurat_integrated)
prop_test <- permutation_test(
	prop_test, cluster_identity = "Updated_Annotation",
	sample_1 = "CT26", sample_2 = "CT26_CAF",
	sample_identity = "Sample")

#permutation_plot(prop_test) +
#  ggtitle("CT26 vs. CT26+CAF")
#
#filename <- "/home/hnatri/PIPAC_spatial/CT26_CAF_celltypeprops.pdf"
#
#pdf(file=filename,
#    width = 5.5,
#    height = 6)
#
#permutation_plot(prop_test) +
#  ggtitle("CT26 vs. CT26+CAF")
#
#dev.off()

```

## DEGs, CAFs only

```{r, message = F, warning = F, fig.width = 8, fig.height = 4, eval = T}

#tumor_DEGs <- get_DEGs(seuratdata = seurat_integrated,
#                       celltypes = unique(seurat_integrated$harmony_snn_res.0.8),
#                       groupvar = "PrePost",
#                       group1 = "Treated",
#                       group2 = "Baseline")
#tumor_only <- subset(seurat_integrated, subset = harmony_snn_res.0.8 %in% c(0, 2, 4, 7, 18, 21,
#                                                                            16, 5, 9, 8, 14, 15, 3,
#                                                                            19, 1, 6))

#tumor_only <- c("CAF1", "CAF2", "CAF3", "CAF4", "CAF5", "CAF6", "CAF7",
#                "Proliferating1", "Proliferating2", "Proliferating3",
#                "Tumor1", "Tumor2", "Tumor3", "Tumor4", "Tumor5")

caf_only <- c("CAF1", "CAF2", "CAF3", "CAF4", "CAF5", "CAF6", "CAF7")

caf_only <- subset(seurat_integrated, subset = Updated_Annotation %in% caf_only)

Idents(caf_only) <- caf_only$Sample
caf_DEGs <- FindMarkers(caf_only,
                          ident.1 = "CT26",
                          ident.2 = "CT26_CAF")

# Distribution of p-values and log2FC
hist(caf_DEGs$p_val)
hist(caf_DEGs$p_val_adj)
hist(caf_DEGs$avg_log2FC)

ensembl <- useMart("ensembl", dataset="mmusculus_gene_ensembl")
geneinfo <- getBM(attributes=c("mgi_symbol", "ensembl_gene_id"), 
      filters = 'mgi_symbol', 
      values = rownames(caf_DEGs), 
      mart = ensembl)

caf_DEGs$gene_id <- mapvalues(rownames(caf_DEGs),
                                   from = geneinfo$mgi_symbol,
                                   to = geneinfo$ensembl_gene_id)

caf_DEGs_sig <- caf_DEGs %>%
  filter(p_val_adj < 0.01,
         abs(avg_log2FC) < 1,
         (pct.1 > 0.2 | pct.2 > 0.2))

rownames(caf_DEGs_sig)

caf_DEGs_sig %>%
  filter(avg_log2FC>0) %>% rownames()

caf_DEGs_sig %>%
  filter(avg_log2FC<0) %>% rownames()

```

## Some features upregulated in CT26+CAF

```{r, message = F, warning = F, fig.width = 8, fig.height = 10, eval = T}

caf_DEGs_sig %>%
  filter(avg_log2FC<0) %>% 
  arrange(desc(avg_log2FC)) %>%
  head(20)

upreg <- caf_DEGs_sig %>%
  filter(avg_log2FC<0) %>% 
  arrange(desc(avg_log2FC)) %>%
  rownames()

VlnPlot(caf_only,
        group.by = "Sample",
        features = head(upreg, n = 12),
        pt.size = 0,
        ncol = 4) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

## Some features downregulated in CT26+CAF

```{r, message = F, warning = F, fig.width = 8, fig.height = 10, eval = T}

caf_DEGs_sig %>%
  filter(avg_log2FC>0) %>% 
  arrange(desc(avg_log2FC)) %>%
  head(20)

downreg <- caf_DEGs_sig %>%
  filter(avg_log2FC>0) %>% 
  arrange(desc(avg_log2FC)) %>%
  rownames()

VlnPlot(caf_only,
        group.by = "Sample",
        features = head(downreg, n = 12),
        pt.size = 0,
        ncol = 4) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

## Pathway enrichment, upregulated in CT26+CAF

```{r, message = F, warning = F, fig.width = 5, fig.height = 8, eval = T}

caf_DEGs_up <- caf_DEGs_sig %>%
  filter(avg_log2FC<0)

gse_res <- enrichGO(rownames(caf_DEGs_up),
                    ont ="BP", 
                    keyType = "SYMBOL",
                    OrgDb = org.Mm.eg.db,
                    minGSSize = 3, 
                    maxGSSize = 1000, 
                    pvalueCutoff = 0.01, 
                    qvalueCutoff = 1,
                    universe = rownames(caf_DEGs),
                    pAdjustMethod = "none")

ddp <- dotplot(gse_res, showCategory=10, font.size=10, color="pvalue") +
  theme(legend.text=element_text(size=5),
        legend.position = "top",
        legend.box = "vertical",
        text = element_text(size=6))

dim(gse_res)

ddp

#filename <- "/home/hnatri/PIPAC_spatial/CT26_CAF_pathways.pdf"
#
#pdf(file=filename,
#    width = 4.5,
#    height = 6)
#
#ddp
#
#dev.off()

```

## Pathway enrichment, downregulated in CT26+CAF

```{r, message = F, warning = F, fig.width = 5, fig.height = 8, eval = T}

caf_DEGs_down <- caf_DEGs_sig %>%
  filter(avg_log2FC>0)

gse_res <- enrichGO(rownames(caf_DEGs_down),
                    ont ="BP", 
                    keyType = "SYMBOL",
                    OrgDb = org.Mm.eg.db,
                    minGSSize = 3, 
                    maxGSSize = 1000, 
                    pvalueCutoff = 0.01, 
                    qvalueCutoff = 1,
                    universe = rownames(caf_DEGs),
                    pAdjustMethod = "none")

ddp <- dotplot(gse_res, showCategory=10, font.size=10, color="pvalue") +
  theme(legend.text=element_text(size=5),
        legend.position = "top",
        legend.box = "vertical",
        text = element_text(size=6))

dim(gse_res)

ddp

#filename <- "/home/hnatri/PIPAC_spatial/CT26_CAF_pathways.pdf"
#
#pdf(file=filename,
#    width = 4.5,
#    height = 6)
#
#ddp
#
#dev.off()

```


```{r, message = F, warning = F, fig.width = 5, fig.height = 8, eval = F}


filename <- "/home/hnatri/PIPAC_spatial/CT26_CAF_UMAP.pdf"

pdf(file=filename,
    width = 8,
    height = 5)

DimPlot(seurat_integrated,
        group.by = "Annotation",
        reduction = "harmony_umap",
        cols = ct_col,
        label = T,
        label.box = T,
        repel = T) +
  coord_fixed(ratio = 1) +
  theme_bw()

dev.off()

filename <- "/home/hnatri/PIPAC_spatial/CT26_CAF_UMAP_sample.pdf"

pdf(file=filename,
    width = 9,
    height = 4)

DimPlot(seurat_integrated,
        group.by = "Annotation",
        split.by = "Sample",
        reduction = "harmony_umap",
        cols = ct_col,
        label = F) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

dev.off()

```

## Marker features, CAFs only

CMS4

```{r, message = F, warning = F, fig.width = 8, fig.height = 10, eval = T}

# CSM4
csm4_genes <- c("TGFB1", "TGFB2", "TGFB3", "VIM", "CDH2", "FN1", "ACTA2", "MMP2", "MMP3", "MMP9", "SOX10", "FOXC2", "S100A11", "COL1A1", "DSC", "EPCAM", "KRT7", "KRT14", "TJP1")
  
genenames <- gorth(csm4_genes, source_organism = "hsapiens", target_organism = "mmusculus")

VlnPlot(caf_only,
        group.by = "Sample",
        features = unique(genenames$ortholog_name),
        pt.size = 0,
        ncol = 4) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

## Myeloid markers

Key Mouse MDSC Markers and Subsets \

- Pan-MDSC: CD11b+, Gr-1+ \
- PMN-MDSC (Granulocytic): CD11b+ Ly6G+ Ly6Clow, CD244+, Arg2, Stat3, Jaml. \
- M-MDSC (Monocytic): CD11b+ Ly6G− Ly6Chigh, CCR2+, Arg1, Csf1r. \
- Functional Genes: Arg1 (Arginase-1), Nos2 (iNOS), S100a8, S100a9, ROS.

```{r, message = F, warning = F, fig.width = 12, fig.height = 10, eval = T}

m2 <- c("CD11b", "Gr-1", "Ly6G", "Ly6C", "CD244", "Arg2", "Stat3", "Jaml",
          "CCR2", "Arg1", "Csf1r", "Nos2", "S100a8", "S100a9", "ROS")

VlnPlot(seurat_integrated,
        group.by = "Updated_Annotation",
        features = m2,
        pt.size = 0,
        ncol = 2) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

Key Mouse M2 Marker Genes: \

- Arginase-1 (Arg1): Crucial enzyme for polyamine synthesis. \
- Mannose Receptor (Mrc1/CD206): Surface receptor, often used for identifying M2 macrophages. \
- Chitinase-like 3 (Chil3/Ym1): Secreted lectin involved in inflammation. \
- Resistin-like molecule alpha (Retnla/Fizz1): Secreted protein, common marker. \
- CD163 Scavenger Receptor (Cd163): Hemoglobin scavenger receptor. \
- Insulin-like Growth Factor 1 (Igf1): Growth factor involved in tissue repair. \
- Transcription Factors/Signaling: Pparg, Irf4, and Stat6. \
- Chemokines/Cytokines: Ccl17, Ccl22, Il10, and Tgfb1. \
- Novel Markers: Egr2 and Fpr2 are also identified as markers for M2 polarization.


```{r, message = F, warning = F, fig.width = 12, fig.height = 10, eval = T}

mdsc <- c("CD11b", "Gr-1", "Ly6G", "Ly6C", "CD244", "Arg2", "Stat3", "Jaml",
          "CCR2", "Arg1", "Csf1r", "Nos2", "S100a8", "S100a9", "ROS")

VlnPlot(seurat_integrated,
        group.by = "Updated_Annotation",
        features = mdsc,
        pt.size = 0,
        ncol = 2) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

Core M1 Marker Genes and Molecules (Murine) \

- Enzyme: Nos2 (Inducible Nitric Oxide Synthase) - Often considered the quintessential M1 marker. \
- Cytokines: Il1b (IL-1), Il6, Tnfa, Il12a, Il12b, Il23. \
- Chemokines: Ccl5 (RANTES), Cxcl2, Cxcl10. \
- Surface Markers: Cd80, Cd86, Cd38, Cd68, Cd64, Fcgr1 (CD64), H2-Aa (MHC II), Cd11c. \
- Novel/Specific Markers: Cd38, Fpr2, Gpr18.


```{r, message = F, warning = F, fig.width = 12, fig.height = 20, eval = T}

m1 <- c("Nos2", "Il1b", "Il6", "Tnfa", "Il12a", "Il12b", "Il23", "Ccl5",
        "Cxcl2", "Cxcl10", "Cd80", "Cd38", "Cd68", "Cd64", "Fcgr1", "H2-Aa",
        "Cd11c", "Fpr2", "Gpr18")

VlnPlot(seurat_integrated,
        group.by = "Updated_Annotation",
        features = m1,
        pt.size = 0,
        ncol = 2) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

Mouse immune markers

Mouse Immune Cell Marker Summary (Mus musculus) 

- Total Leukocytes: Ptprc (CD45)
- T Cells (Pan): Cd3e, Cd3g, Cd247 (CD3)
- Helper T Cells: Cd4
- Cytotoxic T Cells: Cd8a, Cd8b1
- Regulatory T Cells (Tregs): Foxp3, Cd25 (Il2ra), Ctla4
- B Cells: Cd19, Ms4a1 (CD20), Cd79a
- NK Cells: Ncr1 (NKp46), Klrb1c (NK1.1), Cd226
- Macrophages (Pan): Adgre1 (F4/80), Cd68, Cd11b (Itgam)
- M1 Macrophages (Pro-inflammatory): Nos2 (iNOS), Cd80, Cd86
- M2 Macrophages (Anti-inflammatory): Cd163, Mrc1 (CD206), Arg1
- Conventional Dendritic Cells (cDCs): Itgax (CD11c), H2-Ab1 (MHCII)
- Plasmacytoid DCs (pDCs): Siglech, Bst2 (CD317)
- Neutrophils: Ly6g, Ly6c, S100a8
- Monocytes: Ly6c, Cd11b (Itgam)
- Hematopoietic Stem/Progenitor Cells: Cd34, Kit (c-Kit) 

Key Activation/Memory Markers

- Naïve T Cells: Sell (CD62L), Ccr7
- Memory T Cells: Cd44, Il7r (CD127)
- Exhaustion Markers: Pdcd1 (PD-1), Cd274 (PD-L1), Havcr2 (TIM-3) 

```{r, message = F, warning = F, fig.width = 12, fig.height = 40, eval = T}

mouse_immune <- c("Ptprc", "Cd3e", "Cd3g", "Cd247", "Cd4", "Cd8a", "Cd8b1", 
                  "Foxp3", "Cd25", "Il2ra", "Ctla4", "Cd19", "Ms4a1", "Cd79a",
                  "Ncr1", "NKp46", "Klrb1c", "Cd226",
                  "Adgre1", "Cd68", "Cd11b", "Itgam",
                  "Nos2", "Cd80", "Cd86",
                  "Cd163", "Mrc1", "Arg1", "Itgax", "Cd11c",
                  "Siglech", "Bst2", "Ly6g", "Ly6c", "S100a8",
                  "Ly6c", "Cd34", "Kit")

VlnPlot(seurat_integrated,
        group.by = "Updated_Annotation",
        features = mouse_immune,
        pt.size = 0,
        ncol = 2) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

## Marker expression across myeloid celltypes in each group

```{r, message = F, warning = F, fig.width = 12, fig.height = 20, eval = T}

VlnPlot(subset(seurat_integrated, subset = Updated_Annotation %in% c("Myeloid1", "Myeloid2", "Myeloid3", "Myeloid4")),
        group.by = "Sample",
        features = unique(c(m1, m2, mdsc)),
        pt.size = 0,
        ncol = 4) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

