---
title: "Integration and annotation of CAF co-injection scRNAseq data"
author: "heinin"
date: "2026-01-31"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Libraries and environment variables

```{r, message = F, warning = F}

library(workflowr)
library(Seurat)
library(googlesheets4)
library(tidyverse)
library(plyr)
library(SoupX)
library(scProportionTest)
library(clusterProfiler)
library(org.Hs.eg.db)
library(org.Mm.eg.db)
library(biomaRt)
library(gprofiler2)

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(1234)
options(future.globals.maxSize = 30000 * 1024^2)
reduction <- "umap"

source("/home/hnatri/SingleCellBestPractices/scripts/preprocessing_qc_module.R")
source("/home/hnatri/SingleCellBestPractices/scripts/integration_module.R")
source("/home/hnatri/Utilities/utilities.R")
source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

# Markers
main_plot_features <- c("PTPRC", "CD3D", "CD3E", "CD4", "CD8A", "JAK1", "JAK2", "STAT4", "STAT3", "TIGIT", "GZMB", "SELL", "CD19", "CD68", "CD44", "MARCO", "APOE", "C1QB", "C1QBP", "CD163", "SPP1", "VCAN", "MUC5AC", "NOTCH3", "RGS5", "MS4A1", "PGA5", "PLVAP", "CCL21", "LYVE", "SPARCL1", "FN1", "DCN", "LUM", "COL1A2", "PDPN", "ACTA2", "LMNA", "EGR3", "TP53", "JUN", "FOS", "BRAF", "KIT", "SOX9", "RNF43", "EPCAM", "PECAM", "CEACAM", "TGFB1", "JUNB", "TCF7", "GZMA", "GZMK", "CD47", "CD74", "LYZ", "TREM2", "CD247", "COL5A1", "SPARC", "CD36", "CTLA4", "RGS1", "CCL2", "SELENOP", "LYVE1", "CTSD", "MMP9", "CDK1", "FLT1", "RAMP2", "CCL14", "MCAM", "NRP1", "CEACAM6", "IL32", "TGFBI", "VEGFA", "LAMP1", "IGFBP7", "POSTN", "C1S", "C1R", "CSF1", "NOTCH1", "FOXP3", "IL2RB", "TUBA1B", "S100A9", "C3", "CXCR2", "SDC1", "CD38", "IGHG1", "PRDX4", "ICAM1", "GATA2", "CXCL3")

mycafs <- c("TGFB1", "ACTA2", "TAGLN") #"HOPX", "MYL9"
icafs <- c("IL6", "CXCL12", "CXCL14") #"CXCL13", 
abcafs <- c("CD74", "HLA-DPB1", "HLA-DPB2", "HLA-DQA1", "HLA-DRA", "HLA-DRB1")
ficafs <- c("SFRP2", "SFRP4", "COMP", "CXCL14", "PLA2G2A", "ASPN")

main_plot_features <- c(main_plot_features, mycafs, icafs, abcafs, ficafs)
main_plot_features <- unique(main_plot_features)

main_plot_features <- gorth(main_plot_features, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

```

## Import data

```{r, message = F, warning = F, eval = T, eval = F}

seurat_data <- readRDS("/scratch/hnatri/PIPAC/PDX_CAF_Seurat_filtered.rds")

hist(seurat_data$percent.mt_RNA)
table(seurat_data$Sample)

```

## Integration

```{r, message = F, warning = F, eval = F}

DefaultAssay(seurat_data) <- "RNA"
seurat_data <-  SCTransform(seurat_data, 
                            vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA"),
                            verbose = F)

s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

mmus_s <- gorth(s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m <- gorth(g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

seurat_data <- CellCycleScoring(seurat_data,
                                s.features = mmus_s,
                                g2m.features = mmus_g2m,
                                set.ident = F,
                                verbose = F) 

seurat_data <-  SCTransform(seurat_data, 
                            vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA",
                                                "S.Score", "G2M.Score", "Phase"),
                            verbose = F)

seurat_data <- RunPCA(seurat_data, assay = "SCT",
                      verbose = F)

seurat_integrated <- IntegrateLayers(object = seurat_data,
                                     method = HarmonyIntegration,
                                     orig.reduction = "pca",
                                     new.reduction = 'harmony',
                                     normalization.method = "SCT",
                                     k.weight = 20,
                                     verbose = FALSE)

#saveRDS(seurat_integrated, "/scratch/hnatri/PIPAC/PDX_CAF_Seurat_integrated.rds")

```

## Dimensionality reduction and clustering

```{r, message = F, warning = F, fig.width = 5, fig.height = 4, eval = F}

#seurat_integrated <- readRDS("/scratch/hnatri/PIPAC/PDX_CAF_Seurat_integrated.rds")

seurat_integrated <-  SCTransform(seurat_integrated, 
                                  vars.to.regress = c("percent.mt_RNA", "nCount_RNA", "nFeature_RNA",
                                                      "S.Score", "G2M.Score", "Phase"),
                                  verbose = F)

seurat_integrated <- RunPCA(seurat_integrated,
                            assay = "SCT",
                            verbose = F)

umap_pcs <- get_pcs(seurat_integrated, reduction_name = "pca")

seurat_integrated <- RunUMAP(object = seurat_integrated,
                             assay = "SCT",
                             reduction = "harmony",
                             reduction.name = "harmony_umap",
                             reduction.key = "harmony_umap_",
                             dims = 1:min(umap_pcs),
                             return.model = TRUE,
                             verbose = F)

seurat_integrated <- FindNeighbors(object = seurat_integrated,
                                 assay = "SCT",
                                 reduction = "harmony",
                                 dims = 1:min(umap_pcs),
                                 graph.name = c("harmony_nn",
                                                "harmony_snn"),
                                 verbose = F)

seurat_integrated <- FindClusters(object = seurat_integrated,
                                resolution = c(0.1,0.2,0.3,0.5,0.8,1),
                                graph.name = "harmony_snn",
                                 verbose = F)

DimPlot(seurat_integrated,
        group.by = "harmony_snn_res.0.8",
        reduction = "harmony_umap",
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

table(seurat_integrated$harmony_snn_res.0.8)

saveRDS(seurat_integrated, "/scratch/hnatri/PIPAC/PDX_CAF_Seurat_integrated.rds")

```

## Annotation

SoupX RNA

```{r, message = F, warning = F, fig.width = 22, fig.height = 56, eval = F}

#seurat_integrated <- readRDS("/scratch/hnatri/PIPAC/PDX_CAF_Seurat_integrated.rds")

DefaultAssay(seurat_integrated) <- "SoupX_RNA"
seurat_integrated <- JoinLayers(seurat_integrated)
seurat_integrated <- NormalizeData(seurat_integrated)

FeaturePlot(seurat_integrated,
            features = sort(main_plot_features),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 5) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

RNA

```{r, message = F, warning = F, fig.width = 4, fig.height = 4}

seurat_integrated <- readRDS("/scratch/hnatri/PIPAC/PDX_CAF_Seurat_integrated.rds")

seurat_integrated <- PercentageFeatureSet(seurat_integrated, pattern = "^mt-", col.name = "percent.mt_RNA", assay = "RNA")

DefaultAssay(seurat_integrated) <- "RNA"
seurat_integrated <- JoinLayers(seurat_integrated)
seurat_integrated <- NormalizeData(seurat_integrated)

DimPlot(seurat_integrated,
        group.by = "harmony_snn_res.0.8",
        reduction = "harmony_umap",
        label = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

table(seurat_integrated$harmony_snn_res.0.8)

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8}

mini_features <- gorth(c("PTPRC", "CD3E", "CD8", "SPP1", "CD68",
                         "FN", "COL1A1", "EPCAM", "VIM"), source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

FeaturePlot(seurat_integrated,
            features = mini_features,
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 3) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 16}

clusters <- as.factor(sort(unique(seurat_integrated$harmony_snn_res.0.8)))
cluster_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(clusters))
names(cluster_col) <- levels(clusters)

create_dotplot_heatmap(seurat_object = seurat_integrated,
                       plot_features = sort(main_plot_features),
                       group_var = "harmony_snn_res.0.8",
                       group_colors = cluster_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

### Top cluster markers

```{r, message = F, warning = F, fig.width = 5, fig.height = 4}

Idents(seurat_integrated) <- seurat_integrated$harmony_snn_res.0.8
cluster_markers <- FindAllMarkers(seurat_integrated,
                                  return.thresh = 0.01,
                                  logfc.threshold = 0.5,
                                  min.pct = 0.20,
                                  verbose = F)

table(cluster_markers$cluster)

top_cluster_markers <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:10)

top_cluster_markers_save <- cluster_markers %>%
  arrange(dplyr::desc(avg_log2FC)) %>%
  group_by(cluster) %>%
  dplyr::slice(1:30)

#write.table(top_cluster_markers_save, "/home/hnatri/PIPAC_spatial/PDX_CAF_cluster_markers.tsv",
#            quote = F, sep = "\t")

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 24}

create_dotplot_heatmap(seurat_object = seurat_integrated,
                       plot_features = unique(top_cluster_markers$gene),
                       group_var = "harmony_snn_res.0.8",
                       group_colors = cluster_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

```{r, message = F, warning = F, fig.width = 22, fig.height = 66}

FeaturePlot(seurat_integrated,
            features = unique(top_cluster_markers$gene),
            cols = c("gray98", "tomato3"),
            reduction = "harmony_umap",
            raster = T,
            order = T,
            ncol = 8) &
  coord_fixed(ratio = 1) &
  theme_bw() &
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 8, eval = T}

# Annotation df
annotations <- data.frame("cluster" = c(0, seq(1, 24)),
                          "annotation" = c("Tumor8", # 0
                                           "Tumor1",
                                           "Tumor2",
                                           "Tumor9", # 3
                                           "Tumor5", # 4
                                           "CAF7",
                                           "CAF3",
                                           "Tumor3",
                                           "CAF4", #8
                                           "CAF2",
                                           "Myeloid1",
                                           "Tumor4",
                                           "Myeloid2", # 12 
                                           "Myeloid5", #
                                           "Lymphoid1_CD8_TIGIT",
                                           "Lymphoid2", # 15
                                           "CAF5",
                                           "Lymphoid3_GZMB", # 17
                                           "Tumor7",
                                           "Lymphoid3_B", #
                                           "Tumor6", # 20
                                           "CAF6",
                                           "Myeloid3",
                                           "CAF1", # 23
                                           "Myeloid4"))

seurat_integrated$Annotation <- mapvalues(seurat_integrated$harmony_snn_res.0.8,
                                from = annotations$cluster,
                                to = annotations$annotation)

seurat_integrated$Annotation <- factor(seurat_integrated$Annotation,
                                       levels = sort(unique(as.character(seurat_integrated$Annotation))))

# Updating annotations
#seurat_integrated$Updated_Annotation <- seurat_integrated$Annotation
#seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified1", "Lymphoid3", seurat_integrated$Updated_Annotation)
#seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified2", "Lymphoid4", seurat_integrated$Updated_Annotation)
#seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified6", "Lymphoid5", seurat_integrated$Updated_Annotation)
#seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified3", "Myeloid3", seurat_integrated$Updated_Annotation)
#seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified4", "Myeloid4", seurat_integrated$Updated_Annotation)
#seurat_integrated$Updated_Annotation <- gsub("Immune_Unclassified5", "Unclassified", seurat_integrated$Updated_Annotation)
#
#seurat_integrated$Updated_Annotation <- factor(seurat_integrated$Updated_Annotation,
#                                       levels = sort(unique(as.character(seurat_integrated$Updated_Annotation))))
#
#table(seurat_integrated$Updated_Annotation)

saveRDS(seurat_integrated, "/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_annotated.rds")

```


```{r, message = F, warning = F, fig.width = 6, fig.height = 6, eval = T}

#seurat_integrated <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_annotated.rds")

# Top markers for each celltype
markers <- presto::wilcoxauc(seurat_integrated,
                             group_by = "Annotation",
                             assay = "data",
                             seurat_assay = "RNA")
top_markers <- markers %>% group_by(group) %>% slice_max(order_by = auc, n = 50)

write.table(top_markers, "/home/hnatri/PDX_celltype_markers_top50.tsv",
           quote = F, sep = "\t", row.names = F)

cts <- as.factor(sort(unique(as.character(seurat_integrated$Annotation))))
ct_col <- colorRampPalette(brewer.pal(10, "Paired"))(nb.cols <- length(cts))
names(ct_col) <- levels(cts)

DimPlot(seurat_integrated,
        group.by = "Annotation",
        reduction = "harmony_umap",
        cols = ct_col,
        label = T,
        label.box = T,
        repel = T) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()

```

```{r, message = F, warning = F, fig.width = 12, fig.height = 12, eval = T}

DimPlot(seurat_integrated,
        group.by = "Annotation",
        split.by = "Annotation",
        reduction = "harmony_umap",
        cols = ct_col,
        ncol = 8) +
  coord_fixed(ratio = 1) +
  theme_bw() +
  NoLegend()
  
```


```{r, message = F, warning = F, fig.width = 8, fig.height = 23, eval = T}

create_dotplot_heatmap(seurat_object = seurat_integrated,
                       plot_features = unique(top_cluster_markers$gene),
                       group_var = "Annotation",
                       group_colors = ct_col,
                       column_title = "",
                       row_km = 5,
                       col_km = 5,
                       row.order = NULL,
                       col.order = NULL)

```

```{r, message = F, warning = F, fig.width = 8, fig.height = 24, eval = F}

#VlnPlot(seurat_integrated,
#        group.by = "Annotation",
#        features = main_plot_features,
#        pt.size = 0,
#        ncol = 2) &
#  #geom_boxplot() &
#  theme_bw() &
#  NoLegend() &
#  RotatedAxis() &
#  xlab("")

```

## CAF markers

```{r, message = F, warning = F, fig.width = 8, fig.height = 24, eval = T}

#grep("H2", rownames(seurat_integrated), value = T)
caf_markers <- c("Ly6C", "Il6", "Ccl2", "Cxcl1", "Cxcl2", "Cxcl12",
                 "Pdgfra", "Agtr1", "Cfd", "Il1a", "Il1b", "Ccl8",
                 "Ccl27", "Acta2", "Myl9", "Tagln", "Tpm2", "Col1a1",
                 "Col1a2", "Col3a1")
caf_markers <- unique(caf_markers)

VlnPlot(seurat_integrated,
        group.by = "Annotation",
        features = caf_markers,
        pt.size = 0,
        ncol = 2) &
  #geom_boxplot() &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

## Myeloid markers

Key Mouse MDSC Markers and Subsets \

- Pan-MDSC: CD11b+, Gr-1+ \
- PMN-MDSC (Granulocytic): CD11b+ Ly6G+ Ly6Clow, CD244+, Arg2, Stat3, Jaml. \
- M-MDSC (Monocytic): CD11b+ Ly6G− Ly6Chigh, CCR2+, Arg1, Csf1r. \
- Functional Genes: Arg1 (Arginase-1), Nos2 (iNOS), S100a8, S100a9, ROS.

```{r, message = F, warning = F, fig.width = 8, fig.height = 15, eval = T}

mdsc <- c("Cd11b", "Gr-1", "Ly6G", "Ly6C", "Cd244", "Arg2", "Stat3", "Jaml",
          "Ccr2", "Arg1", "Csf1r", "Nos2", "S100a8", "S100a9", "Ros")

VlnPlot(seurat_integrated,
        group.by = "Annotation",
        features = mdsc,
        pt.size = 0,
        ncol = 2) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

Key Mouse M2 Marker Genes: \

- Arginase-1 (Arg1): Crucial enzyme for polyamine synthesis. \
- Mannose Receptor (Mrc1/CD206): Surface receptor, often used for identifying M2 macrophages. \
- Chitinase-like 3 (Chil3/Ym1): Secreted lectin involved in inflammation. \
- Resistin-like molecule alpha (Retnla/Fizz1): Secreted protein, common marker. \
- CD163 Scavenger Receptor (Cd163): Hemoglobin scavenger receptor. \
- Insulin-like Growth Factor 1 (Igf1): Growth factor involved in tissue repair. \
- Transcription Factors/Signaling: Pparg, Irf4, and Stat6. \
- Chemokines/Cytokines: Ccl17, Ccl22, Il10, and Tgfb1. \
- Novel Markers: Egr2 and Fpr2 are also identified as markers for M2 polarization.


```{r, message = F, warning = F, fig.width = 8, fig.height = 22, eval = T}

m2 <- c("Arg1", "Mrc1", "Cd206", "Chil3", "Ym1", "Retnla", "Fizz1",
        "Cd163", "Igf1", "Pparg", "Irf4", "Stat6", "Ccl17", "Ccl22", "Il10",
        "Tgfb1", "Egr2", "Fpr2")

VlnPlot(seurat_integrated,
        group.by = "Annotation",
        features = m2,
        pt.size = 0,
        ncol = 2) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

Core M1 Marker Genes and Molecules (Murine) \

- Enzyme: Nos2 (Inducible Nitric Oxide Synthase) - Often considered the quintessential M1 marker. \
- Cytokines: Il1b (IL-1), Il6, Tnfa, Il12a, Il12b, Il23. \
- Chemokines: Ccl5 (RANTES), Cxcl2, Cxcl10. \
- Surface Markers: Cd80, Cd86, Cd38, Cd68, Cd64, Fcgr1 (CD64), H2-Aa (MHC II), Cd11c. \
- Novel/Specific Markers: Cd38, Fpr2, Gpr18.

```{r, message = F, warning = F, fig.width = 8, fig.height = 22, eval = T}

m1 <- c("Nos2", "Il1b", "Il6", "Tnfa", "Il12a", "Il12b", "Il23", "Ccl5",
        "Cxcl2", "Cxcl10", "Cd80", "Cd38", "Cd68", "Cd64", "Fcgr1", "H2-Aa",
        "Cd11c", "Fpr2", "Gpr18")

VlnPlot(seurat_integrated,
        group.by = "Annotation",
        features = m1,
        pt.size = 0,
        ncol = 2) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

## Mouse immune markers

Mouse Immune Cell Marker Summary (Mus musculus) \

- Total Leukocytes: Ptprc (CD45)\
- T Cells (Pan): Cd3e, Cd3g, Cd247 (CD3)\
- Helper T Cells: Cd4\
- Cytotoxic T Cells: Cd8a, Cd8b1\
- Regulatory T Cells (Tregs): Foxp3, Cd25 (Il2ra), Ctla4\
- B Cells: Cd19, Ms4a1 (CD20), Cd79a\
- NK Cells: Ncr1 (NKp46), Klrb1c (NK1.1), Cd226\
- Macrophages (Pan): Adgre1 (F4/80), Cd68, Cd11b (Itgam)\
- M1 Macrophages (Pro-inflammatory): Nos2 (iNOS), Cd80, Cd86\
- M2 Macrophages (Anti-inflammatory): Cd163, Mrc1 (CD206), Arg1\
- Conventional Dendritic Cells (cDCs): Itgax (CD11c), H2-Ab1 (MHCII)\
- Plasmacytoid DCs (pDCs): Siglech, Bst2 (CD317)\
- Neutrophils: Ly6g, Ly6c, S100a8\
- Monocytes: Ly6c, Cd11b (Itgam)\
- Hematopoietic Stem/Progenitor Cells: Cd34, Kit (c-Kit) \

Key Activation/Memory Markers\

- Naïve T Cells: Sell (CD62L), Ccr7\
- Memory T Cells: Cd44, Il7r (CD127)\
- Exhaustion Markers: Pdcd1 (PD-1), Cd274 (PD-L1), Havcr2 (TIM-3) \

```{r, message = F, warning = F, fig.width = 8, fig.height = 40, eval = T}

mouse_immune <- c("Ptprc", "Cd3e", "Cd3g", "Cd247", "Cd4", "Cd8a", "Cd8b1", 
                  "Foxp3", "Cd25", "Il2ra", "Ctla4", "Cd19", "Ms4a1", "Cd79a",
                  "Ncr1", "NKp46", "Klrb1c", "Cd226",
                  "Adgre1", "Cd68", "Cd11b", "Itgam",
                  "Nos2", "Cd80", "Cd86",
                  "Cd163", "Mrc1", "Arg1", "Itgax", "Cd11c",
                  "Siglech", "Bst2", "Ly6g", "Ly6c", "S100a8",
                  "Ly6c", "Cd34", "Kit")

VlnPlot(seurat_integrated,
        group.by = "Annotation",
        features = mouse_immune,
        pt.size = 0,
        ncol = 2) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

## EMT markers

EMT marker on tumor cells\

Epithelial (down in EMT): CDH1 (E-cadherin), TACSTD2 (EPCAM), CLDN1.\
Mesenchymal (up in EMT): VIM (Vimentin ), FN1 (Fibronetin), CDH2 (N-cadherin), ITGA5.\
EMT TFs (up): SNAI1, SNAI2, ZEB1, ZEB2, TWIST1.\
Invasion/ECM (up): MMP2, MMP9, COL1A1

Cdh1, Epcam, Krt8, Krt18, Ocln , Tjp1 and transcription
factors including Snai1, Snai2, Twist1, Zeb1, Zeb2 up: Cdh2, Vim, Fn1,
Col1a1, Acta2\

```{r, message = F, warning = F, fig.width = 8, fig.height = 17, eval = T}

#grep("Mmp2", rownames(seurat_integrated), value = T)

emt_markers <- c("Cdh1", "Epcam", "Vim", "Fn1", "Cdh2", "Itga5",
                 "Snai1", "Snai2", "Twist1", "Zeb1", "Zeb2", "Mmp2",
                 "Mmp9", "Col1a1")

VlnPlot(seurat_integrated,
        group.by = "Annotation",
        features = emt_markers,
        pt.size = 0,
        ncol = 2) &
  theme_bw() &
  NoLegend() &
  RotatedAxis() &
  xlab("")

```

## QC metrics by celltype

```{r, message = F, warning = F, fig.width = 12, fig.height = 9, eval = T}

#grep("^mt-", rownames(seurat_integrated), value = T)
#seurat_integrated <- PercentageFeatureSet(seurat_integrated, pattern = "^mt-", col.name = "percent.mt_RNA", assay = "RNA")
#hist(seurat_integrated$percent.mt_RNA, breaks = 100)

VlnPlot(seurat_integrated,
        features = c("nCount_RNA", "nFeature_RNA", "percent.mt_RNA"),
        group.by = "Annotation",
        pt.size = 0,
        ncol = 1) &
  NoLegend()

VlnPlot(seurat_integrated,
        features = c("G2M.Score", "S.Score"),
        group.by = "Annotation",
        pt.size = 0,
        ncol = 1) &
  NoLegend()

```

## Cell numbers by celltype

```{r, message = F, warning = F, fig.width = 12, fig.height = 9, eval = T}

table(seurat_integrated$Annotation)

#Rscript -e "rmarkdown::render('PDX_CAF_scRNAseq_analysis.Rmd')"

```

