---
title: "CMS classification of PDX educated mouse tumors"
author: "heinin"
date: "2026-01-31"
output:
  workflowr::wflow_html:
    code_folding: hide
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Libraries and environment variables

```{r, message = F, warning = F}

library(workflowr)
library(Seurat)
library(googlesheets4)
library(tidyverse)
library(plyr)
library(org.Hs.eg.db)
library(org.Mm.eg.db)
library(biomaRt)
library(Biobase) # for CMScaller, if input is ExpressionSet
library(CMScaller)
library(gprofiler2)

setwd("/home/hnatri/PIPAC_spatial/")
set.seed(1234)
options(future.globals.maxSize = 30000 * 1024^2)
reduction <- "umap"

source("/home/hnatri/PIPAC_spatial/code/PIPAC_colors_themes.R")
source("/home/hnatri/PIPAC_spatial/code/plot_functions.R")

```

## Import data

```{r, message = F, warning = F, eval = T}

seurat_data <- readRDS("/tgen_labs/banovich/PIPAC/CAF_scRNAseq/PDX_CAF_Seurat_annotated.rds")

table(seurat_data$Sample)

```

## Pseudobulk analysis

```{r, message = F, warning = F, eval = T}

# Pseudobulk of tumor
tumor_caf <- setdiff(unique(seurat_data$harmony_snn_res.0.8), c(t_cells, myeloid, immune))
tumor_only <- subset(seurat_data, subset = harmony_snn_res.0.8 %in% tumor_caf)

DefaultAssay(tumor_only) <- "RNA"
tumor_only <- NormalizeData(tumor_only)
Idents(tumor_only) <- tumor_only$Sample
pbcounts <- AggregateExpression(tumor_only,
                                normalization.method = "none")
pbcounts <- pbcounts[["RNA"]]

```

## Running classification

```{r, message = F, warning = F, eval = F}

ensembl <- useMart("ensembl", dataset="hsapiens_gene_ensembl")
geneinfo <- getBM(attributes=c("hgnc_symbol", "ensembl_gene_id", 
                               "entrezgene_id"), 
      filters = 'hgnc_symbol', 
      values = rownames(pbcounts), 
      mart = ensembl)

pbcounts <- as.data.frame(pbcounts)
pbcounts$gene <- mapvalues(rownames(pbcounts),
                           from = geneinfo$hgnc_symbol,
                           to = geneinfo$entrezgene_id)

pbcounts$gene <- ifelse(pbcounts$gene %in% rownames(pbcounts), NA, pbcounts$gene)

pbcounts <- pbcounts[complete.cases(pbcounts),]

pbcounts <- dplyr::group_by(pbcounts, gene) %>%
  mutate(row_number() == 1) %>%
  filter(`row_number() == 1` == TRUE)

pbcounts <- as.data.frame(pbcounts)
rownames(pbcounts) <- pbcounts$gene
pbcounts <- pbcounts[,c(1, 2)]

# Not applying normalization
res <- CMScaller(emat=pbcounts, RNAseq=T, FDR=0.1)

head(res, n=3)

```

## Module scores

```{r, message = F, warning = F, fig.width = 5, fig.height = 4, eval = T}

load("/home/hnatri/PIPAC_spatial/templates.CMS.rda")

genenames <- gorth(templates.CMS$symbol, source_organism = "hsapiens", target_organism = "mmusculus")

templates.CMS$gene <- mapvalues(templates.CMS$symbol,
                                from = genenames$input,
                                to = genenames$ortholog_name)

features_list <- split(templates.CMS, templates.CMS$class)
features_list <- lapply(features_list, function(xx){
  xx$gene
})

```

```{r, message = F, warning = F, fig.width = 5, fig.height = 4, eval = F}

tumor_only <- AddModuleScore(tumor_only,
                             features = features_list,
                             name = "CMSmodule")

VlnPlot(tumor_only,
        features = c("CMSmodule1", "CMSmodule2", "CMSmodule3", "CMSmodule4"),
        group.by = "Sample",
        pt.size = 0,
        ncol = 2) &
  geom_boxplot(width=0.6, fill="white") &
  NoLegend() &
  xlab("")

```

## scGSEA

```{r, message = F, warning = F, fig.width = 7, fig.height = 6}

res_df <- readRDS("/scratch/hnatri/PIPAC/scGSVA_out.rds")

for(i in names(features_list)){
  seurat_data@meta.data[,i] <- res_df[,i]
}

# Pseudobulk of tumor
t_cells <- c(12, 13)
myeloid <- c(10, 11)
immune <- c(17, 21, 20, 22, 24, 23)
tumor_caf <- setdiff(unique(seurat_data$harmony_snn_res.0.8), c(t_cells, myeloid, immune))
tumor_only <- subset(seurat_data, subset = harmony_snn_res.0.8 %in% tumor_caf)

VlnPlot(tumor_only,
        features = c("CMS4"),
        group.by = "Sample",
        pt.size = 0,
        ncol = 1) &
  geom_boxplot(width=0.6, fill="white") &
  NoLegend() &
  xlab("") &
  ylab("CMS4 signature") &
  ggtitle("")

g1 <- tumor_only@meta.data %>%
  filter(Sample == "CT26")

g2 <- tumor_only@meta.data %>%
  filter(Sample == "CT26_CAF")

summary(g1$CMS4)
summary(g2$CMS4)
t.test(g1$CMS4, g2$CMS4)

#filename <- "/home/hnatri/PIPAC_spatial/CT26_CAF_CMS4signature.pdf"
#
#pdf(file=filename,
#    width = 4,
#    height = 4)
#
#VlnPlot(tumor_only,
#        features = c("CMS4"),
#        group.by = "Sample",
#        pt.size = 0,
#        ncol = 1) &
#  #geom_boxplot(width=0.6, fill="white") &
#  NoLegend() &
#  xlab("") &
#  ylab("CMS4 signature") &
#  ggtitle("")
#
#dev.off()

VlnPlot(tumor_only,
        features = c("CMS1", "CMS2", "CMS3", "CMS4"),
        group.by = "Sample",
        pt.size = 0,
        ncol = 2) &
  geom_boxplot(width=0.6, fill="white") &
  NoLegend() &
  xlab("")
  #ylab("CMS4 signature") &
  #ggtitle("")

#plot(tumor_only$CMSmodule1, tumor_only$CMS1)

```

## Individual genes

### CMS1

```{r, message = F, warning = F, fig.width = 15, fig.height = 50}

DefaultAssay(tumor_only) <- "RNA"
tumor_only <- JoinLayers(tumor_only)

VlnPlot(tumor_only,
        features = sort(features_list$CMS1),
        group.by = "Sample",
        pt.size = 0,
        ncol = 6) &
  geom_boxplot(width=0.6, fill="white") &
  theme_classic() &
  NoLegend() &
  xlab("")

```

### CMS2

```{r, message = F, warning = F, fig.width = 15, fig.height = 37}

VlnPlot(tumor_only,
        features = sort(features_list$CMS2),
        group.by = "Sample",
        pt.size = 0,
        ncol = 6) &
  geom_boxplot(width=0.6, fill="white") &
  theme_classic() &
  NoLegend() &
  xlab("")

```

### CMS3

```{r, message = F, warning = F, fig.width = 15, fig.height = 37}

VlnPlot(tumor_only,
        features = sort(features_list$CMS3),
        group.by = "Sample",
        pt.size = 0,
        ncol = 6) &
  geom_boxplot(width=0.6, fill="white") &
  theme_classic() &
  NoLegend() &
  xlab("")

```

### CMS4

```{r, message = F, warning = F, fig.width = 15, fig.height = 70}

VlnPlot(tumor_only,
        features = sort(features_list$CMS4),
        group.by = "Sample",
        pt.size = 0,
        ncol = 6) &
  geom_boxplot(width=0.6, fill="white") &
  theme_classic() &
  NoLegend() &
  xlab("")

```

Regarding CMS classication .. CMS4 was derived from bulk RNA-seq data. Single cell analysis has found that that tumor cells in CMS4 tumors are either iCMS2 (enerocyte like) or iCMS3 (goblet ell like). This paper suggests that iCMS3 has more predicliction for peritoneal mets and associated with F-apCAF and inflammed but dysfunctional immune profile
